---
title: "Alert Utility Case Confirmation"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    number-depth: 4
    code-fold: true
    embed-resources: true
    theme:
      - custom.scss
params:
  testing: false
execute:
  warning: false
message: false
cache: false
---

```{r}
#| label: pkg-reinstall
#| echo: false

# detach("package:OutbreakExtractR", unload=TRUE)
# devtools::document("../OutbreakExtractR")
# install.packages("../OutbreakExtractR", type="source", repos=NULL)
# params <- list(testing=FALSE)
```

```{r}
#| label: setup
#| echo: false

## Load necessary libraries
library(OutbreakExtractR)
library(tidyverse)
library(scales)
library(sf)
library(broom)
library(knitr)
library(kableExtra)
library(data.table)
library(patchwork)
library(ggpubr)
library(purrr)
library(stringr)


new_dpath <- "../data"
plot_path <- paste0(new_dpath, "/exploratory")
pre_path <- "../data/"

testing <- params$testing ## process the testing dataset or full evaluation of outbreaks
test_cntry <- if(testing){"COD"}
cCh_threshold <- 5

## set parameters for reading files
evalperiod_code <- "ew52"
delayperiod_code <- "dw4" ## delay period for base analyses
ag_nweeks_code <- "nweeks8"
```

Updated `r Sys.Date()`

`r paste("Testing is", testing, ifelse(testing, paste0("with test country ", test_cntry, "."), "so results are for the full dataset."))`

Alertgroup `r ag_nweeks_code`; Delay period `r delayperiod_code`; Evaluation period `r evalperiod_code`

```{css classDummyTable, echo=FALSE}

.dummyTable .table thead {
  visibility: collapse;
}
.dummyTable .table tbody {
  visibility: collapse;
}
.dummyTable .table {
  border-collapse: inherit;
  padding-bottom: none;
}
.dummyTable .table.table.table-sm.table-striped.small {
  margin: auto;
}
```

```{r}
#| label: funcs-reactable-template
#| echo: false

#' @description Generic function to prepare reactable
#' @param print_table dataframe ready to be printed
#' @param col_names character vector of column names
#' @returns reactable
reactable_template <- function(print_table, col_names){
  reactable::reactable(print_table,
                       columns = col_names,
                       style = list(fontFamily = "sans-serif"),
                       highlight = TRUE,
                       striped = TRUE,
                       compact = TRUE,
                       filterable = TRUE,
                       defaultPageSize = 20
  )
}
```

```{r}
#| label: utility_functions
#| echo: false

quibble2 <- function(colname, q = c(0.025, 0.25, 0.5, 0.75, 0.975)) {
  tibble("{{ colname }}_q" := paste0("q", stringr::str_remove(as.character(q), "0.")), "{{ colname }}" := quantile(colname, q, na.rm = TRUE))
}

prepare_fig_dat <- function(df){
  OutbreakExtractR::label_alerts_short(df, is_ordered = TRUE) %>%
  dplyr::mutate(alert_lab = forcats::fct_rev(alert_lab)) %>%
  OutbreakExtractR::label_alerttypes()
}

```


```{r}
#| label: load-data
#| echo: false

## Set Parameters
incl_trend_alerts <- FALSE ## we do not use trend-based alerts for this sensitivity analysis
impact_thresh <- 300 ## outcome threshold for ppv, missed outbreaks, delay
use_filtered_linked <- TRUE ## TRUE for dataset that CA cleaned
which_setting <- "epidemic" ## endemic, epidemic, all

## Read Data
alerts <- readr::read_rds(paste0(new_dpath, "/alerts.rds"))
ags <- readr::read_rds(paste0(new_dpath, "/alert_groups_", ag_nweeks_code, ".rds"))
evalC <- readr::read_rds(paste0(new_dpath, "/evaluation_cases_", delayperiod_code, "_", evalperiod_code, ".rds"))

if(use_filtered_linked){
  linked <- readr::read_rds(paste0(new_dpath, "/filtered_linked_outbreaks_alert_groups_", ag_nweeks_code, ".rds"))
} else{
  linked <- readr::read_rds(paste0(new_dpath, "/linked_outbreaks_alert_groups_", ag_nweeks_code, ".rds"))
}

pre_extract <- readr::read_rds(file.path(pre_path, "time_series_preoutbreak_extraction_public_apr_2025.rds"))
pre_extract <- dplyr::mutate(pre_extract, pop_brk = cut(pop, breaks = c(0, 50000, 500000, max(pre_extract$pop)), labels = c("<50k", "[50k, 500k)", "\U2265 500k"), include.lowest=T, right=FALSE))

```

The following locations are include in this report: `r which_setting`

```{r}
#| label: clean-base-data
#| echo: false

keep_low_stringency <- function(df, low_stringency_threshold){
  df <- df %>% OutbreakExtractR::add_alert_stringency() %>%
  filter(alert_stringency %in% c(1:low_stringency_threshold)) ## keep only low stringency alerts
  return(df)
}

alerts_cl <- OutbreakExtractR::add_pop3_adm_columns(alerts, pre_extract) %>%
  OutbreakExtractR::add_country_column() %>%
  OutbreakExtractR::filter_ms_data(which_setting = which_setting) %>%
  filter(alert_type != "trend") %>%
  keep_low_stringency(low_stringency_threshold = 5)

ags_cl <- OutbreakExtractR::add_pop3_adm_columns(ags, pre_extract) %>%
  OutbreakExtractR::add_country_column() %>%
  OutbreakExtractR::filter_ms_data(which_setting = which_setting) %>%
  filter(alert_type != "trend") %>%
  keep_low_stringency(low_stringency_threshold = 5)

evalC_cl <- OutbreakExtractR::add_pop3_adm_columns(evalC, pre_extract) %>%
  OutbreakExtractR::add_alert_columns(alerts) %>%
  OutbreakExtractR::filter_ms_data(which_setting = which_setting)%>%
  filter(alert_type != "trend") %>%
  keep_low_stringency(low_stringency_threshold = 5)

```

## Data availability for case confirmation 

```{r}

alerts_wo_trend <- alerts_cl %>%
  filter(alert_type != "trend")

## Using data.table to perform non-equi join where we join based on whether the TL is within 4 weeks of the alert TL
a <- as.data.table(alerts_wo_trend)
p <- as.data.table(pre_extract)

a[, TL := as.Date(TL)]
p[, TL := as.Date(TL)]

## define temporal window
a[, TL_start := TL - 21L]   ## TL - 3 weeks
a[, TL_end   := TL]         ## TL (alert week)

## Non-equi join
result <- p[
  a,
  on = .(location,
         TL >= TL_start,
         TL <= TL_end),
  allow.cartesian = TRUE
][,
  .(total_confirmed_cases = sum(cCh, na.rm = TRUE),
    total_suspected_cases = sum(sCh, na.rm = TRUE)),
  by = alert_id
]

a_with_counts <- a[result, on = .(alert_id)] %>%
  as.data.frame() %>%
  prepare_fig_dat()

a_with_counts <- a_with_counts %>%
  mutate(confirmed_prop = ifelse(total_suspected_cases > 0,
                                 total_confirmed_cases / total_suspected_cases,
                                 NA),
         unconfirmed_prop = 1-confirmed_prop)

## important: here we filter the dataset to only keep the alerts that have total confirmed cases during the alert period according to the threshold we want to explore
alerts_w_cCh_threshold <- a_with_counts %>%
  select(alert_id, alert_lab, total_confirmed_cases, total_suspected_cases, confirmed_prop, location) %>%
  filter(total_confirmed_cases >= cCh_threshold)

message(paste0("the number of alerts in the unfiltered dataset is ",length(a_with_counts$alert_id)))

message(paste0("the number of alerts in the filtered dataset with cCh threshold =  ", cCh_threshold, " is " ,length(alerts_w_cCh_threshold$alert_id)))


```

The number of alerts in the unfiltered dataset is `r length(a_with_counts$alert_id)`

The number of alerts in the filtered dataset with cCh threshold = `r cCh_threshold` is `r length(alerts_w_cCh_threshold$alert_id)`

```{r, fig.width=12}

# a barchart per type showing the proportion per alert_number (bars side-by-side)

# Helper: coerce alert_number to a factor for nicer plotting order when appropriate
if (!"alert_number" %in% names(a_with_counts)) stop("a_with_counts must contain column 'alert_number'")

a_with_counts <- a_with_counts %>%
  mutate(alert_lab = as.factor(alert_lab))

alert_types <- unique(a_with_counts$alert_type)

barchart_plots <- list()

for (atype in alert_types) {
  df_prop <- a_with_counts %>%
    filter(alert_type == atype) %>%
    group_by(alert_lab) %>%
    summarise(
      n_alerts = n(),
      n_with_confirmed = sum((total_confirmed_cases > 0), na.rm = TRUE),
      prop_with_confirmed = ifelse(n_alerts > 0, n_with_confirmed / n_alerts, NA_real_)
    ) %>%
    ungroup()

  if (nrow(df_prop) == 0) next

  p_bar <- ggplot(df_prop, aes(x = alert_lab, y = prop_with_confirmed)) +
    geom_col(width = 0.7) +
    scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0, 1)) +
    labs(
      title = paste0("Proportion of Alerts with \u2265 1 Confirmed Case — Alert Type: ", atype),
      x = "alert_number",
      y = "Proportion with at least one confirmed case"
    ) +
    theme_minimal()

  print(p_bar)
  barchart_plots[[atype]] <- p_bar
  
  df_raw <- df_prop %>%
    select(alert_lab, n_alerts, n_with_confirmed) %>%
    pivot_longer(cols = c(n_alerts, n_with_confirmed),
                 names_to = "metric",
                 values_to = "count")

  p_bar_raw <- ggplot(df_raw,
                      aes(x = alert_lab,
                          y = count,
                          fill = metric)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.7) +
    labs(
      title = paste0("Raw Alert Counts — Alert Type: ", atype),
      x = "alert_number",
      y = "Count",
      fill = "Metric"
    ) +
    scale_fill_manual(
      values = c("n_alerts" = "#1f78b4",
                 "n_with_confirmed" = "#33a02c"),
      labels = c(
        n_alerts = "Total Alerts",
        n_with_confirmed = "Alerts with ≥1 Confirmed Case"
      )
    ) +
    theme_minimal()

  print(p_bar_raw)
  barchart_plots[[paste0(atype, "_raw")]] <- p_bar_raw

}

## summary table 
summary_table <- a_with_counts %>%
  group_by(alert_type, alert_lab) %>%
  summarise(
    n_alerts = n(),
    n_with_confirmed = sum((total_confirmed_cases > 0), na.rm = TRUE),
    prop_with_confirmed = ifelse(n_alerts > 0, n_with_confirmed / n_alerts, NA_real_)
  ) %>%
  ungroup()

print(summary_table)



```

### Proportion of alerts that have (not NA) case confirmation data for at least one week in the alert period and proportion of alerts that have (not NA) case confirmation data for all weeks in the alert period

```{r}
alerts_wo_trend2 <- alerts_cl %>%
  filter(alert_type != "trend") %>%
  prepare_fig_dat()

a_dt <- as.data.table(alerts_wo_trend2)
p_dt <- as.data.table(pre_extract)
a_dt[, TL_alert := as.Date(TL)]
p_dt[, TL_p := as.Date(TL)]
p_dt[, TL := NULL]

## defining the temporal window (alert week + 3 weeks prior)
a_dt[, TL_start := TL_alert - 21L]   ## 21 days before
a_dt[, TL_end   := TL_alert]         ## alert week

joined <- p_dt[
  a_dt,
  on = .(
    location,
    TL_p >= TL_start,
    TL_p <= TL_end
  ),
  allow.cartesian = TRUE
]

alert_week_summary <- joined[, .(
  n_weeks_matched   = .N,                  
  n_nonNA_weeks     = sum(!is.na(cCh)),     ## how many have non-NA cCh
  has_atleast1      = sum(!is.na(cCh)) > 0, ## TRUE if ≥1 week non-NA
  has_all4          = sum(!is.na(cCh)) == 4 ## TRUE if all 4 weeks non-NA
), by = alert_id]

a_combined <- a_dt[alert_week_summary, on = .(alert_id)] %>%
  as.data.frame() %>%
  mutate(alert_lab = as.factor(alert_lab))

summary_prop <- a_combined %>%
  group_by(alert_type, alert_lab) %>%
  summarise(
    n_alerts = n(),
    n_atleast1 = sum(has_atleast1, na.rm = TRUE),
    n_all4     = sum(has_all4, na.rm = TRUE),

    prop_atleast1 = n_atleast1 / n_alerts,
    prop_all4     = n_all4     / n_alerts
  ) %>%
  ungroup()

print(summary_prop)


alert_types <- unique(a_combined$alert_type)
barchart_atleast1 <- list()
barchart_all4     <- list()

for (atype in alert_types) {
  
  df_plot <- summary_prop %>% filter(alert_type == atype)
  if (nrow(df_plot) == 0) next
  
  p1 <- ggplot(df_plot, aes(x = alert_lab, y = prop_atleast1)) +
    geom_col(width = 0.7) +
    scale_y_continuous(labels = percent_format(accuracy = 1),
                       limits = c(0,1)) +
    labs(
      title = paste0("Prop. Alerts with ≥1 Week non-NA cCh, Type: ", atype),
      x = "alert definition", y = "Proportion (≥1 week non-NA)"
    ) +
    theme_minimal()
  
  print(p1)
  barchart_atleast1[[atype]] <- p1
  
  p2 <- ggplot(df_plot, aes(x = alert_lab, y = prop_all4)) +
    geom_col(width = 0.7) +
    scale_y_continuous(labels = percent_format(accuracy = 1),
                       limits = c(0,1)) +
    labs(
      title = paste0("Prop. Alerts with All 4 Weeks non-NA cCh, Type: ", atype),
      x = "alert definition", y = "Proportion (all 4 weeks non-NA)"
    ) +
    theme_minimal()
  
  print(p2)
  barchart_all4[[atype]] <- p2
  
  df_raw <- df_plot %>%
    select(alert_lab, n_alerts, n_atleast1, n_all4) %>%
    pivot_longer(
      cols = c(n_alerts, n_atleast1, n_all4),
      names_to = "metric",
      values_to = "count"
    )

  p_raw <- ggplot(df_raw, aes(x = alert_lab, y = count, fill = metric)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.7) +
    labs(
      title = paste0("Raw Alert Counts, Type: ", atype),
      x = "alert definition",
      y = "Count",
      fill = "Metric"
    ) +
    scale_fill_manual(
      values = c(
        n_alerts   = "#1f78b4",
        n_atleast1 = "#33a02c",
        n_all4     = "#e31a1c"
      ),
      labels = c(
        n_alerts   = "Total Alerts",
        n_atleast1 = "Alerts w/ ≥1 non-NA week",
        n_all4     = "Alerts w/ all 4 non-NA weeks"
      )
    ) +
    theme_minimal()

  print(p_raw)

}


```
### Proportion of outbreaks that have (not NA) case confirmation data for at least one week in the alert period and proportion of outbreaks that have (not NA) case confirmation data for all weeks in the alert period

```{r}

library(data.table)
library(dplyr)
library(ggplot2)
library(scales)

l_dt <- as.data.table(linked)

outbreak_windows <- l_dt[, .(
  location       = first(location),
  outbreak_start = first(outbreak_start),
  outbreak_end   = first(outbreak_end)
), by = outbreak_UID]

p_dt <- as.data.table(pre_extract)
p_dt[, TL_p := as.Date(TL)]
p_dt[, TL := NULL]


joined_outbreak <- p_dt[
  outbreak_windows,
  on = .(
    location,
    TL_p >= outbreak_start,
    TL_p <= outbreak_end
  ),
  allow.cartesian = TRUE
]

outbreak_summary <- joined_outbreak[, .(
  n_weeks_matched = .N,                ## number of outbreak weeks found in preoutbreak time series
  n_nonNA_weeks   = sum(!is.na(cCh)),  ## how many weeks have non-NA cCh
  has_atleast1    = sum(!is.na(cCh)) > 0,
  has_all_weeks   = sum(!is.na(cCh)) == .N
), by = outbreak_UID]

outbreak_combined <- outbreak_windows[outbreak_summary, on = "outbreak_UID"]

linked_outbreak <- linked %>%
  select(outbreak_UID, linkage, location) %>%
  distinct() %>%
  inner_join(outbreak_combined, by = "outbreak_UID")

summary_overall <- outbreak_combined %>%
  summarise(
    n_outbreaks   = n(),
    prop_atleast1 = mean(has_atleast1, na.rm = TRUE),
    prop_all      = mean(has_all_weeks, na.rm = TRUE)
  )

print(summary_overall)

ggplot(summary_overall, aes(x = "all outbreaks", y = prop_atleast1)) +
  geom_col(width = 0.5) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0,1)) +
  labs(
    title = "Proportion of Outbreaks with ≥1 Week non-NA cCh",
    x = "",
    y = "Proportion"
  ) +
  theme_minimal()

ggplot(summary_overall, aes(x = "all outbreaks", y = prop_all)) +
  geom_col(width = 0.5) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0,1)) +
  labs(
    title = "Proportion of Outbreaks with All Weeks non-NA cCh",
    x = "",
    y = "Proportion"
  ) +
  theme_minimal()

summary_raw <- outbreak_combined %>% summarise( n_outbreaks = n(), 
                                                n_with_atleast1 = sum(has_atleast1, na.rm = TRUE),
                                                n_with_all = sum(has_all_weeks, na.rm = TRUE) ) %>%
  tidyr::pivot_longer( cols = everything(), names_to = "metric", values_to = "count" ) 


ggplot(summary_raw, aes(x = metric, y = count, fill = metric)) +
  geom_col(width = 0.6) +
  labs( title = "Raw Outbreak Counts (overall)", x = "", y = "Count", fill = "Metric" ) +
  scale_fill_manual(values = c( n_outbreaks = "#1f78b4", n_with_atleast1 = "#33a02c", n_with_all = "#e31a1c" ), labels = c(n_outbreaks = "Total outbreaks", n_with_atleast1 = "Outbreaks w/ ≥1 non-NA week",                                                                                                              n_with_all = "Outbreaks w/ all weeks non-NA" )) +
  theme_minimal()

## outbreak data
l_dt <- as.data.table(linked)

outbreak_windows <- l_dt[, .(
  location       = first(location),
  outbreak_start = first(outbreak_start),
  outbreak_end   = first(outbreak_end)
), by = outbreak_UID]


p_dt <- as.data.table(pre_extract)
p_dt[, TL_p := as.Date(TL)]
p_dt[, TL := NULL]


joined_outbreak <- p_dt[
  outbreak_windows,
  on = .(
    location,
    TL_p >= outbreak_start,
    TL_p <= outbreak_end
  ),
  allow.cartesian = TRUE
]

outbreak_summary <- joined_outbreak[, .(
  n_weeks_matched = .N,
  n_nonNA_weeks   = sum(!is.na(cCh)),
  has_atleast1    = sum(!is.na(cCh)) > 0,
  has_all_weeks   = sum(!is.na(cCh)) == .N,
  total_cCh       = sum(cCh, na.rm = TRUE)   # <-- key addition
), by = outbreak_UID]


outbreak_combined <- outbreak_windows[outbreak_summary, on = "outbreak_UID"]

linked_outbreak <- linked %>%
  select(outbreak_UID, linkage, location) %>%
  distinct() %>%
  inner_join(outbreak_combined, by = "outbreak_UID")

outbreak_combined_filtered <- outbreak_combined %>%
  #filter(total_cCh >= cCh_threshold) %>%
  mutate(outbreak_UID = paste0(location, "_", outbreak_start, "_", outbreak_end))

summary_overall <- outbreak_combined_filtered %>%
  summarise(
    n_outbreaks   = n(),
    prop_atleast1 = mean(has_atleast1, na.rm = TRUE),
    prop_all      = mean(has_all_weeks, na.rm = TRUE)
  )

print(summary_overall)


ggplot(summary_overall, aes(x = "filtered outbreaks", y = prop_atleast1)) +
  geom_col(width = 0.5) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0,1)) +
  labs(
    title = paste0(
      "Proportion of Outbreaks with ≥1 Week non-NA cCh",
      "\n(Filter: total cCh ≥ ", cCh_threshold, ")"
    ),
    x = "",
    y = "Proportion"
  ) +
  theme_minimal()


ggplot(summary_overall, aes(x = "filtered outbreaks", y = prop_all)) +
  geom_col(width = 0.5) +
  scale_y_continuous(labels = percent_format(accuracy = 1), limits = c(0,1)) +
  labs(
    title = paste0(
      "Proportion of Outbreaks with All Weeks non-NA cCh",
      "\n(Filter: total cCh ≥ ", cCh_threshold, ")"
    ),
    x = "",
    y = "Proportion"
  ) +
  theme_minimal()

summary_raw <- outbreak_combined_filtered %>%
  summarise(
    n_outbreaks      = n(),
    n_with_atleast1  = sum(has_atleast1, na.rm = TRUE),
    n_with_all       = sum(has_all_weeks, na.rm = TRUE)
  ) %>%
  tidyr::pivot_longer(
    cols = everything(),
    names_to = "metric",
    values_to = "count"
  )

ggplot(summary_raw, aes(x = metric, y = count, fill = metric)) +
  geom_col(width = 0.6) +
  labs(
    title = paste0("Raw Outbreak Counts (Filtered)\nFilter: total cCh ≥ ", cCh_threshold),
    x = "",
    y = "Count",
    fill = "Metric"
  ) +
  scale_fill_manual(
    values = c(
      n_outbreaks     = "#1f78b4",
      n_with_atleast1 = "#33a02c",
      n_with_all      = "#e31a1c"
    ),
    labels = c(
      n_outbreaks     = "Total outbreaks (filtered)",
      n_with_atleast1 = "Outbreaks w/ ≥1 non-NA week",
      n_with_all      = "Outbreaks w/ all weeks non-NA"
    )
  ) +
  theme_minimal()


```


```{r}
#| label: prepare-linked-data
#| echo: true

## subset to outbreaks exceeding a threshold
print(paste(impact_thresh, "suspected cases is the outbreak filtering threshold"))

if(use_filtered_linked){

  linked_cl <- OutbreakExtractR::add_pop3_adm_columns_outbreak(linked, pre_extract) %>% ## N.B. function changes the outbreak_UID separators to underscores
    OutbreakExtractR::add_country_column() %>%
    OutbreakExtractR::add_alert_columns_outbreak(alerts) %>%
    OutbreakExtractR::filter_ms_data(which_setting = which_setting) 

  ## checks
  print("Number of outbreak_UIDs missing at least one of 24 alert definitions (linked)")
  linked %>% group_by(outbreak_UID) %>% count() %>% filter(n<24) %>% nrow
  print("Number of outbreak_UIDs missing at least one of 24 alert definitions")
  linked_cl %>% group_by(outbreak_UID) %>% count() %>% filter(n<24) %>% nrow

} else{
  message("Unfiltered links to outbreak data were used. Filtering to one alert definition per outbreak via automated selection.")

  linked_cl_tmp <- OutbreakExtractR::add_pop3_adm_columns_outbreak(linked, pre_extract) %>%
    OutbreakExtractR::add_country_column() %>%
    OutbreakExtractR::add_alert_columns_outbreak(alerts) %>%
    OutbreakExtractR::filter_ms_data(which_setting = which_setting) 

  linked_cl_1row <- dplyr::group_by(linked_cl_tmp, outbreak_UID, alert_number) %>%
    dplyr::count() %>%
    dplyr::filter(n <= 1) %>%
    dplyr::ungroup()
  linked_cl_mrow <- dplyr::group_by(linked_cl_tmp, outbreak_UID, alert_number) %>%
    dplyr::count() %>%
    dplyr::filter(n > 1) %>%
    dplyr::ungroup()

  ## clean outbreak-alert linkages with multiple linkage options
  clean_mrow_linkages <- function(linked_mrow_dat, linked_full_dat, i){
    id <- linked_mrow_dat[i,]$outbreak_UID
    alert_num <- linked_mrow_dat[i,]$alert_number

    tmp <- dplyr::filter(linked_full_dat, outbreak_UID == id & alert_number == alert_num)

    if ("overlap" %in% tmp$linkage){
      tmp2 <- dplyr::mutate(tmp, keep = dplyr::if_else(linkage == "overlap", TRUE, FALSE))
    } else if ("closest before" %in% tmp$linkage){
      tmp2 <- dplyr::mutate(tmp, keep = dplyr::if_else(linkage == "closest before", TRUE, FALSE))
    }

    rc <- dplyr::filter(tmp2, keep) %>% dplyr::select(-keep)
    if(nrow(rc) != 1){
      stop("More than 1 row was returned")
    }
    return(rc)
  }

  linked_mrow_singles <- 1:nrow(linked_cl_mrow) |>
    purrr::map(\(x) clean_mrow_linkages(linked_cl_mrow, linked_cl_tmp, i = x)) %>%
    purrr::list_rbind() %>%
    dplyr::ungroup()

  linked_1rows <- dplyr::left_join(dplyr::select(linked_cl_1row, outbreak_UID, alert_number), linked_cl_tmp, by = c("outbreak_UID", "alert_number"))

  ## checks
  print("Number of outbreak UIDs in pre and post cleaning are the same")
  nrow(linked_mrow_singles) + nrow(linked_1rows) == nrow(dplyr::distinct(linked_cl_tmp, outbreak_UID, alert_number))

  linked_cl <- dplyr::bind_rows(linked_1rows, linked_mrow_singles) %>%
  filter(alert_type != "trend") ## one linkage per alert def - outbreak
}

linked_cl_large <- dplyr::filter(linked_cl, total_sCh > impact_thresh) %>%
  ## only keep outbreaks above impact threshold
  filter(alert_type != "trend")

## filtered outbreaks for cCh VERSION 2
linked_cl_large_filtered <- linked_cl_large %>%
  keep_low_stringency(low_stringency_threshold = 5) %>%
  ## set alert fields to NA / FALSE if alert_id not in alerts_w_cCh_threshold
  mutate(
    is_triggered = if_else(alert_id %in% alerts_w_cCh_threshold$alert_id,
                           is_triggered, FALSE),
    alert_id = if_else(alert_id %in% alerts_w_cCh_threshold$alert_id,
                        alert_id, NA_character_),
    alert_group_start = if_else(alert_id %in% alerts_w_cCh_threshold$alert_id,
                                alert_group_start, as.Date(NA)),
    alert_group_end = if_else(alert_id %in% alerts_w_cCh_threshold$alert_id,
                              alert_group_end, as.Date(NA)),
    linkage = if_else(alert_id %in% alerts_w_cCh_threshold$alert_id,
                      linkage, NA_character_)
  )

```

Identify proportion of target population as \>1 year olds by country according to UNWPP2024 in the alert year.

```{r}
#| label: prep_target_pop
#| echo: false

library(wpp2024)
library(countrycode)
data(popAge1dt)

iso3_uq <- sort(unique(evalC_cl$country))
codes <- dplyr::select(countrycode::codelist, iso3c, un.name.en) %>%
  dplyr::filter(iso3c %in% iso3_uq) %>%
  dplyr::mutate(un.name.en = dplyr::case_when(iso3c == "CIV" ~ "Cote d'Ivoire",
                              .default = un.name.en))

iso3_targetpop <- popAge1dt %>%
  dplyr::mutate(target_pop = dplyr::if_else(age != 0, pop, 0)) %>%
  dplyr::group_by(name, year) %>%
  dplyr::summarise(target_pop = sum(target_pop), total_pop = sum(pop)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(prop_targeted = round(target_pop/total_pop, 2)) %>%
  dplyr::right_join(codes, by = c("name" = "un.name.en")) %>%
  dplyr::relocate(iso3c, .after = "name") %>%
  dplyr::rename(country = iso3c)

targetpop_cl <- dplyr::select(iso3_targetpop, country, year, prop_targeted)

merge_prop_targeted <- dplyr::distinct(ags_cl, alert_id, TL_first_alert, country) %>%
  dplyr::mutate(TL_year = lubridate::year(TL_first_alert)) %>%
  dplyr::left_join(targetpop_cl, by = c("country", "TL_year" = "year")) %>%
  dplyr::select(alert_id, prop_targeted)

```

# Alert utility

1.  Alerts preceding highest potential impact
2.  Alerts preceding highest potential efficiency
3.  Alerts with high PPV (outcomes with \>`r impact_thresh` cases)
4.  Proportion of outbreaks with \>`r impact_thresh` cases that were missed
5.  Delay from outbreak start for outbreaks \>`r impact_thresh` cases


## Dimensions of alert utility

```{r}
#| label: prep_underlying_data
#| echo: false

### impact and efficiency ###
prep_impact_eff_dat <- dplyr::left_join(evalC_cl, merge_prop_targeted, by = c("alert_id")) %>%
  dplyr::rename(impact = total_sCh) %>%
  dplyr::mutate(eff = impact / (prop_targeted*pop) * 1000)

means_impact_eff_dat <- prep_impact_eff_dat %>%
  dplyr::group_by(alert_number, alert_type, pop_brk) %>%
  dplyr::summarise(impact_mean = mean(impact), eff_mean = mean(eff)) %>%
  dplyr::ungroup()

### ppv ###
prep_ppv_dat <- dplyr::mutate(evalC_cl, true_positives := dplyr::if_else(total_sCh > impact_thresh, TRUE, FALSE))

### missed outbreaks and delayed detection ###
prep_missed_delay_dat <- linked_cl_large %>%
  dplyr::mutate(delay = as.numeric(alert_group_start-outbreak_start, units = "weeks"))

means_delay_dat <- prep_missed_delay_dat %>%
  dplyr::group_by(alert_number, alert_type, pop_brk) %>%
  dplyr::summarise(delay_mean = mean(delay, na.rm = TRUE)) %>%
  dplyr::ungroup()

n_outbreaks <- linked_cl_large %>%
  dplyr::distinct(outbreak_UID, pop_brk) %>%
  dplyr::group_by(pop_brk) %>%
  dplyr::summarise(n_outbreaks = n())

sum(n_outbreaks$n_outbreaks) == length(unique(linked_cl_large$outbreak_UID)) ## total number of outbreaks

### important: missed outbreaks and delayed detection for the cCh threshold, matching outbreaks to alert locations above the cCh threshold ###   
linked_cl_large_cCh <- linked_cl_large_filtered %>%
  filter(location %in% alerts_w_cCh_threshold$location)

prep_missed_delay_dat_cCh <- linked_cl_large_cCh %>%
  dplyr::mutate(delay = as.numeric(alert_group_start-outbreak_start, units = "weeks"))

means_delay_dat_cCh <- prep_missed_delay_dat_cCh %>%
  dplyr::group_by(alert_number, alert_type, pop_brk) %>%
  dplyr::summarise(delay_mean = mean(delay, na.rm = TRUE)) %>%
  dplyr::ungroup()

n_outbreaks_cCh <- linked_cl_large_cCh %>%
  dplyr::distinct(outbreak_UID, pop_brk) %>%
  dplyr::group_by(pop_brk) %>%
  dplyr::summarise(n_outbreaks = n())

sum(n_outbreaks_cCh$n_outbreaks) == length(unique(linked_cl_large_cCh$outbreak_UID)) ## total number of outbreaks

```

## Sensitivity Analysis for the specific threshold

```{r}
#| label: prep_underlying_data_cCh
#| echo: false

### impact and efficiency ###
prep_impact_eff_dat_cCh <- dplyr::left_join(evalC_cl, merge_prop_targeted, by = c("alert_id")) %>%
  dplyr::rename(impact = total_sCh) %>%
  dplyr::mutate(eff = impact / (prop_targeted*pop) * 1000) %>%
  filter(alert_id %in% alerts_w_cCh_threshold$alert_id)

means_impact_eff_dat_cCh <- prep_impact_eff_dat_cCh %>%
  dplyr::group_by(alert_number, alert_type, pop_brk) %>%
  dplyr::summarise(impact_mean = mean(impact), eff_mean = mean(eff)) %>%
  dplyr::ungroup()

### ppv ###
evalC_cl_cCh <- evalC_cl %>% filter(alert_id %in% alerts_w_cCh_threshold$alert_id)
prep_ppv_dat_cCh <- dplyr::mutate(evalC_cl_cCh, true_positives := dplyr::if_else(total_sCh > impact_thresh, TRUE, FALSE))

### missed outbreaks and delayed detection ###
prep_missed_delay_dat_cCh <- linked_cl_large_cCh %>%
  dplyr::mutate(delay = as.numeric(alert_group_start-outbreak_start, units = "weeks"))

means_delay_dat_cCh <- prep_missed_delay_dat_cCh %>%
  dplyr::group_by(alert_number, alert_type, pop_brk) %>%
  dplyr::summarise(delay_mean = mean(delay, na.rm = TRUE)) %>%
  dplyr::ungroup()

n_outbreaks_cCh <- linked_cl_large_cCh %>%
  dplyr::distinct(outbreak_UID, pop_brk) %>%
  dplyr::group_by(pop_brk) %>%
  dplyr::summarise(n_outbreaks = n())

sum(n_outbreaks_cCh$n_outbreaks) == length(unique(linked_cl_large_cCh$outbreak_UID)) ## total number of outbreaks

```


```{r}
#| label: prep_fig_data
#| echo: false

fig_impact_dat <- dplyr::group_by(prep_impact_eff_dat, alert_number, alert_type, pop_brk) %>%
  dplyr::reframe(quibble2(impact)) %>%
  tidyr::pivot_wider(names_from = impact_q, names_glue = "{.value}_{impact_q}", values_from = impact) %>%
  dplyr::full_join(dplyr::select(means_impact_eff_dat, alert_number, alert_type, pop_brk, impact_mean), by = c("alert_number", "alert_type", "pop_brk")) %>%
  prepare_fig_dat()

fig_eff_dat <- dplyr::group_by(prep_impact_eff_dat, alert_number, alert_type, pop_brk) %>%
  dplyr::reframe(quibble2(eff)) %>%
  tidyr::pivot_wider(names_from = eff_q, names_glue = "{.value}_{eff_q}", values_from = eff) %>%
  dplyr::full_join(dplyr::select(means_impact_eff_dat, alert_number, alert_type, pop_brk, eff_mean), by = c("alert_number", "alert_type", "pop_brk")) %>%
  prepare_fig_dat()

fig_ppv_dat <- dplyr::group_by(prep_ppv_dat, alert_number, alert_type, pop_brk) %>%
  dplyr::summarise(alerts = n(), true_positives = sum(true_positives), ppv = round(true_positives/alerts, 2)) %>%
  prepare_fig_dat()

fig_missed_dat <- dplyr::group_by(prep_missed_delay_dat, alert_number, alert_type, pop_brk) %>%
  dplyr::summarise(n_detected = sum(is_triggered)) %>%
  dplyr::left_join(n_outbreaks, by = c("pop_brk")) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(n_missed = n_outbreaks - n_detected,
                missed_prop = n_missed/n_outbreaks) %>%
  prepare_fig_dat()

fig_delay_dat <- dplyr::group_by(prep_missed_delay_dat, alert_number, alert_type, pop_brk) %>%
  dplyr::reframe(quibble2(delay)) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_wider(names_from = delay_q, names_glue = "{.value}_{delay_q}", values_from = delay) %>%
  dplyr::full_join(means_delay_dat, by = c("alert_number", "alert_type", "pop_brk")) %>%
  prepare_fig_dat()

fig_complete_dat <- dplyr::full_join(fig_impact_dat, fig_eff_dat, by = c("alert_number", "alert_type", "pop_brk", "alert_lab")) %>%
  dplyr::full_join(fig_ppv_dat, by = c("alert_number", "alert_type", "pop_brk", "alert_lab")) %>%
  dplyr::full_join(fig_missed_dat, by = c("alert_number", "alert_type", "pop_brk", "alert_lab")) %>%
  dplyr::full_join(fig_delay_dat, by = c("alert_number", "alert_type", "pop_brk", "alert_lab"))

## 30 Oct save as file to compare epidemic/endemic locations and test for statistically significant differences
add_setting_and_save <- function(df, name) {
  df$which_setting <- which_setting
  saveRDS(df, paste0(pre_path, name, which_setting, ".rds"))
}

add_setting_and_save(fig_complete_dat, "compare_significant_missed_ppv_chisquare_")

fig_under_50k <- dplyr::filter(fig_complete_dat, pop_brk == "< 50k")
fig_50k_500k <- dplyr::filter(fig_complete_dat, pop_brk == "[50k, 500k)")
fig_over_500k <- dplyr::filter(fig_complete_dat, pop_brk == "\U2265 500k")

```

```{r}
#| label: prep_fig_data_cCh
#| echo: false

fig_impact_dat_cCh <- dplyr::group_by(prep_impact_eff_dat_cCh, alert_number, alert_type, pop_brk) %>%
  dplyr::reframe(quibble2(impact)) %>%
  tidyr::pivot_wider(names_from = impact_q, names_glue = "{.value}_{impact_q}", values_from = impact) %>%
  dplyr::full_join(dplyr::select(means_impact_eff_dat_cCh, alert_number, alert_type, pop_brk, impact_mean), by = c("alert_number", "alert_type", "pop_brk")) %>%
  prepare_fig_dat()

fig_eff_dat_cCh <- dplyr::group_by(prep_impact_eff_dat_cCh, alert_number, alert_type, pop_brk) %>%
  dplyr::reframe(quibble2(eff)) %>%
  tidyr::pivot_wider(names_from = eff_q, names_glue = "{.value}_{eff_q}", values_from = eff) %>%
  dplyr::full_join(dplyr::select(means_impact_eff_dat_cCh, alert_number, alert_type, pop_brk, eff_mean), by = c("alert_number", "alert_type", "pop_brk")) %>%
  prepare_fig_dat()

fig_ppv_dat_cCh <- dplyr::group_by(prep_ppv_dat_cCh, alert_number, alert_type, pop_brk) %>%
  dplyr::summarise(alerts = n(), true_positives = sum(true_positives), ppv = round(true_positives/alerts, 2)) %>%
  prepare_fig_dat()

fig_missed_dat_cCh <- dplyr::group_by(prep_missed_delay_dat_cCh, alert_number, alert_type, pop_brk) %>%
  dplyr::summarise(n_detected = sum(is_triggered)) %>%
  dplyr::left_join(n_outbreaks_cCh, by = c("pop_brk")) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(n_missed = n_outbreaks - n_detected,
                missed_prop = n_missed/n_outbreaks) %>%
  prepare_fig_dat()

fig_delay_dat_cCh <- dplyr::group_by(prep_missed_delay_dat_cCh, alert_number, alert_type, pop_brk) %>%
  dplyr::reframe(quibble2(delay)) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_wider(names_from = delay_q, names_glue = "{.value}_{delay_q}", values_from = delay) %>%
  dplyr::full_join(means_delay_dat_cCh, by = c("alert_number", "alert_type", "pop_brk")) %>%
  prepare_fig_dat()

fig_complete_dat_cCh <- dplyr::full_join(fig_impact_dat_cCh, fig_eff_dat_cCh, by = c("alert_number", "alert_type", "pop_brk", "alert_lab")) %>%
  dplyr::full_join(fig_ppv_dat_cCh, by = c("alert_number", "alert_type", "pop_brk", "alert_lab")) %>%
  dplyr::full_join(fig_missed_dat_cCh, by = c("alert_number", "alert_type", "pop_brk", "alert_lab")) %>%
  dplyr::full_join(fig_delay_dat_cCh, by = c("alert_number", "alert_type", "pop_brk", "alert_lab"))

add_setting_and_save(fig_complete_dat_cCh, paste0("compare_significant_missed_ppv_chisquare_cCh_", cCh_threshold, "_"))

fig_under_50k_cCh <- dplyr::filter(fig_complete_dat_cCh, pop_brk == "< 50k")
fig_50k_500k_cCh <- dplyr::filter(fig_complete_dat_cCh, pop_brk == "[50k, 500k)")
fig_over_500k_cCh <- dplyr::filter(fig_complete_dat_cCh, pop_brk == "\U2265 500k")

```

```{r}
#| label: func-figs
#| echo: false

impact_lineplot <- function(df, title_lab = "Impact", col_pal = "Dark2"){
  df_mean <- dplyr::summarise(df, impact_total_mean = mean(impact_mean, na.rm = TRUE))
  print(paste("Impact mean:", df_mean$impact_total_mean))

  ggplot2::ggplot(df, aes(y = alert_lab)) +
    geom_point(aes(x = impact_mean, colour = alert_type)) +
    geom_point(aes(x = impact_q5, colour = alert_type), shape = "x", size = 4) +
    geom_linerange(aes(xmin = impact_q25, xmax = impact_q75, colour = alert_type)) +
    geom_vline(data = df_mean, aes(xintercept = impact_total_mean), colour = "black", linetype = 2, alpha = 0.7) +
    scale_colour_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Suspected Cases", labels = scales::label_number(scale = 0.001, suffix = "k")) +
    theme_bw() +
    theme(legend.position = "none", axis.title.y = element_blank(), plot.margin = unit(c(1,1,1,1), "mm")) +
    ggtitle(title_lab)
}

eff_lineplot <- function(df, title_lab = "Efficiency", col_pal = "Dark2"){
  df_mean <- dplyr::summarise(df, eff_total_mean = mean(eff_mean, na.rm = TRUE))
  print(paste("Efficiency mean:", df_mean$eff_total_mean))

  ggplot2::ggplot(df, aes(y = alert_lab)) +
    geom_point(aes(x = eff_mean, colour = alert_type)) +
    geom_point(aes(x = eff_q5, colour = alert_type), shape = "x", size = 4) +
    geom_linerange(aes(xmin = eff_q25, xmax = eff_q75, colour = alert_type)) +
    geom_vline(data = df_mean, aes(xintercept = eff_total_mean), colour = "black", linetype = 2, alpha = 0.7) +
    scale_colour_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Cases/1k target pop", labels = scales::label_number(big.mark = "", accuracy = 1), limits = c(0, max(df$eff_mean, na.rm = TRUE)*1.1)) +
    theme_bw() +
    theme(legend.position = "none", axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank(), plot.margin = unit(c(1,1,1,1), "mm")) +
    ggtitle(title_lab)
}

ppv_barplot <- function(df, title_lab = "PPV", col_pal = "Dark2"){
  df_mean <- dplyr::summarise(df, ppv_mean = mean(ppv, na.rm = TRUE))
  print(paste("PPV mean:", df_mean$ppv_mean))

  ggplot2::ggplot(df, aes(x = ppv, y = alert_lab)) +
    geom_col(aes(fill = alert_type)) +
    geom_vline(data = df_mean, aes(xintercept = ppv_mean), colour = "black", linetype = 2, alpha = 0.7) +
    scale_x_continuous("PPV", labels = scales::label_number(accuracy = 0.1), breaks = seq(0, 1, by = 0.2), limits = c(0, 1)) +
    scale_y_discrete("Alert Type") +
    scale_fill_brewer("Alert Type", palette = col_pal) +
    theme_bw() +
    theme(legend.position = "none", axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank(), plot.margin = unit(c(1,1,1,3), "mm")) +
    ggtitle(title_lab)
}

missed_barplot <- function(df, title_lab = "Missed", col_pal = "Dark2"){
  df_mean <- dplyr::summarise(df, missed_mean = mean(missed_prop, na.rm = TRUE))
  print(paste("Missed mean:", df_mean$missed_mean))

  ggplot2::ggplot(df, aes(x = missed_prop, y = alert_lab)) +
    geom_col(aes(fill = alert_type)) +
    geom_vline(data = df_mean, aes(xintercept = missed_mean), colour = "black", linetype = 2, alpha = 0.7) +
    scale_fill_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Prop of outbreaks", labels = scales::label_number(accuracy = 0.1), breaks = seq(0, 1, by = 0.2), limits = c(0, 1)) +
    theme_bw() +
    theme(legend.position = "none", axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), plot.margin = unit(c(1,1,1,1), "mm")) +
    ggtitle(title_lab)
}

delay_lineplot <- function(df, title_lab = "Delay", col_pal = "Dark2"){
  df_mean <- dplyr::summarise(df, delay_total_mean = mean(delay_mean, na.rm = TRUE))
  print(paste("Delay mean:", df_mean$delay_total_mean))

  ggplot2::ggplot(df, aes(y = alert_lab)) +
    geom_point(aes(x = delay_mean, colour = alert_type)) +
    geom_point(aes(x = delay_q5, colour = alert_type), shape = "x", size = 4) +
    geom_vline(data = df_mean, aes(xintercept = delay_total_mean), colour = "black", linetype = 2, alpha = 0.7) +
    geom_linerange(aes(xmin = delay_q25, xmax = delay_q75, colour = alert_type)) +
    scale_colour_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Weeks of delay", labels = scales::label_number()) +
    theme_bw() +
    theme(legend.position = "right", axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), plot.margin = unit(c(1,1,1,1), "mm")) +
    ggtitle(title_lab)
}
```



```{r}
#| label: boxplots-and-barcharts-mainpop
#| echo: true
#| fig.width: 10

col_pal <- "Dark2"

fig_impact_raw <- prep_impact_eff_dat %>% prepare_fig_dat()
fig_eff_raw    <- prep_impact_eff_dat %>% prepare_fig_dat()
fig_delay_raw  <- prep_missed_delay_dat %>% prepare_fig_dat()

## 30 Oct save as file to compare epidemic/endemic locations and test for statistically significant differences
add_setting_and_save(fig_impact_raw, "compare_significant_impact_testmeans_")
add_setting_and_save(fig_eff_raw, "compare_significant_efficiency_testmeans_")
add_setting_and_save(fig_delay_raw, "compare_significant_delay_testmeans_")

pop_groups <- c("< 50k", "[50k, 500k)", "\U2265 500k")

impact_data_list <- lapply(pop_groups, function(pg) fig_impact_raw %>% filter(pop_brk == pg))
names(impact_data_list) <- pop_groups

eff_data_list <- lapply(pop_groups, function(pg) fig_eff_raw %>% filter(pop_brk == pg))
names(eff_data_list) <- pop_groups

delay_data_list <- lapply(pop_groups, function(pg) fig_delay_raw %>% filter(pop_brk == pg))
names(delay_data_list) <- pop_groups

## function to add vertical dashed line for overall mean
add_mean_line <- function(df, xvar) {
  overall_mean <- mean(df[[xvar]], na.rm = TRUE)
  ymin <- 0.5
  ymax <- length(unique(df$alert_lab)) + 0.5
  geom_segment(aes(x = overall_mean, xend = overall_mean, y = ymin, yend = ymax),
               linetype = 2, color = "black", alpha = 0.7, inherit.aes = FALSE)
}

center_x_on_mean <- function(df, xvar) {
  x_mean <- mean(df[[xvar]], na.rm = TRUE)
  x_lower <- quantile(df[[xvar]], 0.05, na.rm = TRUE)
  x_upper <- quantile(df[[xvar]], 0.95, na.rm = TRUE)
  xlim_min <- min(x_lower, x_mean - 0.2 * (x_upper - x_lower))
  xlim_max <- max(x_upper, x_mean + 0.2 * (x_upper - x_lower))
  
  c(xlim_min, xlim_max)
}

## impact boxplot
create_impact_boxplot <- function(df, pop_label) {
  xlim_vals <- center_x_on_mean(df, "impact")
  ggplot(df, aes(x = impact, y = alert_lab, fill = alert_type)) +
    geom_boxplot(outlier.shape = 21, outlier.fill = NA) +
    stat_summary(fun = mean, geom = "point", size = 2.5, fill = "white") +
    add_mean_line(df, "impact") +
    scale_fill_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Suspected Cases", labels = label_number(scale = 0.001, suffix = "k")) +
    coord_cartesian(xlim = xlim_vals) +
    theme_bw() +
    theme(
      legend.position = "none",
      axis.title.y = element_blank(),
      axis.text.y = element_text(size = 9, color = "black"),
      axis.ticks.y = element_line(),
      plot.margin = unit(c(1,1,1,1), "mm")
    ) +
    ggtitle("Impact")
}

## efficiency boxplot
create_eff_boxplot <- function(df, xvar, metric_label, pop_label) {
  xlim_vals <- center_x_on_mean(df, xvar)
  ggplot(df, aes_string(x = xvar, y = "alert_lab", fill = "alert_type")) +
    geom_boxplot(outlier.shape = 21, outlier.fill = NA) +
    stat_summary(fun = mean, geom = "point", size = 2.5, fill = "white") +
    add_mean_line(df, xvar) +
    scale_fill_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Cases/1k target pop", labels = label_number(big.mark = "", accuracy = 1)) +
    coord_cartesian(xlim = xlim_vals) +
    theme_bw() +
    theme(
      legend.position = "none",
      axis.title.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.text.y = element_blank(),
      plot.margin = unit(c(1,1,1,1), "mm")
    ) +
    ggtitle(metric_label)
}

## delay boxplot (legend visible)
create_delay_boxplot <- function(df, xvar, metric_label, pop_label) {
  xlim_vals <- center_x_on_mean(df, xvar)
  ggplot(df, aes_string(x = xvar, y = "alert_lab", fill = "alert_type")) +
    geom_boxplot(outlier.shape = 21, outlier.fill = NA) +
    stat_summary(fun = mean, geom = "point", size = 2.5, fill = "white") +
    add_mean_line(df, xvar) +
    scale_fill_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Weeks of delay", labels = label_number(big.mark = "", accuracy = 1)) +
    coord_cartesian(xlim = xlim_vals) +
    theme_bw() +
    theme(
      legend.position = "right",
      axis.title.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.text.y = element_blank(),
      plot.margin = unit(c(1,1,1,1), "mm")
    ) +
    ggtitle(metric_label)
}

## boxplots per population group
impact_boxplots <- lapply(impact_data_list, create_impact_boxplot)
eff_boxplots    <- lapply(eff_data_list, function(df) create_eff_boxplot(df, "eff", "Efficiency", NULL))
delay_boxplots  <- lapply(delay_data_list, function(df) create_delay_boxplot(df, "delay", "Delay", NULL))

ws <- c(4, 4, 2, 2, 4)

gridExtra::grid.arrange(
  impact_boxplots[["[50k, 500k)"]],
  eff_boxplots[["[50k, 500k)"]],
  ppv_barplot(fig_50k_500k),
  missed_barplot(fig_50k_500k),
  delay_boxplots[["[50k, 500k)"]],
  nrow = 1, widths = ws, padding = unit(0.6, "line")
)

```

```{r}
#| label: boxplots-and-barcharts-mainpop_cCh
#| echo: true
#| fig.width: 10

col_pal <- "Dark2"

fig_impact_raw_cCh <- prep_impact_eff_dat_cCh %>% prepare_fig_dat()
fig_eff_raw_cCh    <- prep_impact_eff_dat_cCh %>% prepare_fig_dat()
fig_delay_raw_cCh  <- prep_missed_delay_dat_cCh %>% prepare_fig_dat()

## 30 Oct save as file to compare epidemic/endemic locations and test for statistically significant differences
add_setting_and_save(fig_impact_raw_cCh, paste0("compare_significant_impact_testmeans_cCh_", cCh_threshold, "_"))
add_setting_and_save(fig_eff_raw_cCh,  paste0("compare_significant_efficiency_testmeans_cCh_", cCh_threshold, "_"))
add_setting_and_save(fig_delay_raw_cCh,  paste0("compare_significant_delay_testmeans_cCh_", cCh_threshold, "_"))

pop_groups <- c("< 50k", "[50k, 500k)", "\U2265 500k")

impact_data_list_cCh <- lapply(pop_groups, function(pg) fig_impact_raw_cCh %>% filter(pop_brk == pg))
names(impact_data_list_cCh) <- pop_groups

eff_data_list_cCh <- lapply(pop_groups, function(pg) fig_eff_raw_cCh %>% filter(pop_brk == pg))
names(eff_data_list_cCh) <- pop_groups

delay_data_list_cCh <- lapply(pop_groups, function(pg) fig_delay_raw_cCh %>% filter(pop_brk == pg))
names(delay_data_list_cCh) <- pop_groups

## function to add vertical dashed line for overall mean
add_mean_line <- function(df, xvar) {
  overall_mean <- mean(df[[xvar]], na.rm = TRUE)
  ymin <- 0.5
  ymax <- length(unique(df$alert_lab)) + 0.5
  geom_segment(aes(x = overall_mean, xend = overall_mean, y = ymin, yend = ymax),
               linetype = 2, color = "black", alpha = 0.7, inherit.aes = FALSE)
}

center_x_on_mean <- function(df, xvar) {
  x_mean <- mean(df[[xvar]], na.rm = TRUE)
  x_lower <- quantile(df[[xvar]], 0.05, na.rm = TRUE)
  x_upper <- quantile(df[[xvar]], 0.95, na.rm = TRUE)
  xlim_min <- min(x_lower, x_mean - 0.2 * (x_upper - x_lower))
  xlim_max <- max(x_upper, x_mean + 0.2 * (x_upper - x_lower))
  
  c(xlim_min, xlim_max)
}

## impact boxplot
create_impact_boxplot <- function(df, pop_label) {
  xlim_vals <- center_x_on_mean(df, "impact")
  ggplot(df, aes(x = impact, y = alert_lab, fill = alert_type)) +
    geom_boxplot(outlier.shape = 21, outlier.fill = NA) +
    stat_summary(fun = mean, geom = "point", size = 2.5, fill = "white") +
    add_mean_line(df, "impact") +
    scale_fill_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Suspected Cases", labels = label_number(scale = 0.001, suffix = "k")) +
    coord_cartesian(xlim = xlim_vals) +
    theme_bw() +
    theme(
      legend.position = "none",
      axis.title.y = element_blank(),
      axis.text.y = element_text(size = 9, color = "black"),
      axis.ticks.y = element_line(),
      plot.margin = unit(c(1,1,1,1), "mm")
    ) +
    ggtitle("Impact")
}

## efficiency boxplot
create_eff_boxplot <- function(df, xvar, metric_label, pop_label) {
  xlim_vals <- center_x_on_mean(df, xvar)
  ggplot(df, aes_string(x = xvar, y = "alert_lab", fill = "alert_type")) +
    geom_boxplot(outlier.shape = 21, outlier.fill = NA) +
    stat_summary(fun = mean, geom = "point", size = 2.5, fill = "white") +
    add_mean_line(df, xvar) +
    scale_fill_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Cases/1k target pop", labels = label_number(big.mark = "", accuracy = 1)) +
    coord_cartesian(xlim = xlim_vals) +
    theme_bw() +
    theme(
      legend.position = "none",
      axis.title.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.text.y = element_blank(),
      plot.margin = unit(c(1,1,1,1), "mm")
    ) +
    ggtitle(metric_label)
}

## delay boxplot (legend visible)
create_delay_boxplot <- function(df, xvar, metric_label, pop_label) {
  xlim_vals <- center_x_on_mean(df, xvar)
  ggplot(df, aes_string(x = xvar, y = "alert_lab", fill = "alert_type")) +
    geom_boxplot(outlier.shape = 21, outlier.fill = NA) +
    stat_summary(fun = mean, geom = "point", size = 2.5, fill = "white") +
    add_mean_line(df, xvar) +
    scale_fill_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Weeks of delay", labels = label_number(big.mark = "", accuracy = 1)) +
    coord_cartesian(xlim = xlim_vals) +
    theme_bw() +
    theme(
      legend.position = "right",
      axis.title.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.text.y = element_blank(),
      plot.margin = unit(c(1,1,1,1), "mm")
    ) +
    ggtitle(metric_label)
}

## boxplots per population group
impact_boxplots_cCh <- lapply(impact_data_list_cCh, create_impact_boxplot)
eff_boxplots_cCh    <- lapply(eff_data_list_cCh, function(df) create_eff_boxplot(df, "eff", "Efficiency", NULL))
delay_boxplots_cCh  <- lapply(delay_data_list_cCh, function(df) create_delay_boxplot(df, "delay", "Delay", NULL))

ws <- c(4, 4, 2, 2, 4)

gridExtra::grid.arrange(
  impact_boxplots_cCh[["[50k, 500k)"]],
  eff_boxplots_cCh[["[50k, 500k)"]],
  ppv_barplot(fig_50k_500k_cCh),
  missed_barplot(fig_50k_500k_cCh),
  delay_boxplots_cCh[["[50k, 500k)"]],
  nrow = 1, widths = ws, padding = unit(0.6, "line")
)

```


### Dimension distributions, 50-500k without filtering for confirmed cases

```{r}
#| label: figs-50k-500k
#| fig-cap: Utility score dimensions for 50k-500k population group. The underlying metric for each utility score dimension is displayed for all alert definitions. The vertical black dashed line depicts the mean across alert definitions and colors indicate the alert type. For line plots, 'X' indicates median of the distribution, a point indicates mean, and line indicates IQR.
#| fig-width: 9
#| fig-height: 6
#| echo: true

ws <- c(1, .7, .7, .7, 1.25)

gridExtra::grid.arrange(impact_lineplot(fig_50k_500k), eff_lineplot(fig_50k_500k), ppv_barplot(fig_50k_500k), missed_barplot(fig_50k_500k), delay_lineplot(fig_50k_500k), nrow=1, widths = ws, padding = unit(0.6, "line"))

```

### Dimension distributions, 50-500k w confirmed cases in the alert period higher than the threshold

```{r}
#| label: figs-50k-500k_cCh
#| fig-cap: Utility score dimensions for 50k-500k population group. The underlying metric for each utility score dimension is displayed for all alert definitions. The vertical black dashed line depicts the mean across alert definitions and colors indicate the alert type. For line plots, 'X' indicates median of the distribution, a point indicates mean, and line indicates IQR.
#| fig-width: 9
#| fig-height: 6
#| echo: true

ws <- c(1, .7, .7, .7, 1.25)

gridExtra::grid.arrange(impact_lineplot(fig_50k_500k_cCh), eff_lineplot(fig_50k_500k_cCh), ppv_barplot(fig_50k_500k_cCh), missed_barplot(fig_50k_500k_cCh), delay_lineplot(fig_50k_500k_cCh), nrow=1, widths = ws, padding = unit(0.6, "line"))
```

### Supplement

#### Dimension distributions, \<50k without filtering for confirmed cases

```{r}
#| label: figs-under-50k
#| fig-cap: Utility score dimensions for under 50k population group. The underlying metric for each utility score dimension is displayed for all alert definitions. The vertical black dashed line depicts the mean across alert definitions and colors indicate the alert type. For line plots, 'X' indicates median of the distribution, a point indicates mean, and line indicates IQR.
#| fig-width: 9
#| fig-height: 6
#| echo: true

gridExtra::grid.arrange(impact_lineplot(fig_under_50k), eff_lineplot(fig_under_50k), ppv_barplot(fig_under_50k), missed_barplot(fig_under_50k), delay_lineplot(fig_under_50k), nrow=1, widths = ws, padding = unit(0.6, "line"))

```


#### Dimension distributions, \<50k w confirmed cases in the alert period higher than the threshold

```{r}
#| label: figs-under-50k_cCh
#| fig-cap: Utility score dimensions for under 50k population group. The underlying metric for each utility score dimension is displayed for all alert definitions. The vertical black dashed line depicts the mean across alert definitions and colors indicate the alert type. For line plots, 'X' indicates median of the distribution, a point indicates mean, and line indicates IQR.
#| fig-width: 9
#| fig-height: 6
#| echo: true

gridExtra::grid.arrange(impact_lineplot(fig_under_50k_cCh), eff_lineplot(fig_under_50k_cCh), ppv_barplot(fig_under_50k_cCh), missed_barplot(fig_under_50k_cCh), delay_lineplot(fig_under_50k_cCh), nrow=1, widths = ws, padding = unit(0.6, "line"))

```

#### Dimension distributions \>500k without filtering for confirmed cases

```{r}
#| label: figs-over-500k
#| fig-cap: Utility score dimensions for over 500k population group. The underlying metric for each utility score dimension is displayed for all alert definitions. The vertical black dashed line depicts the mean across alert definitions and colors indicate the alert type. For line plots, 'X' indicates median of the distribution, a point indicates mean, and line indicates IQR.
#| fig-width: 9
#| fig-height: 6
#| echo: true

gridExtra::grid.arrange(impact_lineplot(fig_over_500k), eff_lineplot(fig_over_500k), ppv_barplot(fig_over_500k), missed_barplot(fig_over_500k), delay_lineplot(fig_over_500k), nrow=1, widths = ws, padding = unit(0.6, "line"))
```

#### Dimension distributions \>500k w confirmed cases in the alert period higher than the threshold

```{r}
#| label: figs-over-500k_cCh
#| fig-cap: Utility score dimensions for over 500k population group. The underlying metric for each utility score dimension is displayed for all alert definitions. The vertical black dashed line depicts the mean across alert definitions and colors indicate the alert type. For line plots, 'X' indicates median of the distribution, a point indicates mean, and line indicates IQR.
#| fig-width: 9
#| fig-height: 6
#| echo: true

gridExtra::grid.arrange(impact_lineplot(fig_over_500k_cCh), eff_lineplot(fig_over_500k_cCh), ppv_barplot(fig_over_500k_cCh), missed_barplot(fig_over_500k_cCh), delay_lineplot(fig_over_500k_cCh), nrow=1, widths = ws, padding = unit(0.6, "line"))
```


## Sensitivity Analysis across thresholds

```{r}
#| label: cCh-comparison-plots
#| eval: true
#| echo: false
#| warning: false
#| message: false
#| fig.width: 14
#| fig.height: 8
#| results: asis

which_setting <- "epidemic" 
thresholds    <- c(0, 1, 2, 5, 10, 20)

## absolute path to load data, there is some problem with loading from relative dirs
#pre_path <- "C:/Users/chalam/Documents/outbreaks_project/main_ecl_dec2024/inputs"

for (popgroup in pop_groups){
  for(th in thresholds){
  f <- file.path(pre_path, paste0("compare_significant_missed_ppv_chisquare_cCh_", th, "_", which_setting, ".rds"))
  if(!file.exists(f)) stop("File not found: ", f)
  message("Found: ", f)
}

load_cCh_all <- function(th) {
  
  read_file <- function(prefix){
    fname <- file.path(pre_path, paste0(prefix, th, "_", which_setting, ".rds"))
    if(!file.exists(fname)) stop("Cannot open file: ", fname)
    readRDS(fname) %>% mutate(threshold = th)
  }
  
  list(
    complete   = read_file("compare_significant_missed_ppv_chisquare_cCh_"),
    impact_raw = read_file("compare_significant_impact_testmeans_cCh_"),
    eff_raw    = read_file("compare_significant_efficiency_testmeans_cCh_"),
    delay_raw  = read_file("compare_significant_delay_testmeans_cCh_")
  )
}

all_loaded <- map(thresholds, function(th) {
  message("Loading threshold: ", th)
  load_cCh_all(th)
})

combined_complete   <- map_df(all_loaded, "complete")
combined_impact_raw <- map_df(all_loaded, "impact_raw")
combined_eff_raw    <- map_df(all_loaded, "eff_raw")
combined_delay_raw  <- map_df(all_loaded, "delay_raw")

make_factor_safe <- function(df){
  if(nrow(df) > 0) df %>% mutate(threshold = factor(threshold, levels = thresholds))
  else df
}

## only looking at the required population group
combined_complete   <- make_factor_safe(combined_complete %>% filter(pop_brk == popgroup )) 
combined_impact_raw <- make_factor_safe(combined_impact_raw %>% filter(pop_brk == popgroup ))
combined_eff_raw    <- make_factor_safe(combined_eff_raw %>% filter(pop_brk == popgroup ))
combined_delay_raw  <- make_factor_safe(combined_delay_raw %>% filter(pop_brk == popgroup ))

message("All data loaded successfully")

## table
table_combined <- combined_complete %>%
  filter(pop_brk == popgroup) %>%
  select(alert_number, alert_type, alert_lab, impact_mean, eff_mean, ppv, missed_prop, delay_mean, threshold) %>%
  distinct() 

# table_pretty <- table_combined %>%
#   select(alert_type, alert_lab, impact_mean, eff_mean, ppv, missed_prop, delay_mean, threshold) %>%
#   distinct() %>%
#   arrange(alert_type, alert_lab) %>%
#   rename(
#     "Alert Type" = alert_type,
#     "Alert Definition" = alert_lab,
#     "Mean Impact" = impact_mean,
#     "Mean Effectiveness" = eff_mean,
#     "PPV" = ppv,
#     "Missed Proportion" = missed_prop,
#     "Mean Delay" = delay_mean,
#     "Threshold" = threshold
#   )

table_pretty <- table_combined %>%
  select(alert_type, alert_lab, impact_mean, eff_mean, ppv, missed_prop, delay_mean, threshold) %>%
  distinct() %>%
  arrange(alert_type, alert_lab) %>%
  rename(
    "Alert Type" = alert_type,
    "Alert Definition" = alert_lab,
    "Impact" = impact_mean,
    "Efficiency" = eff_mean,
    "PPV" = ppv,
    "Missed" = missed_prop,
    "Timeliness" = delay_mean,
    "Threshold" = threshold
  )

table_print <- table_pretty %>%
  kable(
    format = "html",       
    caption = "Dimension Metrics Across cCh Thresholds",
    digits = 2
  ) %>%
  kable_styling(
    full_width = FALSE,
    position = "center"
  ) 

print(table_print)
  
## scatterplot

df_long <- table_pretty %>%
  pivot_longer(
    cols = c("Impact", "Efficiency", "PPV", "Missed", "Timeliness"),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  mutate(
    Metric = factor(Metric, levels = c("Impact", "Efficiency", "PPV", "Missed", "Timeliness"))
  )


metric_scatterplot <- ggplot(df_long, aes(x = Threshold, y = Value)) +
  geom_point() +
  facet_grid(Metric ~ `Alert Definition` + `Alert Type`, scales = "free_y") +
  theme_bw() +
  labs(
    x = "Threshold", 
    y = "Metric Value", 
    title = paste0("Metrics Across Thresholds by Alert Type and Definition , ",popgroup )
  )

print(metric_scatterplot)

baseline <- table_pretty %>%
  filter(
    str_detect(`Alert Definition`, regex("≥ 50 weekly", ignore_case = TRUE)),
    Threshold == 0
  ) %>%
  pivot_longer(
    cols = c("Impact", "Efficiency", "PPV",
             "Missed", "Timeliness"),
    names_to = "Metric",
    values_to = "Baseline"
  ) %>% 
  select(Metric, Baseline) %>%
  mutate(
    Metric = factor(Metric, levels = c("Impact", "Efficiency", "PPV", "Missed", "Timeliness"))
  )


df_long <- table_pretty %>%
  pivot_longer(
    cols = c("Impact", "Efficiency", "PPV",
             "Missed", "Timeliness"),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  mutate(
    Metric = factor(Metric, levels = c("Impact", "Efficiency", "PPV", "Missed", "Timeliness"))
  )

df_filtered <- df_long %>%
  filter(
    !str_detect(`Alert Definition`, regex("≥ 50 weekly", ignore_case = TRUE))
  )

df_raw <- df_filtered %>%
  left_join(baseline, by = "Metric") %>%
  mutate(
    Threshold = as.numeric(as.character(Threshold))
  )

compare_to_baseline_plot <- ggplot(df_raw, aes(
  x = Value, 
  y = `Alert Definition`,
  color = Threshold
)) +
  geom_point(size = 3, alpha = 0.9) +
  scale_color_gradient(
    low = "#b2df8a",   
    high = "#1f78b4"   
  ) +
  geom_vline(
    data = baseline,
    aes(xintercept = Baseline),
    linetype = "dashed",
    color = "grey40"
  ) +
  facet_wrap(~ Metric, scales = "free_x") +
  theme_minimal(base_size = 16) +
  labs(
    title = paste0("Utility Dimension Metrics Across cCh Thresholds , ", popgroup),
    subtitle = "Baseline (≥50 weekly at threshold 0) shown as dashed line",
    x = "",
    y = "Alert Definition",
    color = "Threshold"
  )

print(compare_to_baseline_plot)

}

```


```{r}
#| eval: true

compare_case_alerts_to_50weekly_blackline <- function(
  popgroup,
  which_setting = "epidemic",
  pre_path = "C:/Users/chalam/Documents/outbreaks_project/main_ecl_dec2024/inputs",
  thresholds = 0:5
) {

  # Helper to load data per threshold
  load_cCh_all <- function(th) {
    fname <- file.path(pre_path, paste0("compare_significant_missed_ppv_chisquare_cCh_", th, "_", which_setting, ".rds"))
    if(!file.exists(fname)) stop("Cannot open file: ", fname)
    readRDS(fname) %>% mutate(threshold = th)
  }

  # Load and combine
  combined_complete <- map_df(thresholds, function(th) {
    message("Loading threshold: ", th)
    load_cCh_all(th)
  })

  # Filter to population and case alerts
  df_cases <- combined_complete %>%
    filter(pop_brk == popgroup, alert_type == "case") %>%
    mutate(threshold = factor(threshold, levels = thresholds)) %>%
    select(alert_lab, impact_mean, eff_mean, ppv, missed_prop, delay_mean, threshold)

  if (nrow(df_cases) == 0) stop("No 'case' alerts found for this population group.")

baseline <- df_cases %>%
    filter(alert_lab == "≥ 50 weekly", threshold == 0) %>%  # no space
    pivot_longer(
        cols = c(impact_mean, eff_mean, ppv, missed_prop, delay_mean),
        names_to = "Metric",
        values_to = "Baseline"
    ) %>%
    mutate(
        Metric = recode(
            Metric,
            impact_mean = "Impact",
            eff_mean    = "Efficiency",
            ppv         = "PPV",
            missed_prop = "Missed",
            delay_mean  = "Timeliness"
        )
    ) %>%
    select(Metric, Baseline) %>%
  mutate(
    Metric = factor(Metric, levels = c("Impact", "Efficiency", "PPV", "Missed", "Timeliness"))
  )

# Filter out baseline from comparison points
df_compare <- df_cases %>%
    filter(!(alert_lab == "≥ 50 weekly")) %>%  # no space
    pivot_longer(
        cols = c(impact_mean, eff_mean, ppv, missed_prop, delay_mean),
        names_to = "Metric",
        values_to = "Value"
    ) %>%
    left_join(baseline, by = "Metric") %>%
    mutate(threshold = as.numeric(as.character(threshold))) %>%
    mutate(Metric = recode(
        Metric,
        impact_mean = "Impact",
        eff_mean    = "Efficiency",
        ppv         = "PPV",
        missed_prop = "Missed",
        delay_mean  = "Timeliness"
    )) %>%
  mutate(
    Metric = factor(Metric, levels = c("Impact", "Efficiency", "PPV", "Missed", "Timeliness"))
  )


  # Plot
  p <- ggplot(df_compare, aes(x = Value, y = alert_lab, color = threshold)) +
    geom_point(size = 3, alpha = 0.9) +
    scale_color_gradient(low = "#b2df8a", high = "#1f78b4") +
    geom_vline(
      data = baseline,
      aes(xintercept = Baseline),
      linetype = "dashed",
      color = "black",
      size = 0.8
    ) +
    facet_wrap(~ Metric, scales = "free_x") +
    theme_minimal() +
    labs(
      title = paste0("Case Alert Metrics Across Thresholds 0–5, ", popgroup),
      subtitle = "Baseline: ≥50 weekly at threshold 0 (black dashed line)",
      x = "Metric Value",
      y = "Alert Definition",
      color = "Threshold"
    )

  print(p)

  invisible(df_compare)
}

# Run the comparison plot
for (pop_group in pop_groups){
  compare_case_alerts_to_50weekly_blackline(popgroup = pop_group)
}


```

## Spatial and temporal coverage across thresholds

### Spatial coverage

```{r}
#| label: cCh-spatial-coverage-plots
#| eval: false

for (thresh in thresholds) {
  
  alerts_w_cCh_threshold <- a_with_counts %>%
    dplyr::select(alert_id, alert_lab, total_confirmed_cases, total_suspected_cases, confirmed_prop, location) %>%
    dplyr::filter(total_confirmed_cases >= thresh)

  outbreak_shapes <- readr::read_rds(file.path(pre_path, "outbreak_shapefiles.rds"))

  alerts_w_cCh_threshold_lctnpr <- alerts_w_cCh_threshold %>%
    dplyr::select(alert_id, location) %>%
    dplyr::left_join(pre_extract, by = "location") %>%
    dplyr::select(alert_id, location, location_period_id, spatial_scale) %>%
    dplyr::distinct()

  if (!inherits(outbreak_shapes, "sf")) {
    outbreak_shapes <- sf::st_as_sf(outbreak_shapes)
  }

  if (is.na(sf::st_crs(outbreak_shapes))) {
    sf::st_crs(outbreak_shapes) <- 4326
  } else if (sf::st_crs(outbreak_shapes) != sf::st_crs(4326)) {
    outbreak_shapes <- sf::st_transform(outbreak_shapes, 4326)
  }

  africa <- giscoR::gisco_get_countries(
    region = "Africa",
    resolution = "3"
  )

  shape_join <- alerts_w_cCh_threshold_lctnpr %>%
    left_join(
      outbreak_shapes,
      by = c("location_period_id" = "lctn_pr")
    ) %>%
    distinct(location, .keep_all = TRUE) %>%
    sf::st_as_sf() %>%
    sf::st_make_valid()

  shape_join <- shape_join %>%
    mutate(scale_group = case_when(
      spatial_scale %in% c("admin1", "admin1 composite") ~ "Admin1",
      spatial_scale %in% c("admin2", "admin2 composite") ~ "Admin2",
      spatial_scale %in% c("admin3", "admin3 composite") ~ "Admin3",
      TRUE ~ spatial_scale
    )) %>%
    mutate(scale_group = factor(scale_group, levels = c("Admin1", "Admin2", "Admin3")))

  scale_colors <- c("Admin1" = "lightblue1",
                  "Admin2" = "lightblue3",
                  "Admin3" = "lightblue4")

## the map
  map <- ggplot() +
    geom_sf(data = africa, fill = "gray95", color = "gray80", linewidth = 0.2) +
    geom_sf(data = shape_join %>% filter(scale_group == "Admin1"), 
          aes(fill = scale_group), color = "white", lwd = 0.2) +
    geom_sf(data = shape_join %>% filter(scale_group == "Admin2"), 
          aes(fill = scale_group), color = "white", lwd = 0.2) +
    geom_sf(data = shape_join %>% filter(scale_group == "Admin3"), 
          aes(fill = scale_group), color = "white", lwd = 0.2, alpha = 0.9) +
    scale_fill_manual(values = scale_colors, name = "Spatial Scale") +
    theme_minimal() +
    labs(
      title = paste0("Alert locations filtered for cCh ",thresh, " ,", which_setting)
    ) +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      plot.subtitle = element_text(size = 12),
      plot.caption = element_text(size = 10, color = "gray40"),
      legend.position = "bottom",
      legend.title = element_text(size = 11, face = "bold"),
      legend.text = element_text(size = 10),
      axis.title = element_blank()
    ) +
    coord_sf(xlim = c(-20, 55), ylim = c(-35, 38), expand = FALSE) 


  print(map)

}

```


```{r}
#| label: cCh-spatial-coverage-plots-2

combined_sf <- list()

for (thresh in thresholds) {

  alerts_w_cCh_threshold <- a_with_counts %>%
    select(alert_id, alert_lab, total_confirmed_cases, total_suspected_cases,
           confirmed_prop, location) %>%
    filter(total_confirmed_cases >= thresh)

  alerts_w_cCh_threshold_lctnpr <- alerts_w_cCh_threshold %>%
    select(alert_id, location) %>%
    left_join(pre_extract, by = "location") %>%
    select(alert_id, location, location_period_id, spatial_scale) %>%
    distinct() %>%
    mutate(threshold = thresh)

  combined_sf[[as.character(thresh)]] <- alerts_w_cCh_threshold_lctnpr
}

combined_df <- bind_rows(combined_sf)

outbreak_shapes <- readr::read_rds(file.path(pre_path, "outbreak_shapefiles.rds"))

if (!inherits(outbreak_shapes, "sf")) {
  outbreak_shapes <- st_as_sf(outbreak_shapes)
}
if (is.na(st_crs(outbreak_shapes))) {
  st_crs(outbreak_shapes) <- 4326
} else if (st_crs(outbreak_shapes) != st_crs(4326)) {
  outbreak_shapes <- st_transform(outbreak_shapes, 4326)
}

combined_sf <- combined_df %>%
  left_join(outbreak_shapes, by = c("location_period_id" = "lctn_pr")) %>%
  distinct(location, threshold, .keep_all = TRUE) %>%
  st_as_sf() %>%
  st_make_valid()

combined_sf <- combined_sf %>%
  mutate(scale_group = case_when(
    spatial_scale %in% c("admin1", "admin1 composite") ~ "Admin1",
    spatial_scale %in% c("admin2", "admin2 composite") ~ "Admin2",
    spatial_scale %in% c("admin3", "admin3 composite") ~ "Admin3",
    TRUE ~ spatial_scale
  )) %>%
  mutate(scale_group = factor(scale_group, levels = c("Admin1", "Admin2", "Admin3")))

africa <- giscoR::gisco_get_countries(region = "Africa", resolution = "3")

scale_colors <- c(
  "Admin1" = "lightblue1",
  "Admin2" = "lightblue3",
  "Admin3" = "lightblue4"
)

ggplot() +
  geom_sf(data = africa, fill = "gray95", color = "gray80", linewidth = 0.25) +

  geom_sf(data = combined_sf,
          aes(fill = scale_group),
          color = "white", lwd = 0.2, alpha = 0.9) +

  scale_fill_manual(values = scale_colors, name = "Administrative Level") +

  facet_wrap(~ threshold) +

  labs(
    title = paste0("Alert locations (", which_setting, ")"),
    subtitle = "One panel per cCh threshold"
  ) +

  theme_minimal() +
  theme(
    strip.text = element_text(size = 12, face = "bold"),
    plot.title = element_text(size = 18, face = "bold"),
    legend.position = "bottom",

    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    panel.grid = element_blank()
  ) +

  coord_sf(xlim = c(-20, 55), ylim = c(-35, 38), expand = FALSE)


```
### Temporal coverage

```{r}
#| label: cCh-temporal-coverage-plots
#| fig.width: 20
#| fig.height: 20

global_week_index <- a_with_counts %>%
  mutate(
    iso_year = isoyear(TL),
    iso_week = isoweek(TL),
    yearweek = paste0(iso_year, "-W", sprintf("%02d", iso_week))
  ) %>%
  distinct(yearweek) %>%
  arrange(yearweek) %>%
  pull(yearweek)

# x axis breaks every 30 weeks globally
global_week_breaks <- global_week_index[seq(1, length(global_week_index), 30)]

for (thresh in thresholds) {

  alerts_w_cCh_threshold <- a_with_counts %>%
    select(alert_id, alert_lab, TL, total_confirmed_cases, total_suspected_cases,
           confirmed_prop, location) %>%
    filter(total_confirmed_cases >= thresh)

  alerts_w_cCh_threshold_temp <- alerts_w_cCh_threshold %>%
    select(alert_id, TL, alert_lab, location) %>%
    left_join(pre_extract, by = c("location", "TL")) %>%
    select(alert_id, alert_lab, TL, TR) %>%
    distinct()

  weekly_counts <- alerts_w_cCh_threshold_temp %>%
    mutate(
      iso_year = isoyear(TL),
      iso_week = isoweek(TL),
      yearweek = paste0(iso_year, "-W", sprintf("%02d", iso_week))
    ) %>%
    count(alert_lab, yearweek)

  # fill missing weeks using the global week index
  weekly_complete <- weekly_counts %>%
    complete(alert_lab, yearweek = global_week_index, fill = list(n = 0))

  temp_density_plot <- ggplot(weekly_complete, aes(x = yearweek, y = 1, fill = n)) +
    geom_tile() +
    facet_wrap(~alert_lab) +
    scale_fill_viridis_c(limits = c(0, 100)) +
    scale_x_discrete(breaks = global_week_breaks) +   
    theme_minimal(base_size = 20) +
    theme(
      axis.text.y = element_blank(),
      axis.title.y = element_blank(),
      panel.grid = element_blank(),
      axis.text.x = element_text(angle = 90, hjust = 1, size = 7)
    ) +
    labs(
      title = paste0("Weekly Alert Density for cCh threshold: ", thresh),
      x = "ISO Week",
      fill = "Alert Count"
    )

  print(temp_density_plot)
}


```
