---
title: "Alert Utility Score"
format:
  html:
    toc: true
    toc-depth: 4
    number-sections: true
    number-depth: 4
    code-fold: true
    embed-resources: true
    theme:
      - custom.scss
params:
  testing: false
  test_cntry: "COD"
  remove_censored: false
  evalperiod_code: "ew52"
  delayperiod_code: "dw4"
  ag_nweeks_code: "nweeks8"
  incl_trend_alerts: true
  impact_thresh: 300 
  use_filtered_linked: true 
  which_setting: "epidemic"
  new_dpath: "../../main_ecl_dec2024/outputs"
  plot_path: paste0(new_dpath, "/exploratory")
  pre_path: "../../main_ecl_dec2024/inputs/" 
  fig_path: "manuscript_figures"
  use_iqr: false
  ocv_status: "all"
execute:
  warning: false
message: false
cache: false
---

```{r}
#| label: pkg-reinstall
#| echo: false

# detach("package:OutbreakExtractR", unload=TRUE)
# devtools::document("../OutbreakExtractR")
# install.packages("../OutbreakExtractR", type="source", repos=NULL)
# params <- list(testing=FALSE)
```

```{r}
#| label: setup
#| echo: false

## Load necessary libraries
library(OutbreakExtractR)
library(tidyverse)
library(scales)
library(sf)
library(broom)
library(knitr)
library(kableExtra)
library(sf)
library(giscoR)
library(colorspace)
library(ggnewscale)
library(giscoR)
library(patchwork)
library(cowplot)

## change pre_path and new_dpath to the locations where the preoutbreak time series and alert/outcome summary data are stored respectively
new_dpath <- params$new_dpath
plot_path <- params$plot_path
pre_path <- params$pre_path
fig_path <- params$fig_path

testing <- params$testing ## process the testing dataset or full evaluation of outbreaks
test_cntry <- if(testing){"COD"}
remove_censored <- params$remove_censored ## whether to remove censored observations

## set parameters for reading files
evalperiod_code <- params$evalperiod_code
delayperiod_code <- params$delayperiod_code ## delay period for base analyses
ag_nweeks_code <- params$ag_nweeks_code

## choose whether to use SD or IQR metrics
use_iqr <- params$use_iqr

## whether to run pipeline with all locations, vaccinated ones, or unvaccinated ones
ocv_status <- params$ocv_status ## choose one of: "all", "vaccinated", "unvaccinated"

```

Updated `r Sys.Date()`

`r paste("Testing is", testing, ifelse(testing, paste0("with test country ", test_cntry, "."), "so results are for the full dataset."))`

Alertgroup `r ag_nweeks_code`; Delay period `r delayperiod_code`; Evaluation period `r evalperiod_code`

```{css classDummyTable, echo=FALSE}

.dummyTable .table thead {
  visibility: collapse;
}
.dummyTable .table tbody {
  visibility: collapse;
}
.dummyTable .table {
  border-collapse: inherit;
  padding-bottom: none;
}
.dummyTable .table.table.table-sm.table-striped.small {
  margin: auto;
}
```

```{r}
#| label: funcs-reactable-template
#| echo: false

#' @description Generic function to prepare reactable
#' @param print_table dataframe ready to be printed
#' @param col_names character vector of column names
#' @returns reactable
reactable_template <- function(print_table, col_names){
  reactable::reactable(print_table,
                       columns = col_names,
                       style = list(fontFamily = "sans-serif"),
                       highlight = TRUE,
                       striped = TRUE,
                       compact = TRUE,
                       filterable = TRUE,
                       defaultPageSize = 20
  )
}
```

# Prepare Data

Remove country-level data. Present results for a single population group (50k-500k) in main analysis. Other population groupings will go in supplement. Endemic locations are defined as those with at least 3 years of time series data and at least 50% of the weeks in the surveillance time series had non-zero suspected cholera cases.

```{r}
#| label: load-data
#| echo: false

## Set Parameters
incl_trend_alerts <- params$incl_trend_alerts
impact_thresh <- params$impact_thresh ## outcome threshold for ppv, missed outbreaks, delay
use_filtered_linked <- params$use_filtered_linked ## TRUE for dataset that CA cleaned
which_setting <- params$which_setting ## endemic, epidemic, all

## Read Data
alerts <- readr::read_rds(paste0(new_dpath, "/alerts.rds"))
ags <- readr::read_rds(paste0(new_dpath, "/alert_groups_", ag_nweeks_code, ".rds"))
evalC <- readr::read_rds(paste0(new_dpath, "/evaluation_cases_", delayperiod_code, "_", evalperiod_code, ".rds"))

if(use_filtered_linked){
  linked <- readr::read_rds(paste0(new_dpath, "/filtered_linked_outbreaks_alert_groups_", ag_nweeks_code, ".rds"))
} else{
  linked <- readr::read_rds(paste0(new_dpath, "/linked_outbreaks_alert_groups_", ag_nweeks_code, ".rds"))
}

pre_extract <- readr::read_rds(file.path(pre_path, "time_series_preoutbreak_extraction.rds"))
pre_extract <- dplyr::mutate(pre_extract, pop_brk = cut(pop, breaks = c(0, 50000, 500000, max(pre_extract$pop)), labels = c("<50k", "[50k, 500k)", "\U2265 500k"), include.lowest=T, right=FALSE))

## optionally remove censored alerts for chosen delay value
if (remove_censored) {

  observed_weeks <- readRDS("../../main_ecl_dec2024/outputs/evaluation_time_censoring.rds")

  ## keep only the delay corresponding to delayperiod_code
  chosen_delay <- as.numeric(gsub("dw", "", delayperiod_code))

  censored_alerts <- observed_weeks %>%
    dplyr::filter(delay == chosen_delay, O_weeks < 52) %>%
    dplyr::pull(alert_id) %>%
    unique()

  alerts  <- alerts  %>% dplyr::filter(!alert_id %in% censored_alerts)
  ags     <- ags     %>% dplyr::filter(!alert_id %in% censored_alerts)
  evalC   <- evalC   %>% dplyr::filter(!alert_id %in% censored_alerts)
  linked  <- linked  %>% dplyr::filter(!alert_id %in% censored_alerts)
}

## % of locations with at least one alert for manuscript text

pre_subnat <- pre_extract %>%
  dplyr::filter(stringr::str_count(location, "::") > 1)

total_locations <- pre_subnat %>%
  dplyr::distinct(location) %>%
  nrow()

alert_locations <- alerts %>%
  dplyr::filter(stringr::str_count(location, "::") > 1) %>%
  dplyr::distinct(location)

n_with_alert <- pre_subnat %>%
  dplyr::distinct(location) %>%
  dplyr::semi_join(alert_locations, by = "location") %>%
  nrow()

pct_with_alert <- round(100 * n_with_alert / total_locations, 1)
alert_text <- glue::glue(
  "{pct_with_alert}% (n = {n_with_alert}) of subnational locations ",
  "reported at least one alert of any definition."
)
alert_text


number_of_countries <- length(unique(pre_subnat$country))
country_text <- glue::glue(
  "(n = {number_of_countries}) number of countries"
)

country_text

```

The following locations are include in this report: `r which_setting`

```{r}
#| label: clean-base-data
#| echo: false

alerts_cl <- OutbreakExtractR::add_pop3_adm_columns(alerts, pre_extract) %>%
  OutbreakExtractR::add_country_column() %>%
  OutbreakExtractR::filter_ms_data(which_setting = which_setting)
ags_cl <- OutbreakExtractR::add_pop3_adm_columns(ags, pre_extract) %>%
  OutbreakExtractR::add_country_column() %>%
  OutbreakExtractR::filter_ms_data(which_setting = which_setting)
evalC_cl <- OutbreakExtractR::add_pop3_adm_columns(evalC, pre_extract) %>%
  OutbreakExtractR::add_alert_columns(alerts) %>%
  OutbreakExtractR::filter_ms_data(which_setting = which_setting)

```
Filtering alert locations according to ocv status selection

```{r}
#| label: filter-locations-by-ocv-status
#| eval: !expr params$ocv_status %in% c("vaccinated", "unvaccinated")
ocv <- read_csv(paste0(pre_path, "/gtfcc_request_timeline_2026-04-02.csv")) %>%
  filter(`Event Type` == "Round") %>%
  mutate(
    ISO3 = if_else(Country == "DR Congo", "COD", ISO3),
    date_round = Date
  ) %>%
  select(ISO3, date_round)

alerts_window <- ags_cl %>%
  select(alert_id, country, alert_number, TL_first_alert) %>%
  distinct() %>%
  left_join(
    evalC_cl %>%
      select(alert_id, evaluation_end_date),
    by = "alert_id"
  ) %>%
  mutate(
    start_window = TL_first_alert - weeks(52),
    end_window = evaluation_end_date
  )

## link OCV campaigns to alerts
alerts_with_ocv <- alerts_window %>%
  left_join(ocv, by = c("country" = "ISO3")) %>%
  mutate(
    in_window = date_round >= start_window & date_round <= end_window
  ) %>%
  group_by(alert_id) %>%
  summarise(
    vaccinated = any(in_window, na.rm = TRUE),
    .groups = "drop"
  )

evalC_cl <- evalC_cl %>%
  left_join(alerts_with_ocv, by = "alert_id")

evalC_full <- evalC_cl

eval_vaccinated <- evalC_cl %>%
  filter(vaccinated == TRUE)

eval_nonvaccinated <- evalC_cl %>%
  filter(vaccinated == FALSE | is.na(vaccinated))

table(evalC_cl$vaccinated, useNA = "ifany")

## proportion vaccinated
mean(evalC_cl$vaccinated, na.rm = TRUE)

## run the rest of the utility pipeline
if (ocv_status == "unvaccinated"){
  evalC_cl <- eval_nonvaccinated %>%
    select(-vaccinated)

  linked <- linked %>%
    dplyr::filter(!alert_id %in% eval_nonvaccinated$alert_id)

} else if (ocv_status == "vaccinated"){
  evalC_cl <- eval_vaccinated %>%
    select(-vaccinated)

  linked <- linked %>%
    dplyr::filter(!alert_id %in% eval_vaccinated$alert_id)

} else {
  evalC_cl <- evalC_cl %>%
    select(-vaccinated)
}

```

```{r, eval: ocv_status %in% c("vaccinated", "unvaccinated")}
#| echo: false
#| message: false
#| warning: false

summary_stats <- evalC_full %>%   
  summarise(
    n_alerts = n(),
    n_vaccinated = sum(vaccinated, na.rm = TRUE),
    pct_vaccinated = 100 * mean(vaccinated, na.rm = TRUE)
  )

n_alerts <- summary_stats$n_alerts
pct_vaccinated <- summary_stats$pct_vaccinated

cat(
  paste0(
    "Out of ", n_alerts, " alerts, ",
    round(pct_vaccinated, 1),
    "% occurred during periods overlapping with OCV campaign rounds."
  )
)

summary_stats
```


```{r}
#| label: location-map
#| echo: true
#| message: false
#| warning: false
#| eval: true

outbreak_shapes <- readr::read_rds(
  file.path(pre_path, "outbreak_shapefiles.rds")
)

if (!inherits(outbreak_shapes, "sf")) {
  outbreak_shapes <- sf::st_as_sf(outbreak_shapes)
}

if (is.na(sf::st_crs(outbreak_shapes))) {
  sf::st_crs(outbreak_shapes) <- 4326
} else if (sf::st_crs(outbreak_shapes) != sf::st_crs(4326)) {
  outbreak_shapes <- sf::st_transform(outbreak_shapes, 4326)
}

africa <- giscoR::gisco_get_countries(
  region = "Africa",
  resolution = "3"
)

shape_join <- outbreak_shapes %>%
  left_join(
    pre_extract %>%
      select(location_period_id, location),
    by = c("lctn_pr" = "location_period_id")
  )
if (which_setting == "endemic") {

  # Load endemic dataset from package
  data("endemic_locs_gte50nz_3ydata", package = "OutbreakExtractR")

  # Alert-linked locations
  alert_locations <- alerts_cl %>%
    dplyr::filter(alert_id %in% evalC_cl$alert_id) %>%
    dplyr::distinct(location) %>%
    dplyr::pull(location)

  # Keep only endemic locations that also have alerts
  endemic_alert_locs <- endemic_locs_gte50nz_3ydata %>%
    dplyr::filter(location %in% alert_locations) %>%
    dplyr::distinct(location)

  endemic_sf <- endemic_alert_locs %>%
    dplyr::left_join(shape_join, by = "location") %>%
    dplyr::distinct(location, .keep_all = TRUE) %>%
    sf::st_as_sf() %>%
    sf::st_make_valid()

  # Derive spatial scale from number of "::" separators in location
  endemic_sf <- endemic_sf %>%
    dplyr::mutate(
      n_levels = stringr::str_count(location, "::") + 1,
      scale_group = dplyr::case_when(
        n_levels == 1 ~ "Admin0",
        n_levels == 2 ~ "Admin1",
        n_levels == 3 ~ "Admin2",
        n_levels >= 4 ~ "Admin3",
        TRUE ~ NA_character_
      ),
      scale_group = factor(
        scale_group,
        levels = c("Admin0", "Admin1", "Admin2", "Admin3")
      )
    )

  scale_colors <- c(
    "Admin0" = "lightblue0",
    "Admin1" = "lightblue1",
    "Admin2" = "lightblue3",
    "Admin3" = "lightblue4"
  )

  ggplot() +
    geom_sf(
      data = africa,
      fill = "gray95",
      color = "gray80",
      linewidth = 0.2
    ) +
    geom_sf(
      data = endemic_sf %>% dplyr::filter(scale_group == "Admin0"),
      aes(fill = scale_group),
      color = "white",
      linewidth = 0.2
    ) +
    geom_sf(
      data = endemic_sf %>% dplyr::filter(scale_group == "Admin1"),
      aes(fill = scale_group),
      color = "white",
      linewidth = 0.2
    ) +
    geom_sf(
      data = endemic_sf %>% dplyr::filter(scale_group == "Admin2"),
      aes(fill = scale_group),
      color = "white",
      linewidth = 0.2
    ) +
    geom_sf(
      data = endemic_sf %>% dplyr::filter(scale_group == "Admin3"),
      aes(fill = scale_group),
      color = "white",
      linewidth = 0.2,
      alpha = 0.9
    ) +
    scale_fill_manual(values = scale_colors, name = "Spatial Scale") +
    theme_minimal() +
    labs(
      title = "Alert locations, endemic"
    ) +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      legend.position = "bottom",
      axis.title = element_blank()
    ) +
    coord_sf(
      xlim = c(-20, 55),
      ylim = c(-35, 38),
      expand = FALSE
    )

} else {

  alert_locations <- alerts_cl %>%
    filter(alert_id %in% evalC_cl$alert_id) %>%
    select(location, spatial_scale) %>%
    distinct(location, .keep_all = TRUE)

  alerts_sf <- alert_locations %>%
    left_join(shape_join, by = "location") %>%
    distinct(location, .keep_all = TRUE) %>%
    sf::st_as_sf() %>%
    sf::st_make_valid()

  alerts_sf <- alerts_sf %>%
    mutate(
      scale_group = case_when(
        spatial_scale %in% c("admin1", "admin1 composite") ~ "Admin1",
        spatial_scale %in% c("admin2", "admin2 composite") ~ "Admin2",
        spatial_scale %in% c("admin3", "admin3 composite") ~ "Admin3",
        TRUE ~ spatial_scale
      ),
      scale_group = factor(
        scale_group,
        levels = c("Admin1", "Admin2", "Admin3")
      )
    )

  scale_colors <- c(
    "Admin1" = "lightblue1",
    "Admin2" = "lightblue3",
    "Admin3" = "lightblue4"
  )

  ggplot() +
    geom_sf(
      data = africa,
      fill = "gray95",
      color = "gray80",
      linewidth = 0.2
    ) +
    geom_sf(
      data = alerts_sf %>% filter(scale_group == "Admin1"),
      aes(fill = scale_group),
      color = "white",
      linewidth = 0.2
    ) +
    geom_sf(
      data = alerts_sf %>% filter(scale_group == "Admin2"),
      aes(fill = scale_group),
      color = "white",
      linewidth = 0.2
    ) +
    geom_sf(
      data = alerts_sf %>% filter(scale_group == "Admin3"),
      aes(fill = scale_group),
      color = "white",
      linewidth = 0.2,
      alpha = 0.9
    ) +
    scale_fill_manual(
      values = scale_colors,
      name = "Spatial Scale"
    ) +
    theme_minimal() +
    labs(
      title = paste0(
        "Alert locations, ",
        which_setting,
        " ",
        ocv_status
      )
    ) +
    theme(
      plot.title = element_text(size = 16, face = "bold"),
      legend.position = "bottom",
      legend.title = element_text(size = 11, face = "bold"),
      legend.text = element_text(size = 10),
      axis.title = element_blank()
    ) +
    coord_sf(
      xlim = c(-20, 55),
      ylim = c(-35, 38),
      expand = FALSE
    )

}


```


```{r}
#| label: prepare-linked-data
#| echo: true

## subset to outbreaks exceeding a threshold
print(paste(impact_thresh, "suspected cases is the outbreak filtering threshold"))

if(use_filtered_linked){

  linked_cl <- OutbreakExtractR::add_pop3_adm_columns_outbreak(linked, pre_extract) %>% ## N.B. function changes the outbreak_UID separators to underscores
    OutbreakExtractR::add_country_column() %>%
    OutbreakExtractR::add_alert_columns_outbreak(alerts) %>%
    OutbreakExtractR::filter_ms_data(which_setting = which_setting)

  ## checks
  print("Number of outbreak_UIDs missing at least one of 24 alert definitions (linked)")
  linked %>% group_by(outbreak_UID) %>% count() %>% filter(n<24) %>% nrow
  print("Number of outbreak_UIDs missing at least one of 24 alert definitions")
  linked_cl %>% group_by(outbreak_UID) %>% count() %>% filter(n<24) %>% nrow

} else{
  message("Unfiltered links to outbreak data were used. Filtering to one alert definition per outbreak via automated selection.")

  linked_cl_tmp <- OutbreakExtractR::add_pop3_adm_columns_outbreak(linked, pre_extract) %>%
    OutbreakExtractR::add_country_column() %>%
    OutbreakExtractR::add_alert_columns_outbreak(alerts) %>%
    OutbreakExtractR::filter_ms_data(which_setting = which_setting)

  linked_cl_1row <- dplyr::group_by(linked_cl_tmp, outbreak_UID, alert_number) %>%
    dplyr::count() %>%
    dplyr::filter(n <= 1) %>%
    dplyr::ungroup()
  linked_cl_mrow <- dplyr::group_by(linked_cl_tmp, outbreak_UID, alert_number) %>%
    dplyr::count() %>%
    dplyr::filter(n > 1) %>%
    dplyr::ungroup()

  ## clean outbreak-alert linkages with multiple linkage options
  clean_mrow_linkages <- function(linked_mrow_dat, linked_full_dat, i){
    id <- linked_mrow_dat[i,]$outbreak_UID
    alert_num <- linked_mrow_dat[i,]$alert_number

    tmp <- dplyr::filter(linked_full_dat, outbreak_UID == id & alert_number == alert_num)

    if ("overlap" %in% tmp$linkage){
      tmp2 <- dplyr::mutate(tmp, keep = dplyr::if_else(linkage == "overlap", TRUE, FALSE))
    } else if ("closest before" %in% tmp$linkage){
      tmp2 <- dplyr::mutate(tmp, keep = dplyr::if_else(linkage == "closest before", TRUE, FALSE))
    }

    rc <- dplyr::filter(tmp2, keep) %>% dplyr::select(-keep)
    if(nrow(rc) != 1){
      stop("More than 1 row was returned")
    }
    return(rc)
  }

  linked_mrow_singles <- 1:nrow(linked_cl_mrow) |>
    purrr::map(\(x) clean_mrow_linkages(linked_cl_mrow, linked_cl_tmp, i = x)) %>%
    purrr::list_rbind() %>%
    dplyr::ungroup()

  linked_1rows <- dplyr::left_join(dplyr::select(linked_cl_1row, outbreak_UID, alert_number), linked_cl_tmp, by = c("outbreak_UID", "alert_number"))

  ## checks
  print("Number of outbreak UIDs in pre and post cleaning are the same")
  nrow(linked_mrow_singles) + nrow(linked_1rows) == nrow(dplyr::distinct(linked_cl_tmp, outbreak_UID, alert_number))

  linked_cl <- dplyr::bind_rows(linked_1rows, linked_mrow_singles) ## one linkage per alert def - outbreak
}

linked_cl_large <- dplyr::filter(linked_cl, total_sCh > impact_thresh) ## only keep outbreaks above impact threshold

```

Identify proportion of target population as \>1 year olds by country according to UNWPP2024 in the alert year.

```{r}
#| label: prep_target_pop
#| echo: false

library(wpp2024)
library(countrycode)
data(popAge1dt)

iso3_uq <- sort(unique(evalC_cl$country))
codes <- dplyr::select(countrycode::codelist, iso3c, un.name.en) %>%
  dplyr::filter(iso3c %in% iso3_uq) %>%
  dplyr::mutate(un.name.en = dplyr::case_when(iso3c == "CIV" ~ "Cote d'Ivoire",
                              .default = un.name.en))

iso3_targetpop <- popAge1dt %>%
  dplyr::mutate(target_pop = dplyr::if_else(age != 0, pop, 0)) %>%
  dplyr::group_by(name, year) %>%
  dplyr::summarise(target_pop = sum(target_pop), total_pop = sum(pop)) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(prop_targeted = round(target_pop/total_pop, 2)) %>%
  dplyr::right_join(codes, by = c("name" = "un.name.en")) %>%
  dplyr::relocate(iso3c, .after = "name") %>%
  dplyr::rename(country = iso3c)

targetpop_cl <- dplyr::select(iso3_targetpop, country, year, prop_targeted)

merge_prop_targeted <- dplyr::distinct(ags_cl, alert_id, TL_first_alert, country) %>%
  dplyr::mutate(TL_year = lubridate::year(TL_first_alert)) %>%
  dplyr::left_join(targetpop_cl, by = c("country", "TL_year" = "year")) %>%
  dplyr::select(alert_id, prop_targeted)

```

# Alert utility

1.  Alerts preceding highest potential impact
2.  Alerts preceding highest potential efficiency
3.  Alerts with high PPV (outcomes with \>`r impact_thresh` cases)
4.  Proportion of outbreaks with \>`r impact_thresh` cases that were missed
5.  Delay from outbreak start for outbreaks \>`r impact_thresh` cases

```{r}
#| label: utility_functions
#| echo: false

quibble2 <- function(colname, q = c(0.025, 0.25, 0.5, 0.75, 0.975)) {
  tibble("{{ colname }}_q" := paste0("q", stringr::str_remove(as.character(q), "0.")), "{{ colname }}" := quantile(colname, q, na.rm = TRUE))
}

prepare_fig_dat <- function(df){
  OutbreakExtractR::label_alerts_short(df, is_ordered = TRUE) %>%
  dplyr::mutate(alert_lab = forcats::fct_rev(alert_lab)) %>%
  OutbreakExtractR::label_alerttypes()
}

```

## Data basics

Number of alerts per alert definition, number of outbreaks linked to each alert definition

```{r}
#| label: tab-data-basics
#| echo: false

nalerts <- dplyr::group_by(ags_cl, pop_brk, alert_type, alert_number) %>%
  dplyr::count() %>%
  dplyr::rename(n_alerts = n)

noutbreaks <- dplyr::filter(linked_cl_large, is_triggered) %>% 
  dplyr::group_by(pop_brk, alert_type, alert_number) %>%
  dplyr::count() %>%
  dplyr::rename(n_outbreaks = n)

df_basics <- dplyr::full_join(nalerts, noutbreaks, by = c("pop_brk", "alert_type", "alert_number")) %>%
  prepare_fig_dat() %>%
  dplyr::select(alert_type, pop_brk, alert_lab, n_alerts, n_outbreaks)

reactable_template(df_basics,
                  col_names = list(alert_type = reactable::colDef(name = "Alert Type"),
                                  pop_brk = reactable::colDef(name = "Pop Group"),
                                  alert_lab = reactable::colDef(name = "Alert Def"),
                                  n_alerts = reactable::colDef(name = "N Alerts"), 
                                  n_outbreaks = reactable::colDef(name = "N Outbreaks")))

```

## Dimensions of alert utility

```{r}
#| label: prep_underlying_data
#| echo: false

### impact and efficiency ###
prep_impact_eff_dat <- dplyr::left_join(evalC_cl, merge_prop_targeted, by = c("alert_id")) %>%
  dplyr::rename(impact = total_sCh) %>%
  dplyr::mutate(eff = impact / (prop_targeted*pop) * 1000)

if(use_iqr){
  means_impact_eff_dat <- prep_impact_eff_dat %>%
  dplyr::group_by(alert_number, alert_type, pop_brk) %>%
  dplyr::summarise(impact_mean = median(impact), eff_mean = median(eff)) %>%
  dplyr::ungroup()
} else {
  means_impact_eff_dat <- prep_impact_eff_dat %>%
  dplyr::group_by(alert_number, alert_type, pop_brk) %>%
  dplyr::summarise(impact_mean = mean(impact), eff_mean = mean(eff)) %>%
  dplyr::ungroup()
}

### ppv ###
prep_ppv_dat <- dplyr::mutate(evalC_cl, true_positives := dplyr::if_else(total_sCh > impact_thresh, TRUE, FALSE))

### missed outbreaks and delayed detection ###
prep_missed_delay_dat <- linked_cl_large %>%
  dplyr::mutate(delay = as.numeric(alert_group_start-outbreak_start, units = "weeks"))

if (use_iqr){
 means_delay_dat <- prep_missed_delay_dat %>%
  dplyr::group_by(alert_number, alert_type, pop_brk) %>%
  dplyr::summarise(delay_mean = median(delay, na.rm = TRUE)) %>%
  dplyr::ungroup()   
} else {
 means_delay_dat <- prep_missed_delay_dat %>%
  dplyr::group_by(alert_number, alert_type, pop_brk) %>%
  dplyr::summarise(delay_mean = mean(delay, na.rm = TRUE)) %>%
  dplyr::ungroup() 
}

n_outbreaks <- linked_cl_large %>%
  dplyr::distinct(outbreak_UID, pop_brk) %>%
  dplyr::group_by(pop_brk) %>%
  dplyr::summarise(n_outbreaks = n())

sum(n_outbreaks$n_outbreaks) == length(unique(linked_cl_large$outbreak_UID)) ## total number of outbreaks

```

```{r}
#| label: prep_fig_data
#| echo: false

fig_impact_dat <- dplyr::group_by(prep_impact_eff_dat, alert_number, alert_type, pop_brk) %>%
  dplyr::reframe(quibble2(impact)) %>%
  tidyr::pivot_wider(names_from = impact_q, names_glue = "{.value}_{impact_q}", values_from = impact) %>%
  dplyr::full_join(dplyr::select(means_impact_eff_dat, alert_number, alert_type, pop_brk, impact_mean), by = c("alert_number", "alert_type", "pop_brk")) %>%
  prepare_fig_dat()

fig_eff_dat <- dplyr::group_by(prep_impact_eff_dat, alert_number, alert_type, pop_brk) %>%
  dplyr::reframe(quibble2(eff)) %>%
  tidyr::pivot_wider(names_from = eff_q, names_glue = "{.value}_{eff_q}", values_from = eff) %>%
  dplyr::full_join(dplyr::select(means_impact_eff_dat, alert_number, alert_type, pop_brk, eff_mean), by = c("alert_number", "alert_type", "pop_brk")) %>%
  prepare_fig_dat()

fig_ppv_dat <- dplyr::group_by(prep_ppv_dat, alert_number, alert_type, pop_brk) %>%
  dplyr::summarise(alerts = n(), true_positives = sum(true_positives), ppv = round(true_positives/alerts, 2)) %>%
  prepare_fig_dat()

fig_missed_dat <- dplyr::group_by(prep_missed_delay_dat, alert_number, alert_type, pop_brk) %>%
  dplyr::summarise(n_detected = sum(is_triggered)) %>%
  dplyr::left_join(n_outbreaks, by = c("pop_brk")) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(n_missed = n_outbreaks - n_detected,
                missed_prop = n_missed/n_outbreaks) %>%
  prepare_fig_dat()

fig_delay_dat <- dplyr::group_by(prep_missed_delay_dat, alert_number, alert_type, pop_brk) %>%
  dplyr::reframe(quibble2(delay)) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_wider(names_from = delay_q, names_glue = "{.value}_{delay_q}", values_from = delay) %>%
  dplyr::full_join(means_delay_dat, by = c("alert_number", "alert_type", "pop_brk")) %>%
  prepare_fig_dat()

fig_complete_dat <- dplyr::full_join(fig_impact_dat, fig_eff_dat, by = c("alert_number", "alert_type", "pop_brk", "alert_lab")) %>%
  dplyr::full_join(fig_ppv_dat, by = c("alert_number", "alert_type", "pop_brk", "alert_lab")) %>%
  dplyr::full_join(fig_missed_dat, by = c("alert_number", "alert_type", "pop_brk", "alert_lab")) %>%
  dplyr::full_join(fig_delay_dat, by = c("alert_number", "alert_type", "pop_brk", "alert_lab"))

## 30 Oct save as file to compare epidemic/endemic locations and test for statistically significant differences
add_setting_and_save <- function(df, name) {
  df$which_setting <- which_setting
  saveRDS(df, paste0(pre_path, name, which_setting, ".rds"))
}

add_setting_and_save(fig_complete_dat, "compare_significant_missed_ppv_chisquare_")

fig_under_50k <- dplyr::filter(fig_complete_dat, pop_brk == "< 50k")
fig_50k_500k <- dplyr::filter(fig_complete_dat, pop_brk == "[50k, 500k)")
fig_over_500k <- dplyr::filter(fig_complete_dat, pop_brk == "\U2265 500k")

```

```{r}
#| label: func-figs
#| echo: false

impact_lineplot <- function(df, title_lab = "Impact", col_pal = "Dark2"){
  df_mean <- dplyr::summarise(df, impact_total_mean = mean(impact_mean, na.rm = TRUE))
  print(paste("Impact mean:", df_mean$impact_total_mean))

  ggplot2::ggplot(df, aes(y = alert_lab)) +
    geom_point(aes(x = impact_mean, colour = alert_type)) +
    geom_point(aes(x = impact_q5, colour = alert_type), shape = "x", size = 4) +
    geom_linerange(aes(xmin = impact_q25, xmax = impact_q75, colour = alert_type)) +
    geom_vline(data = df_mean, aes(xintercept = impact_total_mean), colour = "black", linetype = 2, alpha = 0.7) +
    scale_colour_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Suspected Cases", labels = scales::label_number(scale = 0.001, suffix = "k")) +
    theme_bw() +
    theme(legend.position = "none", axis.title.y = element_blank(), plot.margin = unit(c(1,1,1,1), "mm")) +
    ggtitle(title_lab)
}

eff_lineplot <- function(df, title_lab = "Efficiency", col_pal = "Dark2"){
  df_mean <- dplyr::summarise(df, eff_total_mean = mean(eff_mean, na.rm = TRUE))
  print(paste("Efficiency mean:", df_mean$eff_total_mean))

  ggplot2::ggplot(df, aes(y = alert_lab)) +
    geom_point(aes(x = eff_mean, colour = alert_type)) +
    geom_point(aes(x = eff_q5, colour = alert_type), shape = "x", size = 4) +
    geom_linerange(aes(xmin = eff_q25, xmax = eff_q75, colour = alert_type)) +
    geom_vline(data = df_mean, aes(xintercept = eff_total_mean), colour = "black", linetype = 2, alpha = 0.7) +
    scale_colour_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Cases/1k target pop", labels = scales::label_number(big.mark = "", accuracy = 1), limits = c(0, max(df$eff_mean, na.rm = TRUE)*1.1)) +
    theme_bw() +
    theme(legend.position = "none", axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank(), plot.margin = unit(c(1,1,1,1), "mm")) +
    ggtitle(title_lab)
}

ppv_barplot <- function(df, title_lab = "PPV", col_pal = "Dark2"){
  df_mean <- dplyr::summarise(df, ppv_mean = mean(ppv, na.rm = TRUE))
  print(paste("PPV mean:", df_mean$ppv_mean))

  ggplot2::ggplot(df, aes(x = ppv, y = alert_lab)) +
    geom_col(aes(fill = alert_type)) +
    geom_vline(data = df_mean, aes(xintercept = ppv_mean), colour = "black", linetype = 2, alpha = 0.7) +
    scale_x_continuous("PPV", labels = scales::label_number(accuracy = 0.1), breaks = seq(0, 1, by = 0.2), limits = c(0, 1)) +
    scale_y_discrete("Alert Type") +
    scale_fill_brewer("Alert Type", palette = col_pal) +
    theme_bw() +
    theme(legend.position = "none", axis.title.y = element_blank(), axis.ticks.y = element_blank(), axis.text.y = element_blank(), plot.margin = unit(c(1,1,1,3), "mm")) +
    ggtitle(title_lab)
}

missed_barplot <- function(df, title_lab = "Missed", col_pal = "Dark2"){
  df_mean <- dplyr::summarise(df, missed_mean = mean(missed_prop, na.rm = TRUE))
  print(paste("Missed mean:", df_mean$missed_mean))

  ggplot2::ggplot(df, aes(x = missed_prop, y = alert_lab)) +
    geom_col(aes(fill = alert_type)) +
    geom_vline(data = df_mean, aes(xintercept = missed_mean), colour = "black", linetype = 2, alpha = 0.7) +
    scale_fill_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Prop of outbreaks", labels = scales::label_number(accuracy = 0.1), breaks = seq(0, 1, by = 0.2), limits = c(0, 1)) +
    theme_bw() +
    theme(legend.position = "none", axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), plot.margin = unit(c(1,1,1,1), "mm")) +
    ggtitle(title_lab)
}

delay_lineplot <- function(df, title_lab = "Delay", col_pal = "Dark2"){
  df_mean <- dplyr::summarise(df, delay_total_mean = mean(delay_mean, na.rm = TRUE))
  print(paste("Delay mean:", df_mean$delay_total_mean))

  ggplot2::ggplot(df, aes(y = alert_lab)) +
    geom_point(aes(x = delay_mean, colour = alert_type)) +
    geom_point(aes(x = delay_q5, colour = alert_type), shape = "x", size = 4) +
    geom_vline(data = df_mean, aes(xintercept = delay_total_mean), colour = "black", linetype = 2, alpha = 0.7) +
    geom_linerange(aes(xmin = delay_q25, xmax = delay_q75, colour = alert_type)) +
    scale_colour_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Weeks of delay", labels = scales::label_number()) +
    theme_bw() +
    theme(legend.position = "right", axis.title.y = element_blank(), axis.text.y = element_blank(), axis.ticks.y = element_blank(), plot.margin = unit(c(1,1,1,1), "mm")) +
    ggtitle(title_lab)
}
```



```{r}
#| label: boxplots-and-barcharts-mainpop
#| echo: true
#| fig.width: 10

col_pal <- "Dark2"

fig_impact_raw <- prep_impact_eff_dat %>% prepare_fig_dat()
fig_eff_raw    <- prep_impact_eff_dat %>% prepare_fig_dat()
fig_delay_raw  <- prep_missed_delay_dat %>% prepare_fig_dat()

## 30 Oct save as file to compare epidemic/endemic locations and test for statistically significant differences
add_setting_and_save(fig_impact_raw, "compare_significant_impact_testmeans_")
add_setting_and_save(fig_eff_raw, "compare_significant_efficiency_testmeans_")
add_setting_and_save(fig_delay_raw, "compare_significant_delay_testmeans_")

pop_groups <- c("< 50k", "[50k, 500k)", "\U2265 500k")

impact_data_list <- lapply(pop_groups, function(pg) fig_impact_raw %>% filter(pop_brk == pg))
names(impact_data_list) <- pop_groups

eff_data_list <- lapply(pop_groups, function(pg) fig_eff_raw %>% filter(pop_brk == pg))
names(eff_data_list) <- pop_groups

delay_data_list <- lapply(pop_groups, function(pg) fig_delay_raw %>% filter(pop_brk == pg))
names(delay_data_list) <- pop_groups

## function to add vertical dashed line for overall mean
add_mean_line <- function(df, xvar) {
  overall_mean <- mean(df[[xvar]], na.rm = TRUE)
  ymin <- 0.5
  ymax <- length(unique(df$alert_lab)) + 0.5
  geom_segment(aes(x = overall_mean, xend = overall_mean, y = ymin, yend = ymax),
               linetype = 2, color = "black", alpha = 0.7, inherit.aes = FALSE)
}

center_x_on_mean <- function(df, xvar) {
  x_mean <- mean(df[[xvar]], na.rm = TRUE)
  x_lower <- quantile(df[[xvar]], 0.05, na.rm = TRUE)
  x_upper <- quantile(df[[xvar]], 0.95, na.rm = TRUE)
  xlim_min <- min(x_lower, x_mean - 0.2 * (x_upper - x_lower))
  xlim_max <- max(x_upper, x_mean + 0.2 * (x_upper - x_lower))
  
  c(xlim_min, xlim_max)
}

center_x_on_box <- function(df, xvar, groupvar) {
  lims <- df %>%
    group_by(.data[[groupvar]]) %>%
    summarise(
      q1 = quantile(.data[[xvar]], 0.25, na.rm = TRUE),
      q3 = quantile(.data[[xvar]], 0.75, na.rm = TRUE),
      iqr = q3 - q1,
      lower = q1 - 1.5 * iqr,
      upper = q3 + 1.5 * iqr
    )
  
  c(min(lims$lower), max(lims$upper))
}

## impact boxplot
create_impact_boxplot <- function(df, pop_label) {
  xlim_vals <- center_x_on_mean(df, "impact")
  ggplot(df, aes(x = impact, y = alert_lab, fill = alert_type)) +
    geom_boxplot(outlier.shape = 21, outlier.fill = NA) +
    stat_summary(fun = mean, geom = "point", size = 2.5, fill = "white") +
    add_mean_line(df, "impact") +
    scale_fill_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Suspected Cases", labels = label_number(scale = 0.001, suffix = "k")) +
    coord_cartesian(xlim = xlim_vals) +
    theme_bw() +
    theme(
      legend.position = "none",
      axis.title.y = element_blank(),
      axis.text.y = element_text(size = 9, color = "black"),
      axis.ticks.y = element_line(),
      plot.margin = unit(c(1,1,1,1), "mm")
    ) +
    ggtitle("Impact")
}

## efficiency boxplot
create_eff_boxplot <- function(df, xvar, metric_label, pop_label) {
  xlim_vals <- center_x_on_mean(df, xvar)
  ggplot(df, aes_string(x = xvar, y = "alert_lab", fill = "alert_type")) +
    geom_boxplot(outlier.shape = 21, outlier.fill = NA) +
    stat_summary(fun = mean, geom = "point", size = 2.5, fill = "white") +
    add_mean_line(df, xvar) +
    scale_fill_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Cases/1k target pop", labels = label_number(big.mark = "", accuracy = 1)) +
    coord_cartesian(xlim = xlim_vals) +
    theme_bw() +
    theme(
      legend.position = "none",
      axis.title.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.text.y = element_blank(),
      plot.margin = unit(c(1,1,1,1), "mm")
    ) +
    ggtitle(metric_label)
}

## delay boxplot (legend visible)
create_delay_boxplot <- function(df, xvar, metric_label, pop_label) {
  xlim_vals <- center_x_on_mean(df, xvar)
  ggplot(df, aes_string(x = xvar, y = "alert_lab", fill = "alert_type")) +
    geom_boxplot(outlier.shape = 21, outlier.fill = NA) +
    stat_summary(fun = mean, geom = "point", size = 2.5, fill = "white") +
    add_mean_line(df, xvar) +
    scale_fill_brewer("Alert Type", palette = col_pal) +
    scale_x_continuous("Weeks of delay", labels = label_number(big.mark = "", accuracy = 1)) +
    coord_cartesian(xlim = c(xlim_vals[1], 12)) +
    theme_bw() +
    theme(
      legend.position = "none",
      axis.title.y = element_blank(),
      axis.ticks.y = element_blank(),
      axis.text.y = element_blank(),
      plot.margin = unit(c(1,1,1,1), "mm")
    ) +
    ggtitle(metric_label)
}

## boxplots per population group
impact_boxplots <- lapply(impact_data_list, create_impact_boxplot)
eff_boxplots    <- lapply(eff_data_list, function(df) create_eff_boxplot(df, "eff", "Efficiency", NULL))
delay_boxplots  <- lapply(delay_data_list, function(df) create_delay_boxplot(df, "delay", "Timeliness", NULL))

## Mar 2026 saving as png
ws <- c(4, 4, 2, 2, 4)

p <- gridExtra::arrangeGrob(
  impact_boxplots[["[50k, 500k)"]],
  eff_boxplots[["[50k, 500k)"]],
  ppv_barplot(fig_50k_500k),
  missed_barplot(fig_50k_500k),
  delay_boxplots[["[50k, 500k)"]],
  nrow = 1, widths = ws, padding = unit(0.6, "line")
)

## save
ggplot2::ggsave(paste0(fig_path,"/fig2_boxplot","_",ocv_status ,".png"), p, width = 10, height = 5, dpi = 300)
## print to report
grid::grid.newpage()
grid::grid.draw(p)

## helper: summarise existing columns safely
summarise_metrics <- function(df, metrics, source) {
  present <- intersect(metrics, names(df))
  if(length(present) == 0) return(NULL)
  
  df %>%
    select(pop_brk, alert_type, any_of(present)) %>%
    pivot_longer(cols = any_of(present), names_to = "metric", values_to = "value") %>%
    group_by(pop_brk, metric) %>%
    summarise(mean = mean(value, na.rm = TRUE),
              sd   = sd(value, na.rm = TRUE),
              source = source,
              .groups = "drop")
}

summary_impact <- summarise_metrics(fig_impact_raw, metrics = c("impact_mean", "impact"), source = "fig_impact_raw")
summary_eff <- summarise_metrics(fig_eff_raw, metrics = c("eff_mean", "eff"), source = "fig_eff_raw")
summary_delay <- summarise_metrics(fig_delay_raw, metrics = c("delay_mean", "delay"), source = "fig_delay_raw")
summary_ppv_missed <- summarise_metrics(fig_50k_500k, metrics = c("ppv", "missed_prop"), source = "fig_50k_500k")

metric_summary <- bind_rows(summary_impact, summary_eff, summary_delay, summary_ppv_missed)

print(metric_summary)

metric_summary_fmt <- metric_summary %>%
  mutate(mean_sd = sprintf("%.2f (%.2f)", mean, sd)) %>%
  select(pop_brk, metric, mean_sd, source)

print(metric_summary_fmt)

```

### Dimension distributions, 50-500k

```{r}
#| label: figs-50k-500k
#| fig-cap: Utility score dimensions for 50k-500k population group. The underlying metric for each utility score dimension is displayed for all alert definitions. The vertical black dashed line depicts the mean across alert definitions and colors indicate the alert type. For line plots, 'X' indicates median of the distribution, a point indicates mean, and line indicates IQR.
#| fig-width: 9
#| fig-height: 6
#| echo: true

ws <- c(1, .7, .7, .7, 1.25)

gridExtra::grid.arrange(impact_lineplot(fig_50k_500k), eff_lineplot(fig_50k_500k), ppv_barplot(fig_50k_500k), missed_barplot(fig_50k_500k), delay_lineplot(fig_50k_500k), nrow=1, widths = ws, padding = unit(0.6, "line"))

```

### Supplement

#### Dimension distributions, \<50k

```{r}
#| label: figs-under-50k
#| fig-cap: Utility score dimensions for under 50k population group. The underlying metric for each utility score dimension is displayed for all alert definitions. The vertical black dashed line depicts the mean across alert definitions and colors indicate the alert type. For line plots, 'X' indicates median of the distribution, a point indicates mean, and line indicates IQR.
#| fig-width: 9
#| fig-height: 6
#| echo: true

gridExtra::grid.arrange(impact_lineplot(fig_under_50k), eff_lineplot(fig_under_50k), ppv_barplot(fig_under_50k), missed_barplot(fig_under_50k), delay_lineplot(fig_under_50k), nrow=1, widths = ws, padding = unit(0.6, "line"))

```

#### Dimension distributions \>500k

```{r}
#| label: figs-over-500k
#| fig-cap: Utility score dimensions for over 500k population group. The underlying metric for each utility score dimension is displayed for all alert definitions. The vertical black dashed line depicts the mean across alert definitions and colors indicate the alert type. For line plots, 'X' indicates median of the distribution, a point indicates mean, and line indicates IQR.
#| fig-width: 9
#| fig-height: 6
#| echo: true

gridExtra::grid.arrange(impact_lineplot(fig_over_500k), eff_lineplot(fig_over_500k), ppv_barplot(fig_over_500k), missed_barplot(fig_over_500k), delay_lineplot(fig_over_500k), nrow=1, widths = ws, padding = unit(0.6, "line"))

```

## Dimension comparison: endemic vs epidemic by pop group & alert definition

```{r}
#| results: asis

## Combined Endemic vs Epidemic comparisons (Wilcoxon and Chi-square/Fisher)

all_results <- list()  ## store all metrics together
settings <- c("endemic", "epidemic")
show_only_main <- TRUE
wilcox_metrics <- c("delay", "efficiency", "impact")
metric_cols <- list(
  delay = "delay",
  efficiency = "eff",
  impact = "impact"
)

for (metric in wilcox_metrics) {
  files <- setNames(
    paste0(pre_path, "compare_significant_", metric, "_testmeans_", settings, ".rds"),
    settings
  )
  if (!all(file.exists(files))) next

  df <- do.call(rbind, lapply(names(files), function(s) {
    d <- readRDS(files[[s]])
    d$which_setting <- s
    d
  }))
  df$which_setting <- factor(df$which_setting, levels = settings)
  value_col <- metric_cols[[metric]]

  ## plot histograms to explore distribution of wilcoxon metrics
  p <- ggplot(df, aes(x = .data[[value_col]], fill = which_setting)) +
    geom_histogram(alpha = 0.7, bins = 30) +
    facet_wrap(~which_setting, ncol = 1, scales = "free_y") +
    scale_fill_manual(values = c("endemic" = "#1f78b4", "epidemic" = "#e31a1c")) +
    labs(title = paste("Distribution of", metric, "(Endemic vs Epidemic)"),
         x = metric, y = "Count") +
    theme_minimal(base_size = 13) +
    theme(plot.title = element_text(face = "bold", hjust = 0.5),
          axis.title.x = element_text(face = "bold"),
          axis.title.y = element_text(face = "bold"),
          legend.position = "none")
  print(p)

  ## Wilcoxon test
  wilcox_results <- df %>%
    group_by(pop_brk, alert_lab) %>%
    group_modify(~{
      x <- .x[!is.na(.x[[value_col]]), ]
      if (length(unique(x$which_setting)) == 2) {
        fm <- reformulate("which_setting", response = value_col)
        t <- suppressWarnings(wilcox.test(fm, data = x, exact = FALSE))
        tidy(t)
      } else tibble(statistic = NA_real_, p.value = NA_real_)
    }) %>%
    ungroup() %>%
    mutate(test = "Wilcoxon rank-sum", method = "Wilcoxon")

  ## medians
  medians <- df %>%
    group_by(pop_brk, alert_lab, which_setting) %>%
    summarize(median_value = median(.data[[value_col]], na.rm = TRUE), .groups = "drop") %>%
    pivot_wider(names_from = which_setting, values_from = median_value)

  ## combine
  final_table <- wilcox_results %>%
    left_join(medians, by = c("pop_brk", "alert_lab")) %>%
    rename(endemic = endemic, epidemic = epidemic) %>%
    mutate(metric = metric) %>%
    select(metric, pop_brk, alert_lab, test, method, endemic, epidemic, statistic, p.value)

  ## filter main
  if (show_only_main) {
    final_table <- final_table %>%
      mutate(alert_lab_clean = str_replace_all(alert_lab, "\u2265", ">=")) %>%
      filter(pop_brk == "[50k, 500k)", alert_lab_clean == ">= 50 weekly") %>%
      select(-alert_lab_clean)
  }

  all_results[[metric]] <- final_table
}

## Chi-square / Fisher metrics
chisq_metrics <- c("missed", "ppv")
metric_file <- "compare_significant_missed_ppv_chisquare_"

for (metric in chisq_metrics) {
  files <- setNames(
    paste0(pre_path, metric_file, settings, ".rds"),
    settings
  )
  if (!all(file.exists(files))) next

  df <- do.call(rbind, lapply(names(files), function(s) {
    d <- readRDS(files[[s]])
    d$which_setting <- s
    d
  }))
  df$which_setting <- factor(df$which_setting, levels = settings)

  test_results <- df %>%
    group_by(pop_brk, alert_lab) %>%
    group_modify(~{
      TP_pair     <- tapply(.x$true_positives, .x$which_setting, sum)
      Alerts_pair <- tapply(.x$alerts, .x$which_setting, sum)
      FN_pair     <- tapply(.x$n_missed, .x$which_setting, sum)
      Nout_pair   <- tapply(.x$n_outbreaks, .x$which_setting, sum)

      TP_pair[is.na(TP_pair)]         <- 0
      Alerts_pair[is.na(Alerts_pair)] <- 0
      FN_pair[is.na(FN_pair)]         <- 0
      Nout_pair[is.na(Nout_pair)]     <- 0

      if (metric == "ppv") {
        prop_end <- ifelse(Alerts_pair["endemic"] > 0,
                           TP_pair["endemic"] / Alerts_pair["endemic"], NA)
        prop_epi <- ifelse(Alerts_pair["epidemic"] > 0,
                           TP_pair["epidemic"] / Alerts_pair["epidemic"], NA)
        mat <- matrix(c(TP_pair["endemic"], Alerts_pair["endemic"] - TP_pair["endemic"],
                        TP_pair["epidemic"], Alerts_pair["epidemic"] - TP_pair["epidemic"]),
                      nrow = 2, byrow = TRUE)
      } else {
        prop_end <- ifelse(Nout_pair["endemic"] > 0,
                           FN_pair["endemic"] / Nout_pair["endemic"], NA)
        prop_epi <- ifelse(Nout_pair["epidemic"] > 0,
                           FN_pair["epidemic"] / Nout_pair["epidemic"], NA)
        mat <- matrix(c(FN_pair["endemic"], Nout_pair["endemic"] - FN_pair["endemic"],
                        FN_pair["epidemic"], Nout_pair["epidemic"] - FN_pair["epidemic"]),
                      nrow = 2, byrow = TRUE)
      }

      mat[!is.finite(mat)] <- 0
      mat[mat < 0] <- 0
      if (all(mat == 0)) return(tibble(method = NA_character_, statistic = NA_real_, 
                                        p.value = NA_real_, endemic = prop_end, epidemic = prop_epi))

      ## use expected counts to decide between Chi-square and Fisher
      expected <- tryCatch(chisq.test(mat, correct = FALSE)$expected, error = function(e) NULL)
      if (is.null(expected) || any(expected < 5)) {
        tt <- fisher.test(mat)
        method <- "Fisher exact"
        stat <- NA_real_
      } else {
        tt <- chisq.test(mat, correct = FALSE)
        method <- "Chi-square"
        stat <- tt$statistic
      }

      tibble(method = method, statistic = stat, p.value = tt$p.value,
             endemic = prop_end, epidemic = prop_epi)
    }) %>%
    ungroup() %>%
    mutate(test = "Chi-square/Fisher", metric = metric)

  ## Filter main
  if (show_only_main) {
    test_results <- test_results %>%
      mutate(alert_lab_clean = str_replace_all(alert_lab, "\u2265", ">=")) %>%
      filter(pop_brk == "[50k, 500k)", alert_lab_clean == ">= 50 weekly") %>%
      select(-alert_lab_clean)
  }

  all_results[[metric]] <- test_results
}

combined_table <- bind_rows(all_results) %>%
  mutate(
    `significance level` = case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ "ns"
    ),
    p.value = ifelse(
      !is.na(p.value) & p.value < 0.001,
      cell_spec(round(p.value, 3), color = "white", background = "#2b8c4d"),
      cell_spec(round(p.value, 3), color = "white", background = "#b30000")
    )
  ) %>%
  rename(`population group` = pop_brk, `alert definition` = alert_lab) %>%
  select(metric, `population group`, `alert definition`, test, method, endemic, epidemic, statistic, p.value, `significance level`)

## printing
kable(
  combined_table,
  format = "html",
  escape = FALSE,
  caption = "Wilcoxon and Chi-square/Fisher results (Endemic vs Epidemic)"
) %>%
  kable_styling(full_width = FALSE, position = "center") %>%
  row_spec(0, bold = TRUE, background = "#f8f8f8") %>%
  print()


```

## Trade-offs in utility dimensions

```{r}
#| label: prep_util_data
#| echo: false

def_impact_dat <- fig_impact_dat %>%
  dplyr::group_by(pop_brk) %>%
  dplyr::mutate(std_impact = as.vector(scale(impact_mean, center = TRUE, scale = TRUE))) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(pop_brk, desc(std_impact)) %>%
  dplyr::select(alert_number, alert_type, pop_brk, alert_lab, impact_mean, std_impact)

def_eff_dat <- fig_eff_dat %>%
  dplyr::group_by(pop_brk) %>%
  dplyr::mutate(std_eff = as.vector(scale(eff_mean, center = TRUE, scale = TRUE))) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(pop_brk, desc(std_eff)) %>%
  dplyr::select(alert_number, pop_brk, eff_mean, std_eff)

def_ppv_dat <- fig_ppv_dat %>%
  dplyr::group_by(pop_brk) %>%
  dplyr::mutate(std_ppv = as.vector(scale(ppv, center = TRUE, scale = TRUE))) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(pop_brk, desc(std_ppv)) %>%
  dplyr::select(alert_number, pop_brk, ppv, std_ppv, alerts, true_positives)

## multiply by -1 so that the scale is reversed (so higher standardized value represents higher utility (lower proportion of missed outbreaks and smaller delay mean higher utility))
def_missed_dat <- fig_missed_dat %>%
  dplyr::group_by(pop_brk) %>%
  dplyr::mutate(std_missed = as.vector(scale(missed_prop, center = TRUE, scale = TRUE)) *-1) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(pop_brk, desc(std_missed)) %>%
  dplyr::select(alert_number, pop_brk, missed_prop, std_missed)

## multiply by -1 so that the scale is reversed (so higher standardized value represents higher utility (lower proportion of missed outbreaks and smaller delay mean higher utility))
def_delay_dat <- fig_delay_dat %>%
  dplyr::group_by(pop_brk) %>%
  dplyr::mutate(std_delay = as.vector(scale(delay_mean, center = TRUE, scale = TRUE)) *-1) %>%
  dplyr::ungroup() %>%
  dplyr::select(alert_number, pop_brk, delay_mean, std_delay) %>%
  dplyr::arrange(pop_brk, desc(std_delay))
```

```{r}
#| label: prep_total_util_data
#| echo: false

def_complete_dat <- dplyr::full_join(def_impact_dat, def_eff_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::full_join(def_ppv_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::full_join(def_missed_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::full_join(def_delay_dat, by = c("alert_number", "pop_brk")) %>%
  OutbreakExtractR::add_alert_stringency() %>%
  dplyr::select(pop_brk, alert_lab, alert_number, alert_type, alert_stringency,
                impact_mean, eff_mean, ppv, missed_prop, delay_mean,
                std_impact, std_eff, std_ppv, std_missed, std_delay
  ) %>%
  dplyr::mutate(across(impact_mean:delay_mean, ~round(.x, 1)), across(contains("std_"), ~round(.x, 1))) %>%
  dplyr::mutate(util_score = rowSums(dplyr::across(contains("std_")))) %>%
  dplyr::arrange(desc(util_score)) 
  
## adding standard deviations/iqr and number of alerts/outbreaks denominators per metric

sd_impact_dat <- prep_impact_eff_dat %>%
  dplyr::group_by(alert_number, pop_brk) %>%
  dplyr::summarise(
    sd_impact = ifelse(sum(!is.na(impact)) > 1, sd(impact, na.rm = TRUE), 0)
  ) %>%
  dplyr::ungroup()

iqr_impact_dat <- prep_impact_eff_dat %>%
  dplyr::group_by(alert_number, pop_brk) %>%
  dplyr::summarise(
    iqr_impact = IQR(impact, na.rm = TRUE)
  ) %>%
  dplyr::ungroup()

sd_eff_dat <- prep_impact_eff_dat %>%
  dplyr::group_by(alert_number, pop_brk) %>%
  dplyr::summarise(
    sd_eff = ifelse(sum(!is.na(eff)) > 1, sd(eff, na.rm = TRUE), 0)
  ) %>%
  dplyr::ungroup()

iqr_eff_dat <- prep_impact_eff_dat %>%
  dplyr::group_by(alert_number, pop_brk) %>%
  dplyr::summarise(
    iqr_eff = IQR(eff, na.rm = TRUE)
  ) %>%
  dplyr::ungroup()

sd_delay_dat <- prep_missed_delay_dat %>%
  dplyr::group_by(alert_number, pop_brk) %>%
  dplyr::summarise(
    sd_delay = ifelse(sum(!is.na(delay)) > 1, sd(delay, na.rm = TRUE), 0)
  ) %>%
  dplyr::ungroup()

iqr_delay_dat <- prep_missed_delay_dat %>%
  dplyr::group_by(alert_number, pop_brk) %>%
  dplyr::summarise(
    iqr_delay = IQR(delay, na.rm = TRUE)
  ) %>%
  dplyr::ungroup()

## denominators
ppv_den_dat <- prep_ppv_dat %>%
  dplyr::group_by(alert_number, pop_brk) %>%
  dplyr::summarise(n_alerts = n()) %>%
  dplyr::ungroup()

missed_den_dat <- linked_cl_large %>%
  dplyr::distinct(outbreak_UID, pop_brk) %>%  ## one row per outbreak
  dplyr::group_by(pop_brk) %>%                ## group by population bracket
  dplyr::summarise(n_outbreaks = n()) %>%
  dplyr::ungroup()

## merge with main table
def_complete_dat_added <- def_complete_dat %>%
  dplyr::left_join(sd_impact_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::left_join(sd_eff_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::left_join(sd_delay_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::left_join(ppv_den_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::left_join(missed_den_dat, by = "pop_brk") %>% 
  dplyr::mutate(across(c(sd_impact, sd_eff, sd_delay), ~round(.x, 1))) %>%
  dplyr::select(
    pop_brk, alert_lab, alert_type, alert_stringency,
    impact_mean, sd_impact, eff_mean, sd_eff, ppv, n_alerts,
    missed_prop, n_outbreaks, delay_mean, sd_delay,
    std_impact, std_eff, std_ppv, std_missed, std_delay,
    util_score
  )

write.csv(def_complete_dat_added, "complete_utility_data_mean_sd.csv")

## Mar 2026 version of tables with IQR
def_complete_dat_added_IQR <- def_complete_dat %>%
  dplyr::left_join(iqr_impact_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::left_join(iqr_eff_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::left_join(iqr_delay_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::left_join(ppv_den_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::left_join(missed_den_dat, by = "pop_brk") %>% 
  dplyr::mutate(across(c(iqr_impact, iqr_eff, iqr_delay), ~round(.x, 1))) %>%
  dplyr::select(
    pop_brk, alert_lab, alert_type, alert_stringency,
    impact_mean, iqr_impact, eff_mean, iqr_eff, ppv, n_alerts,
    missed_prop, n_outbreaks, delay_mean, iqr_delay,
    std_impact, std_eff, std_ppv, std_missed, std_delay,
    util_score
  )


```

### Stringency by alert type, 50-500k

```{r}
#| label: fig-stringency-50-500
#| fig-width: 2.5
#| fig-height: 3.5
#| echo: true

f3a <- ggplot2::ggplot(dplyr::filter(def_complete_dat, pop_brk == "[50k, 500k)"),
                       aes(x = alert_stringency, y = util_score)) +
  geom_point(aes(colour = alert_type, alpha = alert_stringency)) +
  geom_hline(yintercept = 0, colour = "black", linetype = 2, alpha = 0.7) +
  scale_x_discrete("Alert Stringency") +
  scale_y_continuous("Utility Score \u2192", limits = c(-4, 4)) +
  scale_colour_brewer("Alert Type", palette = "Dark2") +
  theme_bw() +
  theme(legend.position = "none") +
  guides(alpha = "none")

f3a
```

### Stringency vs Dimension Score, 50-500k

```{r}
#| label: fig-stringency-utility-dimension-50-500
#| fig-width: 5
#| fig-height: 3.5
#| echo: true

def_complete_dat_longdim <- dplyr::select(def_complete_dat, pop_brk, alert_lab, alert_type, alert_stringency, dplyr::contains("std_")) %>%
  tidyr::pivot_longer(contains("std_"), names_to = "dimension", values_to = "score") %>%
  OutbreakExtractR::label_dimensions() 

f3b <- ggplot2::ggplot(dplyr::filter(def_complete_dat_longdim, pop_brk == "[50k, 500k)"),
                       aes(x = alert_stringency, y = score)) +
  geom_point(aes(colour = alert_type, alpha = alert_stringency)) +
  geom_hline(yintercept = 0, colour = "black", linetype = 2, alpha = 0.7) +
  scale_x_discrete("Alert Stringency") +
  scale_y_continuous("Dimension Score \u2192", limits = c(-3, 3)) +
  scale_colour_brewer("Alert Type", palette = "Dark2") +
  theme_bw() +
  guides(alpha = "none") +
  facet_wrap(~dimension, nrow = 2) +
  theme(
    legend.position = c(0.85, 0.25),
    legend.background = element_rect(fill = "white", colour = "black"),
    legend.key.size = unit(0.5, "lines")
  )

f3b
```

### Tradeoffs fig, 50-500k

Fix 3b so that the legend goes in the facet whitespace

```{r}
#| label: fig-tradeoffs-50-500
#| fig-width: 7.5
#| fig-height: 3.5
#| echo: true

ws <- c(.8, 1)

f3a_labeled <- gridExtra::arrangeGrob(
  f3a,
  top = grid::textGrob("A", x = 0, hjust = 0, gp = grid::gpar(fontsize = 14, fontface = "bold"))
)

f3b_labeled <- gridExtra::arrangeGrob(
  f3b,
  top = grid::textGrob("B", x = 0, hjust = 0, gp = grid::gpar(fontsize = 14, fontface = "bold"))
)

p2 <- gridExtra::arrangeGrob(
  f3a_labeled,
  f3b_labeled,
  widths = ws,
  nrow = 1,
  padding = unit(0.6, "line")
)

ggplot2::ggsave(
  paste0(fig_path,"/fig2_tradeoff_scatterplots","_",ocv_status,".png"),
  p2,
  width = 7.5,
  height = 3,
  dpi = 300
)

# print in the report
grid::grid.newpage()
grid::grid.draw(p2)

```

### Supplement Methods

#### Sensitivity of Utility Score Calculation 

-   Rank sum score (TODO) - code needs to be updated. Rank dimensions lowest to highest and sum the ranks. Sum dimension ranks to get V2 of the utility score. Then compare the two utility scores in a scatterplot.
-   Is the best-performing alert definition different?

```{r}
#| label: prep_total_util_data_sens
#| echo: false
#| eval: false

def_complete_dat_v2 <- dplyr::full_join(def_impact_dat, def_eff_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::full_join(def_ppv_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::full_join(def_missed_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::full_join(def_delay_dat, by = c("alert_number", "pop_brk")) %>%
  OutbreakExtractR::add_alert_stringency() %>%
  dplyr::select(pop_brk, alert_lab, alert_type, alert_stringency,
                impact_mean, eff_mean, ppv, missed_prop, delay_mean,
                std_impact, std_eff, std_ppv, std_missed, std_delay
  ) %>%
  dplyr::mutate(across(impact_mean:delay_mean, ~round(.x, 1)), across(contains("std_"), ~round(.x, 1))) %>%
  dplyr::mutate(util_score_v2 = rowSums(dplyr::across(contains("std_")))) %>%
  dplyr::arrange(desc(util_score_v2)) 
  
```



```{r}
#| label: utility_sens_analysis
#| echo: true

library(ggrepel)
library(patchwork)  

## clean labels for plots
clean_labels <- c(
  std_impact = "Std Impact",
  rank_impact = "Ranked Impact",
  std_eff = "Std Efficiency",
  rank_eff = "Ranked Efficiency",
  std_ppv = "Std PPV",
  rank_ppv = "Ranked PPV",
  std_missed = "Std Missed",
  rank_missed = " Ranked Missed",
  std_delay = "Std Delay",
  rank_delay = "Ranked Delay",
  util_score = "Utility Score",
  util_score_v2 = "Utility Score v2",
  pop_brk = "Population"
)

## scatterplot function
plot_comparison <- function(compare_dat, x_col, y_col, color_var = "pop_brk") {
  ggplot(compare_dat, aes(
    x = !!rlang::sym(x_col),
    y = !!rlang::sym(y_col),
    color = !!rlang::sym(color_var)
  )) +
    geom_point(size = 3) +
    geom_smooth(method = "lm", se = FALSE, linetype = "dashed", color = "gray40") +
    ggrepel::geom_text_repel(aes(label = alert_lab), size = 3) +
    theme_minimal(base_size = 12) +
    labs(
      x = clean_labels[[x_col]],
      y = clean_labels[[y_col]],
      color = clean_labels[[color_var]]
    )
}

## Rank-sum utility score function
calc_rank_sum_util_high_good <- function(dat,
                                         metrics_high_better = c("impact_mean", "eff_mean", "ppv"),
                                         metrics_low_better  = c("missed_prop", "delay_mean")) {

  dat %>%
    dplyr::group_by(pop_brk) %>%
    dplyr::mutate(
      ## high-better metrics: high value gives high rank
      across(all_of(metrics_high_better), ~ dplyr::min_rank(.x), .names = "rank_{.col}"),

      ## Low-better metrics: low value gives high rank
      across(all_of(metrics_low_better), ~ dplyr::min_rank(max(.x, na.rm = TRUE) - .x), .names = "rank_{.col}")
    ) %>%
    ## sum all ranks: higher total = better alert
    dplyr::mutate(util_score_v2 = rowSums(dplyr::across(starts_with("rank_")), na.rm = TRUE)) %>%
    dplyr::arrange(pop_brk, dplyr::desc(util_score_v2)) %>%  
    dplyr::ungroup()  
}


## combining datasets and computing utility version 2
def_complete_dat_v2 <- dplyr::full_join(def_impact_dat, def_eff_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::full_join(def_ppv_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::full_join(def_missed_dat, by = c("alert_number", "pop_brk")) %>%
  dplyr::full_join(def_delay_dat, by = c("alert_number", "pop_brk")) %>%
  OutbreakExtractR::add_alert_stringency() %>%
  dplyr::select(pop_brk, alert_lab, alert_type, alert_stringency,
                impact_mean, eff_mean, ppv, missed_prop, delay_mean)

def_complete_dat_sens <- calc_rank_sum_util_high_good(def_complete_dat_v2)

## utility score comparison
compare_scores_util <- def_complete_dat %>%
  dplyr::select(alert_lab, pop_brk, util_score) %>%
  dplyr::left_join(
    def_complete_dat_sens %>%
      dplyr::select(alert_lab, pop_brk, util_score_v2),
    by = c("alert_lab", "pop_brk")
  )

plot_util_total <- plot_comparison(compare_scores_util, x_col = "util_score", y_col = "util_score_v2")
print(plot_util_total)


## compare standardised metrics to ranked raw metrics
compare_scores_dim <- def_complete_dat %>%
  dplyr::select(alert_lab, pop_brk, std_impact, std_eff, std_ppv, std_missed, std_delay) %>%
  dplyr::left_join(
    def_complete_dat_sens %>%
      dplyr::select(alert_lab, pop_brk,
                    rank_impact_mean, rank_eff_mean, rank_ppv,
                    rank_missed_prop, rank_delay_mean),
    by = c("alert_lab", "pop_brk")
  ) %>%
  
  ## renaming for clarity
  dplyr::rename(
    rank_impact = rank_impact_mean,
    rank_eff    = rank_eff_mean,
    rank_ppv    = rank_ppv,
    rank_missed = rank_missed_prop,
    rank_delay  = rank_delay_mean
  )

dimension_pairs <- list(
  impact = c("std_impact", "rank_impact"),
  eff    = c("std_eff", "rank_eff"),
  ppv    = c("std_ppv", "rank_ppv"),
  missed = c("std_missed", "rank_missed"),
  delay  = c("std_delay", "rank_delay")
)

dimension_plots <- lapply(names(dimension_pairs), function(dim) {
  x_col <- dimension_pairs[[dim]][1]
  y_col <- dimension_pairs[[dim]][2]
  plot_comparison(compare_scores_dim, x_col = x_col, y_col = y_col)
})


all_plots <- (
  plot_util_total +
  dimension_plots[[1]] + dimension_plots[[2]] +
  dimension_plots[[3]] + dimension_plots[[4]] + dimension_plots[[5]]
) +
  plot_layout(ncol = 2, guides = "collect") +
  plot_annotation(
    title = "Comparison of Standardised and Raw (Ranked) Metrics Across Dimensions",
    theme = theme(
      plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
      axis.title.y = element_text(margin = margin(r = 10)),  # adds space to y labels
      plot.margin = margin(15, 10, 15, 10)
    )
  )

all_plots

```

#### Sensitivity of Utility Score by country 

-   Examine country-specific utility scores and compare them to the overall one. If there is a lot of heterogeneity in the best-performing alert definition across countries, we can comment on how we really need context-dependent alert definitions.
-   N.B. Include only outbreak-prone subnational locations for the main population group

```{r}
#| label: prep_country_utility_data
#| echo: true

means_impact_eff_country_dat <- prep_impact_eff_dat %>%
  dplyr::filter(pop_brk == "[50k, 500k)") %>% ## we are only interested in the main population group
  dplyr::group_by(alert_number, alert_type, country) %>%
  dplyr::summarise(impact_mean = mean(impact), eff_mean = mean(eff)) %>%
  dplyr::ungroup()

means_delay_country_dat <- prep_missed_delay_dat %>%
  dplyr::filter(pop_brk == "[50k, 500k)") %>% ## we are only interested in the main population group
  dplyr::group_by(alert_number, alert_type, country) %>%
  dplyr::summarise(delay_mean = mean(delay, na.rm = TRUE)) %>%
  dplyr::ungroup()

n_country_outbreaks <- linked_cl_large %>%
  dplyr::filter(pop_brk == "[50k, 500k)") %>% ## we are only interested in the main population group
  dplyr::distinct(outbreak_UID, country) %>%
  dplyr::group_by(country) %>%
  dplyr::summarise(n_outbreaks = n())

fig_impact_country_dat <- dplyr::group_by(prep_impact_eff_dat %>% dplyr::filter(pop_brk == "[50k, 500k)"),
                                          alert_number, alert_type, country) %>%
  dplyr::reframe(quibble2(impact)) %>%
  tidyr::pivot_wider(names_from = impact_q, names_glue = "{.value}_{impact_q}", values_from = impact) %>%
  dplyr::full_join(dplyr::select(means_impact_eff_country_dat, alert_number, alert_type, country, impact_mean), by = c("alert_number", "alert_type", "country")) %>%
  prepare_fig_dat()

fig_eff_country_dat <- dplyr::group_by(prep_impact_eff_dat %>% dplyr::filter(pop_brk == "[50k, 500k)"), alert_number, alert_type, country) %>%
  dplyr::reframe(quibble2(eff)) %>%
  tidyr::pivot_wider(names_from = eff_q, names_glue = "{.value}_{eff_q}", values_from = eff) %>%
  dplyr::full_join(dplyr::select(means_impact_eff_country_dat, alert_number, alert_type, country, eff_mean), by = c("alert_number", "alert_type", "country")) %>%
  prepare_fig_dat()

fig_ppv_country_dat <- dplyr::group_by(prep_ppv_dat %>% dplyr::filter(pop_brk == "[50k, 500k)"),
                                       alert_number, alert_type, country) %>%
  dplyr::summarise(alerts = n(), true_positives = sum(true_positives), ppv = round(true_positives/alerts, 2)) %>%
  prepare_fig_dat()

fig_missed_country_dat <- dplyr::group_by(prep_missed_delay_dat %>% dplyr::filter(pop_brk == "[50k, 500k)"),
  alert_number, alert_type, country) %>%
  dplyr::summarise(n_detected = sum(is_triggered)) %>%
  dplyr::left_join(n_country_outbreaks, by = c("country")) %>%
  dplyr::ungroup() %>%
  dplyr::mutate(n_missed = n_outbreaks - n_detected,
                missed_prop = n_missed/n_outbreaks) %>%
  prepare_fig_dat()

fig_delay_country_dat <- dplyr::group_by(prep_missed_delay_dat %>% dplyr::filter(pop_brk == "[50k, 500k)"),
  alert_number, alert_type, country) %>%
  dplyr::reframe(quibble2(delay)) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_wider(names_from = delay_q, names_glue = "{.value}_{delay_q}", values_from = delay) %>%
  dplyr::full_join(means_delay_country_dat, by = c("alert_number", "alert_type", "country")) %>%
  prepare_fig_dat()

fig_complete_country_dat <- dplyr::full_join(fig_impact_country_dat, fig_eff_country_dat, by = c("alert_number", "alert_type", "country", "alert_lab")) %>%
  dplyr::full_join(fig_ppv_country_dat, by = c("alert_number", "alert_type", "country", "alert_lab")) %>%
  dplyr::full_join(fig_missed_country_dat, by = c("alert_number", "alert_type", "country", "alert_lab")) %>%
  dplyr::full_join(fig_delay_country_dat, by = c("alert_number", "alert_type", "country", "alert_lab"))


def_impact_country_dat <- fig_impact_country_dat %>%
  dplyr::group_by(country) %>%
  dplyr::mutate(std_impact = as.vector(scale(impact_mean, center = TRUE, scale = TRUE))) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(country, desc(std_impact)) %>%
  dplyr::select(alert_number, alert_type, country, alert_lab, impact_mean, std_impact)

def_eff_country_dat <- fig_eff_country_dat %>%
  dplyr::group_by(country) %>%
  dplyr::mutate(std_eff = as.vector(scale(eff_mean, center = TRUE, scale = TRUE))) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(country, desc(std_eff)) %>%
  dplyr::select(alert_number, alert_type, country, alert_lab, eff_mean, std_eff)

def_ppv_country_dat <- fig_ppv_country_dat %>%
  dplyr::group_by(country) %>%
  dplyr::mutate(std_ppv = as.vector(scale(ppv, center = TRUE, scale = TRUE))) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(country, desc(std_ppv)) %>%
  dplyr::select(alert_number, alert_type, country, alert_lab, ppv, std_ppv, alerts, true_positives)

## multiply by -1 so that the scale is reversed (so higher standardized value represents higher utility (lower proportion of missed outbreaks and smaller delay mean higher utility))
def_missed_country_dat <- fig_missed_country_dat %>%
  dplyr::group_by(country) %>%
  dplyr::mutate(std_missed = as.vector(scale(missed_prop, center = TRUE, scale = TRUE)) *-1) %>%
  dplyr::ungroup() %>%
  dplyr::arrange(country, desc(std_missed)) %>%
  dplyr::select(alert_number, alert_type, country, alert_lab, missed_prop, std_missed)

## multiply by -1 so that the scale is reversed (so higher standardized value represents higher utility (lower proportion of missed outbreaks and smaller delay mean higher utility))
def_delay_country_dat <- fig_delay_country_dat %>%
  dplyr::group_by(country) %>%
  dplyr::mutate(std_delay = as.vector(scale(delay_mean, center = TRUE, scale = TRUE)) *-1) %>%
  dplyr::ungroup() %>%
  dplyr::select(alert_number, alert_type, country, alert_lab, delay_mean, std_delay) %>%
  dplyr::arrange(country, desc(std_delay))

def_complete_country_dat <- dplyr::full_join(def_impact_country_dat, def_eff_country_dat, by = c("alert_number","alert_type", "country", "alert_lab")) %>%
  dplyr::full_join(def_ppv_country_dat, by = c("alert_number", "alert_type", "country", "alert_lab")) %>%
  dplyr::full_join(def_missed_country_dat, by = c("alert_number", "alert_type", "country", "alert_lab")) %>%
  dplyr::full_join(def_delay_country_dat, by = c("alert_number","alert_type", "country", "alert_lab")) %>%
  OutbreakExtractR::add_alert_stringency() %>%
  dplyr::select(country, alert_lab, alert_type, alert_stringency,
                impact_mean, eff_mean, ppv, missed_prop, delay_mean,
                std_impact, std_eff, std_ppv, std_missed, std_delay
  ) %>%
  dplyr::mutate(across(impact_mean:delay_mean, ~round(.x, 1)), across(contains("std_"), ~round(.x, 1))) %>%
  dplyr::mutate(util_score = rowSums(dplyr::across(contains("std_")))) %>%
  dplyr::arrange(desc(util_score)) 

write.csv(def_complete_country_dat, "complete_country_utility_data_mean_sd.csv")
  
```

### Supplement Fig Proposal

#### Stringency vs Utility score, all population groups

```{r}
#| label: fig-stringency-utility
#| fig-width: 4
#| fig-height: 4
#| echo: true

ggplot2::ggplot(def_complete_dat, aes(x = util_score, y = alert_stringency)) +
  geom_point(aes(colour = alert_type, alpha = alert_stringency)) +
  scale_x_continuous("Utility Score") +
  scale_y_discrete("Alert Stringency") +
  scale_colour_brewer("Alert Type", palette = "Dark2") +
  theme_bw() +
  theme(legend.position = "right") +
  guides(alpha = "none")

```

#### Dimension score pairs plot, all population groups

```{r}
#| label: fig-score-tradeoffs
#| fig-width: 6
#| fig-height: 6
#| echo: true

fig_score_allpop <- GGally::ggpairs(def_complete_dat, 
                columns = 10:14, 
                ggplot2::aes(colour = alert_type, fill = alert_type, alpha = 0.8),
                upper = "blank", 
                lower = list(continuous = "points"),
                switch = "both",
                axisLabels = "none",
#                columnLabels = c("Impact", "Efficiency", "PPV", "Missed", "Delay")) +
                columnLabels = c("Impact", "Efficiency", "PPV", "Missed", "Timeliness")) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer("Alert Type", palette = "Dark2") +
  theme_bw() 

fig_score_allpop

## Pooled spearman correlations

data_subset <- def_complete_dat[, 10:14]
#labels <- c("Impact", "Efficiency", "PPV", "Missed", "Delay")
labels <- c("Impact", "Efficiency", "PPV", "Missed", "Timeliness")
colnames(data_subset) <- labels

vars <- colnames(data_subset)
pairs <- combn(vars, 2, simplify = FALSE)

corr_results <- map_dfr(pairs, ~{
  test <- cor.test(data_subset[[.x[1]]], data_subset[[.x[2]]], 
                   method = "spearman", exact = FALSE)
  tibble(
    Dimension1 = .x[1],
    Dimension2 = .x[2],
    Correlation = round(test$estimate, 2),
    P_value = test$p.value
  )
})

corr_results %>%
  kable(caption = "Pairwise Spearman correlations with p-values") %>%
  kable_styling(full_width = FALSE)

```

### Supplement Fig Backup

#### Stringency by Total Utility, faceted by population

```{r}
#| label: fig-stringency-score-supp
#| fig-width: 4
#| fig-height: 3
#| echo: true

ggplot2::ggplot(def_complete_dat, aes(x = util_score, y = alert_stringency)) +
  geom_point(aes(colour = alert_type, alpha = alert_stringency)) +
  scale_x_continuous("Utility Score") +
  scale_y_discrete("Stringency") +
  scale_colour_brewer("Alert Type", palette = "Dark2") +
  theme_bw() +
  theme(legend.position = "right") +
  guides(alpha = "none") +
  facet_wrap(~pop_brk, ncol = 1)
```

#### Stringency vs Dimension Score, faceted by population

```{r}
#| label: fig-stringency-utility-dimension
#| fig-width: 5
#| fig-height: 5
#| echo: true

def_complete_dat_longdim <- dplyr::select(def_complete_dat, pop_brk, alert_lab, alert_type, alert_stringency, dplyr::contains("std_")) %>%
  tidyr::pivot_longer(contains("std_"), names_to = "dimension", values_to = "score") %>%
  dplyr::mutate(dimension = stringr::str_remove(dimension, "std_"))

ggplot2::ggplot(def_complete_dat_longdim, aes(x = score, y = alert_stringency)) +
  geom_point(aes(colour = alert_type, alpha = alert_stringency)) +
  scale_x_continuous("Dimension Score") +
  scale_y_discrete("Alert Stringency") +
  scale_colour_brewer("Alert Type", palette = "Dark2") +
  theme_bw() +
  theme(legend.position = "bottom") +
  guides(alpha = "none") +
  facet_grid(pop_brk~ dimension)

```

#### Dimension score pairs plot, 50-500k

```{r}
#| label: fig-dimension-score-pairs
#| fig-width: 6
#| fig-height: 6
#| echo: true

GGally::ggpairs(dplyr::filter(def_complete_dat, pop_brk == "[50k, 500k)"), 
                columns = 10:14, 
                ggplot2::aes(colour = alert_type, fill = alert_type, alpha = 0.8),
                upper = "blank", 
                lower = list(continuous = "points"),
                switch = "both",
                axisLabels = "none",
#                columnLabels = c("Impact", "Efficiency", "PPV", "Missed", "Delay")) +
                columnLabels = c("Impact", "Efficiency", "PPV", "Missed", "Timeliness")) +
  scale_color_brewer(palette = "Dark2") +
  scale_fill_brewer("Alert Type", palette = "Dark2") +
  theme_bw() 

```

## Best-performing alert definitions

### Top 4 Raw Metrics and Total Score Table for Different Dimension Score Cutoffs

-   Added SD for impact, efficiency, and delay metrics
-   Added number of alerts for PPV metric (denominator)
-   Added number of outbreaks for missed metric (denominator)

```{r}
#| label: tab-top5
#| echo: false
#| eval: false

tab_top_def_dat <- def_complete_dat %>%
  dplyr::select(pop_brk, alert_lab, alert_type, impact_mean, eff_mean, ppv, missed_prop, delay_mean, util_score) %>%
  dplyr::group_by(pop_brk) %>%
  dplyr::arrange(pop_brk, desc(util_score)) %>%
  dplyr::mutate(rank = 1:n()) %>%
  dplyr::ungroup() %>%
  dplyr::filter(rank <= 5) %>%
  dplyr::select(-rank)

reactable_template(tab_top_def_dat,
                  col_names = list(pop_brk = reactable::colDef(name = "Population"),
                                   alert_lab = reactable::colDef(name = "Alert Definition"),
                                   alert_type = reactable::colDef(name = "Alert Type"),
                                   impact_mean = reactable::colDef(name = "Impact: cases"),
                                   eff_mean = reactable::colDef(name = "Efficiency: per target pop"),
                                   ppv = reactable::colDef(name = "PPV >300 cases: prop"),
                                   missed_prop = reactable::colDef(name = "Missed Outbreaks: prop"),
                                   delay_mean = reactable::colDef(name = "Delay: weeks from outbreak start"),
                                   util_score = reactable::colDef(name = "Util Score", defaultSortOrder = "desc")))


```


```{r}
#| label: prepare-top4-data
#| echo: true

## automating best definition selection based on quantitative threshold
full_dat <- def_complete_dat %>%
  dplyr::select(pop_brk, alert_lab, util_score,
                std_impact, std_eff, std_ppv, std_missed, std_delay) 

## make datasets long for plotting
full_long <- full_dat %>%
  pivot_longer(cols = c(std_impact, std_eff, std_ppv, std_missed, std_delay),
               names_to = "dimension", values_to = "score") %>%
  OutbreakExtractR::label_dimensions()


```


```{r, fig.width=20, fig.height=15}
#| label: top4-histogram
#| echo: true

## function to automate quantitative cutoff exploration
get_top_for_threshold <- function(full_dat, cutoff_val, no_top = 4) {
  score_cols <- c("std_impact", "std_eff", "std_ppv", "std_missed", "std_delay")
  
  top4_dat <- full_dat %>%
    rowwise() %>%
    mutate(min_score = min(c_across(all_of(score_cols)), na.rm = TRUE)) %>%
    ungroup() %>%
    filter(min_score >= cutoff_val) %>%
    group_by(pop_brk) %>%
    arrange(desc(util_score)) %>%
    slice_head(n = no_top) %>%
    ungroup()
  
  return(top4_dat)
}

## function to make histograms per threshold
make_static_top4_counts <- function(full_long, full_dat, cutoff_val, bins = 20) {

  ## filter top 4 per group based on cutoff
  top4_dat <- get_top_for_threshold(full_dat, cutoff_val, 4)
  
  if(nrow(top4_dat) > 0){
    ## long format
    top4_long <- top4_dat %>%
      pivot_longer(
        cols = c(std_impact, std_eff, std_ppv, std_missed, std_delay),
        names_to = "dimension",
        values_to = "score"
      ) %>%
      OutbreakExtractR::label_dimensions() %>%
      dplyr::select(pop_brk, alert_lab, dimension, score, util_score)
  } else {
    top4_long <- NULL
  }
  
  x_range <- range(full_long$score, na.rm = TRUE)
  
  ## histograms
  p <- ggplot(full_long, aes(x = score)) +
    geom_histogram(
      bins = bins,
      fill = "lightblue",
      color = "white",
      alpha = 0.6
    )  +
    facet_wrap(~ pop_brk + dimension,
               scales = "fixed",
               ncol = length(unique(full_long$dimension))) +
    coord_cartesian(xlim = x_range, expand = TRUE) +
    scale_x_continuous(expand = expansion(mult = 0.05)) +
    labs(
      title = paste0("Cutoff ≥ ", cutoff_val),
      x = "Dimension score",
      y = "Count"
    ) +
    theme_minimal(base_size = 24) +
    theme(
      strip.text = element_text(face = "bold"),
      panel.grid.minor = element_blank(),
      panel.spacing = unit(2, "lines"),
      plot.margin = margin(10, 10, 10, 10)
    )

  ## lines per alert
  if(!is.null(top4_long)){
    p <- p +
      geom_vline(
        data = top4_long,
        aes(
          xintercept = score,
          color = alert_lab,  
          group = alert_lab   
        ),
        linetype = "dashed",
        linewidth = 0.6,
        alpha = 0.4
      ) +
      scale_color_viridis_d(option = "turbo", name = "Top 4 Alerts")
  }

  return(p)
}


# plots for different cutoffs
cutoff_values <- seq(from = -1, to = -0.5, by = 0.1)
for(thresh in cutoff_values){
  p <- make_static_top4_counts(full_long, full_dat, cutoff_val = thresh) +
       ggtitle(paste0("Cutoff Threshold: ", thresh))
  print(p)  
}

```

```{r}
#| label: tab-top4-by-cutoff
#| echo: true
#| results: asis

## choose whether to use SD or IQR metrics

if (use_iqr) {
  metrics_dat <- def_complete_dat_added_IQR
  spread_impact <- "iqr_impact"
  spread_eff <- "iqr_eff"
  spread_delay <- "iqr_delay"
  spread_label <- "IQR"
} else {
  metrics_dat <- def_complete_dat_added
  spread_impact <- "sd_impact"
  spread_eff <- "sd_eff"
  spread_delay <- "sd_delay"
  spread_label <- "SD"
}

top_tables_list <- list()
n <- 4
for (thresh in cutoff_values) {
  
  ## get top alerts above threshold
  top_dat <- get_top_for_threshold(
    full_dat,
    cutoff_val = thresh,
    no_top = n
  )
  
  if (nrow(top_dat) > 0) {
    
    ## merge additional metrics
    top_table <- top_dat %>%
      dplyr::left_join(
        metrics_dat %>%
          dplyr::select(
            pop_brk, alert_lab, alert_type,
            impact_mean, !!spread_impact,
            eff_mean, !!spread_eff,
            ppv, n_alerts,
            missed_prop, n_outbreaks,
            delay_mean, !!spread_delay
          ) %>%
          dplyr::distinct(pop_brk, alert_lab, .keep_all = TRUE),
        by = c("pop_brk", "alert_lab")
      ) %>%
      dplyr::mutate(
        impact_mean = paste0(impact_mean, " (", .data[[spread_impact]], ")"),
        eff_mean = paste0(eff_mean, " (", .data[[spread_eff]], ")"),
        delay_mean = paste0(delay_mean, " (", .data[[spread_delay]], ")"),
        ppv = paste0(ppv, " (", n_alerts, ")"),
        missed_prop = paste0(missed_prop, " (", n_outbreaks, ")")
      ) %>%
      dplyr::select(
        pop_brk, alert_lab, alert_type,
        impact_mean, eff_mean, ppv,
        missed_prop, delay_mean, util_score
      ) %>%
      dplyr::group_by(pop_brk) %>%
      dplyr::arrange(pop_brk, desc(util_score)) %>%
      dplyr::mutate(rank = dplyr::row_number()) %>%
      dplyr::ungroup() %>%
      dplyr::filter(rank <= n) %>%
      dplyr::select(-rank)
    
    
    ## label population groups
    top_table <- top_table %>%
      dplyr::mutate(
        pop_brk = forcats::fct_recode(
          pop_brk,
          "Administrative units with <50,000 people" = "< 50k",
          "Administrative units with 50,000 to 500,000 people" = "[50k, 500k)",
          "Administrative units with ≥500,000 people" = "≥ 500k"
        )
      ) %>%
      dplyr::arrange(pop_brk, desc(util_score))
    
    
    ## count rows per group
    group_counts <- top_table %>%
      dplyr::count(pop_brk) %>%
      dplyr::pull(n)
    
    top_rows <- cumsum(c(1, head(group_counts, -1)))
    
    
    ## rename columns for table
    tab_for_kable <- top_table %>%
      dplyr::select(
        pop_brk,
        alert_lab,
        alert_type,
        impact_mean,
        eff_mean,
        ppv,
        missed_prop,
        delay_mean,
        util_score
      ) %>%
      dplyr::rename(
        "Alert Definition" = alert_lab,
        "Alert Type" = alert_type,
        !!paste0("Impact: cases (", spread_label, ")") := impact_mean,
        !!paste0("Efficiency: cases per 1000 target population (", spread_label, ")") := eff_mean,
        "PPV: proportion of alerts (N alerts)" = ppv,
        "Missed: proportion of outbreaks (N outbreaks)" = missed_prop,
        !!paste0("Delay: weeks from outbreak start (", spread_label, ")") := delay_mean,
        "Utility Score" = util_score
      )
    
    ## save csv
    write.csv(
      tab_for_kable,
      file = paste0(fig_path,"/top_alerts_cutoff_", thresh,"_",ocv_status,".csv"),
      row.names = FALSE
    )
    
    
    ## create kable
    table_kbl <- tab_for_kable %>%
      dplyr::select(-pop_brk) %>%
      kableExtra::kbl(
        caption = paste0(
          "Top Alerts by Utility Score per Population Group (Cutoff ≥ ",
          thresh, ")"
        )
      ) %>%
      kableExtra::kable_styling(
        latex_options = "basic",
        full_width = TRUE,
        position = "center"
      )
    
    
    ## add grouped rows automatically
    start_row <- 1
    
    for (i in seq_along(group_counts)) {
      
      end_row <- start_row + group_counts[i] - 1
      
      table_kbl <- table_kbl %>%
        kableExtra::group_rows(
          group_label = levels(top_table$pop_brk)[i],
          start_row = start_row,
          end_row = end_row
        )
      
      start_row <- end_row + 1
    }
    
    
    ## highlight top alert per group
    table_kbl <- table_kbl %>%
      kableExtra::row_spec(
        top_rows,
        bold = TRUE,
        background = "#d4edda"
      )
    
    
    top_tables_list[[as.character(thresh)]] <- table_kbl
    
  } else {
    
    top_tables_list[[as.character(thresh)]] <-
      htmltools::tags$p(
        paste0(
          "Cutoff ≥ ",
          thresh,
          ": No alerts meet the threshold."
        )
      )
  }
}

for (tbl in top_tables_list) {
  print(tbl)
}

```

### Table with top x alerts per pop group by utility score plus one that has no dimension under the cutoff

#### Main population group

```{r}
#| label: tab-top4-manuscript_mainpop
#| echo: true
#| results: asis

## function to get highest utility alert based on quant cutoff
get_top_for_threshold <- function(full_dat, cutoff_val, no_top = 1, pop_keep = NULL) {

  score_cols <- c("std_impact", "std_eff", "std_ppv", "std_missed", "std_delay")

  dat_with_min <- full_dat %>%
    rowwise() %>%
    mutate(min_score = min(c_across(all_of(score_cols)), na.rm = TRUE)) %>%
    ungroup()

  if (!is.null(pop_keep)) {
    dat_with_min <- dat_with_min %>%
      dplyr::filter(pop_brk %in% pop_keep)
  }

  ## top alert by util_score per pop_brk (no filtering by cutoff)
  top_n_per_pop <- dat_with_min %>%
    group_by(pop_brk) %>%
    arrange(desc(util_score)) %>%
    slice_head(n = no_top) %>%
    ungroup()

  ## best alert per pop that passes all dims
  best_passing_per_pop <- dat_with_min %>%
    filter(min_score >= cutoff_val) %>%
    group_by(pop_brk) %>%
    arrange(desc(util_score)) %>%
    slice_head(n = 1) %>%
    ungroup()

  bind_rows(top_n_per_pop, best_passing_per_pop) %>%
    distinct(pop_brk, alert_lab, .keep_all = TRUE)
}

make_top_alerts_table_select_pop <- function(full_dat, def_dat, cutoff = -1, pop_keep = NULL) {
  
  top_dat <- get_top_for_threshold(
    full_dat,
    cutoff_val = cutoff,
    no_top = 1,
    pop_keep = pop_keep
  )
  
  if (nrow(top_dat) == 0) return(NULL)
  
  long_labels <- function(x) {
    case_when(
      x == "< 50k" ~ "Administrative units with <50,000 people",
      x == "[50k, 500k)" ~ "Administrative units with 50,000 to 500,000 people",
      x == "≥ 500k" ~ "Administrative units with ≥500,000 people",
      TRUE ~ x
    )
  }
  
  top_table <- top_dat %>%
    left_join(
      def_dat %>%
        select(
          pop_brk, alert_lab, alert_type,
          impact_mean, sd_impact,
          eff_mean, sd_eff,
          ppv, n_alerts,
          missed_prop, n_outbreaks,
          delay_mean, sd_delay
        ) %>%
        distinct(pop_brk, alert_lab, .keep_all = TRUE),
      by = c("pop_brk", "alert_lab")
    ) %>%
    mutate(
      impact_mean = paste0(impact_mean, " (", sd_impact, ")"),
      eff_mean = paste0(eff_mean, " (", sd_eff, ")"),
      delay_mean = paste0(delay_mean, " (", sd_delay, ")"),
      ppv = paste0(ppv, " (", n_alerts, ")"),
      missed_prop = paste0(missed_prop, " (", n_outbreaks, ")"),
      best_passing_flag = !is.na(min_score) & min_score >= cutoff,
      pop_brk = factor(long_labels(pop_brk))  # apply descriptive labels
    ) %>%
    arrange(pop_brk, desc(util_score)) %>%
    group_by(pop_brk) %>%
    mutate(rank = row_number()) %>%
    ungroup()
  
  best_pass_rows <- which(top_table$best_passing_flag)
  top_table <- top_table %>% select(-best_passing_flag)
  
  tab_for_kable <- top_table %>%
    select(alert_lab, alert_type, impact_mean, eff_mean, ppv, missed_prop, delay_mean, util_score) %>%
    rename(
      "Alert Definition" = alert_lab,
      "Alert Type" = alert_type,
      "Impact: cases (SD)" = impact_mean,
      "Efficiency: cases per 1000 target population (SD)" = eff_mean,
      "PPV: proportion of alerts (N alerts)" = ppv,
      "Missed: proportion of outbreaks (N outbreaks)" = missed_prop,
      "Delay: weeks from outbreak start (SD)" = delay_mean,
      "Utility Score" = util_score
    )
  
  table_kbl <- tab_for_kable %>%
    kbl(caption = paste0("Top Alerts by Utility Score per Population Group (Cutoff ≥ ", cutoff, ")")) %>%
    kable_styling(latex_options = "basic", full_width = TRUE, position = "center")
  
  # Add group rows
  group_counts <- top_table %>% count(pop_brk) %>% pull(n)
  start <- 1
  for (i in seq_along(group_counts)) {
    table_kbl <- table_kbl %>%
      group_rows(levels(top_table$pop_brk)[i], start_row = start, end_row = start + group_counts[i] - 1)
    start <- start + group_counts[i]
  }
  
  ## highlight top row per group
  top_rows <- cumsum(c(1, head(group_counts, -1)))
  table_kbl <- table_kbl %>% row_spec(top_rows, bold = TRUE, background = "white")
  
  ## highlight best-passing rows
  if (length(best_pass_rows) > 0) {
    table_kbl <- table_kbl %>% row_spec(best_pass_rows, bold = TRUE, background = "white")
  }
  
  return(table_kbl)
}


table_main <- make_top_alerts_table_select_pop(
  full_dat,
  def_complete_dat_added,
  pop_keep = "[50k, 500k)"  ## main population group
)
print(table_main)

```

#### Population groups in the supplement

```{r}
#| label: tab-top4-manuscript_supplement_popgroups
#| echo: true
#| results: asis

table_small_large <- make_top_alerts_table_select_pop(
  full_dat,
  def_complete_dat_added,
  pop_keep = c("< 50k", "≥ 500k")
)
print(table_small_large)


```

#### All population groups

```{r}
#| label: tab-top4-manuscript_all_popgroups
#| echo: true
#| results: asis

## all populations

table_all <- make_top_alerts_table_select_pop(full_dat, def_complete_dat_added)
print(table_all)

```

### NEW 24 Nov: Tables with top x alerts (one per pop group) by utility score and cutoff

```{r}
#| label: tab-top4-version3
#| echo: true
#| results: asis

## function compute failing dimensions 
get_fail_dims <- function(row, cutoff) {
  dims <- c("std_impact", "std_eff", "std_ppv", "std_missed", "std_delay")
  # descriptive names
  name_map <- c(
    std_impact = "impact",
    std_eff = "efficiency",
    std_ppv = "ppv",
    std_missed = "missed",
    std_delay = "delay"
  )
  
  fails <- dims[row[dims] < cutoff]
  if (length(fails) == 0) return(NA_character_)
  
  # convert to descriptive names
  clean_names <- name_map[fails]
  paste0(clean_names, "<", cutoff, collapse = ", ")
}


## function to get TOP 4 for each cutoff
get_top4_for_cutoff <- function(full_dat, cutoff_val) {
  
  score_cols <- c("std_impact", "std_eff", "std_ppv", "std_missed", "std_delay")
  
  dat_with_min <- full_dat %>%
    rowwise() %>%
    mutate(min_score = min(c_across(all_of(score_cols)), na.rm = TRUE)) %>%
    ungroup()
  
  top4 <- dat_with_min %>%
    group_by(pop_brk) %>%
    arrange(desc(util_score)) %>%
    slice_head(n = 4) %>%
    ungroup() %>%
    mutate(is_top4 = TRUE)
  
  return(top4)
}

## loop over cutoff values to perform the sensitivity analysis
cutoff_values <- seq(from = -1, to = -0.5, by = 0.1)

top4_across_cutoffs <- list()

for (cut in cutoff_values) {
  top4_across_cutoffs[[as.character(cut)]] <- 
    get_top4_for_cutoff(full_dat, cutoff_val = cut) %>%
    mutate(cutoff = cut)
}

## combine all cutoffs
combined_top4 <- bind_rows(top4_across_cutoffs)

all_alerts <- full_dat %>%
  select(pop_brk, alert_lab, util_score,
         std_impact, std_eff, std_ppv, std_missed, std_delay)

## per-cutoff indicators
expanded <- all_alerts %>%
  crossing(cutoff = cutoff_values) %>%
  left_join(combined_top4 %>% 
              select(pop_brk, alert_lab, cutoff, is_top4),
            by = c("pop_brk", "alert_lab", "cutoff")) %>%
  mutate(is_top4 = ifelse(is.na(is_top4), FALSE, is_top4))

## failure reasons for each alert-cutoff
expanded <- expanded %>%
  rowwise() %>%
  mutate(fail_dims = ifelse(is_top4,
                            NA_character_,
                            get_fail_dims(cur_data(), cutoff))) %>%
  ungroup()

## 1 column per cutoff
wide <- expanded %>%
  mutate(label = ifelse(is_top4,
                        "top4",
                        ifelse(is.na(fail_dims), "not top4", fail_dims))) %>%
  select(pop_brk, alert_lab, cutoff, label) %>%
  pivot_wider(names_from = cutoff,
              values_from = label,
              names_prefix = "cutoff_")

## table for each population group
final_tables <- wide %>%
  arrange(pop_brk, alert_lab) %>%
  group_split(pop_brk) %>%
  setNames(unique(wide$pop_brk))

table_list <- list()

for (grp in names(final_tables)) {
  tab <- final_tables[[grp]] %>%
    rename(
      `population group` = pop_brk,
      `alert definition` = alert_lab
    )

  table_list[[grp]] <- tab %>%
    kbl(
      caption = paste0("Alert Definition Performance Across Cutoffs, Population Group: ", grp),
      align = "l"
    ) %>%
    kable_styling(full_width = FALSE)
}

### print table per population group
for (nm in names(table_list)) {
  cat("## ", nm, "\n\n")
  print(table_list[[nm]])
  cat("\n\n")
}

## saving as pngs to paste into manuscript google doc
library(webshot2)
library(htmltools)

output_dir <- "table_pngs"
dir.create(output_dir, showWarnings = FALSE)

for (nm in names(table_list)) {
  safe_nm <- gsub("[^A-Za-z0-9_-]+", "_", nm)
  html_file <- file.path(output_dir, paste0("table_", safe_nm, ".html"))
  png_file  <- file.path(output_dir, paste0("table_", safe_nm, ".png"))
  
  htmltools::save_html(
    htmltools::HTML(as.character(table_list[[nm]])),
    file = html_file
  )
  ## html to png
  webshot2::webshot(
    url = html_file,
    file = png_file,
    zoom = 2,        
    vwidth = 1200,   
    vheight = 800    
  )
}


```

### Supplement Backup

#### Complete Metrics Table

```{r}
#| label: tab-complete
#| echo: false
#|
reactable_template(def_complete_dat,
                  col_names = list(alert_type = reactable::colDef(name = "Alert Type"),
                                  pop_brk = reactable::colDef(name = "Pop Group"),
                                  alert_lab = reactable::colDef(name = "Alert Def"),
                                  util_score = reactable::colDef(name = "Util Score", defaultSortOrder = "desc")))
```

#### Raw Metrics and Total Score Table

```{r}
#| label: tab-raw-only
#| echo: false

reactable_template(dplyr::select(def_complete_dat, -contains("std_")),
                  col_names = list(alert_type = reactable::colDef(name = "Alert Type"),
                                  pop_brk = reactable::colDef(name = "Pop Group"),
                                  alert_lab = reactable::colDef(name = "Alert Def"),
                                  util_score = reactable::colDef(name = "Util Score", defaultSortOrder = "desc")))
```

#### Rank Metrics and Rank Score Table

```{r}
#| label: tab-sensitivity-only
#| echo: true
#|
reactable_template(def_complete_dat_sens,
                  col_names = list(alert_type = reactable::colDef(name = "Alert Type"),
                                  pop_brk = reactable::colDef(name = "Pop Group"),
                                  alert_lab = reactable::colDef(name = "Alert Def"),
                                  util_score_v2 = reactable::colDef(name = "Util Score", defaultSortOrder = "desc")))
```

#### Top 4 Alert Definitions by Rank Sum Score

```{r}
#| label: tab-top4-by-group-sens
#| echo: true
#| results: asis

def_top4_by_group_sens <- def_complete_dat_sens %>%
  group_by(pop_brk) %>%
  arrange(desc(util_score_v2)) %>%
  slice_head(n = 4) %>%
  ungroup() %>%
  select(pop_brk, alert_lab, alert_type, alert_stringency, 
         impact_mean, eff_mean, ppv, missed_prop, delay_mean, util_score_v2
  )  %>%
  left_join(
    def_complete_dat %>% select(pop_brk, alert_lab, util_score),
    by = c("pop_brk", "alert_lab")
  )

def_top4_by_group_sens <- def_top4_by_group_sens %>%
  mutate(
    pop_brk = fct_recode(
      pop_brk,
      "Administrative units with <50,000 people" = "< 50k",
      "Administrative units with 50,000 to 500,000 people" = "[50k, 500k)",
      "Administrative units with ≥500,000 people" = "≥ 500k"
    )
  ) %>%
  arrange(pop_brk, desc(util_score_v2))

group_counts <- def_top4_by_group_sens %>%
  count(pop_brk) %>%
  pull(n)

top_rows <- cumsum(c(1, head(group_counts, -1)))

# Rename columns for clarity
tab_for_kable <- def_top4_by_group_sens %>%
  select(
    alert_lab, alert_type, 
    
    impact_mean, eff_mean, ppv, missed_prop, delay_mean,
    util_score_v2, util_score
  ) %>%
  rename(
    "Alert Definition" = alert_lab,
    "Alert Type" = alert_type,
    "Impact: cases" = impact_mean,
    "Efficiency: per 1000 pop" = eff_mean,
    "PPV" = ppv,
    "Missed Proportion" = missed_prop,
    "Delay (weeks)" = delay_mean,
    "Utility Score (V2)" = util_score_v2,
    "Utility Score (Original)" = util_score
  )

## formatted table
table_kbl <- tab_for_kable %>%
  kbl(
    caption = "Top 4 Alerts by Utility Score V2 per Population Group",
    align = "lcccccccc"
  ) %>%
  kable_styling(
    full_width = TRUE,
    position = "center"
  )

start_row <- 1
for (i in seq_along(group_counts)) {
  end_row <- start_row + group_counts[i] - 1
  table_kbl <- table_kbl %>%
    group_rows(
      group_label = levels(def_top4_by_group_sens$pop_brk)[i],
      start_row = start_row,
      end_row = end_row
    )
  start_row <- end_row + 1
}

highlight_top_by_group <- def_top4_by_group_sens %>%
  mutate(
    highlight_flag = case_when(
      pop_brk == "Administrative units with <50,000 people" &
        alert_lab == "≥ 100 total" ~ TRUE,

      pop_brk == "Administrative units with 50,000 to 500,000 people" &
        alert_lab == "≥ 50 weekly" ~ TRUE,

      pop_brk == "Administrative units with ≥500,000 people" &
        alert_lab == "≥ 100 weekly" ~ TRUE,

      TRUE ~ FALSE
    )
  )

table_kbl <- table_kbl %>%
  row_spec(which(highlight_top_by_group$highlight_flag), background = "#d4edda")

table_kbl

```

#### Alert Definitions by Utility Score by Country

```{r}
#| label: tab-all-by-country
#| echo: true

reactable_template(dplyr::select(def_complete_country_dat, -contains("std_")),
                  col_names = list(alert_type = reactable::colDef(name = "Alert Type"),
                                  country = reactable::colDef(name = "Country"),
                                  alert_lab = reactable::colDef(name = "Alert Def"),
                                  util_score = reactable::colDef(name = "Util Score", defaultSortOrder = "desc")))

```


```{r}
#| label: fig-utility-by-country
#| echo: true

ggplot2::ggplot(def_complete_country_dat,
                ggplot2::aes(x = alert_lab, y = country, fill = util_score)) +
  ggplot2::geom_tile(color = "white") +
  ggplot2::scale_fill_viridis_c(option = "D") +
  ggplot2::labs(
    title = "Utility Score by Country and Alert Definition",
    x = "Alert Definition",
    y = "Country",
    fill = "Utility Score"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    axis.text.x = ggplot2::element_text(angle = 45, hjust = 1),
    axis.text.y = ggplot2::element_text(size = 8)
  )


```


####  Best Alert Definition by Utility Score by Country

```{r, out.width='100%'}
#| label: tab-top1-by-country
#| echo: true

score_cols <- c("std_impact", "std_eff", "std_ppv", "std_missed", "std_delay")
cutoff_val <- -1 ## selected cutoff value for std dimension scores based on sensitivity analysis above
def_top1_by_country <- def_complete_country_dat %>%
    rowwise() %>%
    mutate(min_score = min(c_across(all_of(score_cols)), na.rm = TRUE)) %>%
    ungroup() %>%
    filter(min_score >= cutoff_val) %>%
    group_by(country) %>%
    arrange(desc(util_score)) %>%
    slice_head(n = 1) %>%
    ungroup()


reactable_template(dplyr::select(def_top1_by_country, -contains("std_")),
                  col_names = list(alert_type = reactable::colDef(name = "Alert Type"),
                                  country = reactable::colDef(name = "Country"),
                                  alert_lab = reactable::colDef(name = "Alert Def"),
                                  util_score = reactable::colDef(name = "Util Score", defaultSortOrder = "desc")))

```

```{r, out.width='100%'}
#| label: top4-by-country
#| echo: true

score_cols <- c("std_impact", "std_eff", "std_ppv", "std_missed", "std_delay")
cutoff_val <- -1 ## selected cutoff value for std dimension scores based on sensitivity analysis 
def_top4_by_country <- def_complete_country_dat %>%
    filter(!is.na(util_score) & !is.nan(util_score)) %>%
    rowwise() %>%
    mutate(min_score = min(c_across(all_of(score_cols)), na.rm = TRUE)) %>%
    ungroup() %>%
    filter(min_score >= cutoff_val) %>%
    group_by(country) %>%
    arrange(desc(util_score)) %>%
    slice_head(n = 4) %>%
    ungroup()


reactable_template(dplyr::select(def_top4_by_country, -contains("std_")),
                  col_names = list(alert_type = reactable::colDef(name = "Alert Type"),
                                  country = reactable::colDef(name = "Country"),
                                  alert_lab = reactable::colDef(name = "Alert Def"),
                                  util_score = reactable::colDef(name = "Util Score", defaultSortOrder = "desc")))

alert_top4_freq <- def_top4_by_country %>%
  count(alert_lab, name = "n_top4") %>%
  arrange(desc(n_top4))

ggplot(alert_top4_freq, aes(x = reorder(alert_lab, n_top4), y = n_top4)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Frequency of Alert Definitions Appearing in Top 4 Across Countries",
    x = "Alert Definition",
    y = "Count of Top-4 Appearances"
  )

```

```{r}
#| label: fig-country-cv-by-alertdef
#| echo: true

## calculate coefficient of variation (CV) accross countries by alert definition
cv_by_alert <- def_complete_country_dat %>%
  group_by(alert_lab) %>%
  summarise(
    mean_util = mean(util_score, na.rm = TRUE),
    sd_util = sd(util_score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(cv_util = ifelse(mean_util != 0, sd_util / mean_util, NA_real_)) %>%
  filter(!is.na(cv_util))

## dot plot of cv
ggplot(cv_by_alert, aes(x = alert_lab, y = cv_util)) +
  geom_point(size = 3, color = "steelblue") +
  theme_minimal() +
  labs(
    title = "Coefficient of Variation of Utility Score by Alert Definition Across Countries",
    x = "Alert Definition",
    y = "Coefficient of Variation"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

## dot plot of sd
ggplot(cv_by_alert, aes(x = alert_lab, y = sd_util)) +
  geom_point(size = 3, color = "steelblue") +
  theme_minimal() +
  labs(
    title = "Standard Deviation of Utility Score by Alert Definition Across Countries",
    x = "Alert Definition",
    y = "Standard Deviation"
  ) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```


```{r}
#| label: fig4-multipanel
#| fig-height: 9
#| fig-width: 7
#| eval: false

library(tidytext)
library(ggspatial)

score_cols <- c("std_impact","std_eff","std_ppv","std_missed","std_delay")
cutoff_val <- -1

filtered_dat <- def_complete_country_dat %>%
  filter(!is.na(util_score) & !is.nan(util_score)) %>%
  rowwise() %>%
  mutate(min_score = min(c_across(all_of(score_cols)), na.rm = TRUE)) %>%
  ungroup() %>%
  filter(min_score >= cutoff_val)

plot_base_dat <- def_complete_country_dat %>%
  filter(!is.na(util_score) & !is.nan(util_score)) %>%
  rowwise() %>%
  mutate(
    min_score = min(c_across(all_of(score_cols)), na.rm = TRUE),
    passes_cutoff = min_score >= cutoff_val
  ) %>%
  ungroup()

plot_top_alerts <- function(data, n = 3, color_var = c("alert_type","region")) {

  color_var <- match.arg(color_var)

  africa_region_map <- tribble(
    ~iso3, ~region,
    "BEN","Western Africa","BFA","Western Africa","CPV","Western Africa",
    "CIV","Western Africa","GMB","Western Africa","GHA","Western Africa",
    "GIN","Western Africa","GNB","Western Africa","LBR","Western Africa",
    "MLI","Western Africa","MRT","Western Africa","NER","Western Africa",
    "NGA","Western Africa","SEN","Western Africa","SLE","Western Africa",
    "TGO","Western Africa",
    "AGO","Central Africa","CAF","Central Africa","CMR","Central Africa",
    "COD","Central Africa","COG","Central Africa","GNQ","Central Africa",
    "GAB","Central Africa","STP","Central Africa","TCD","Central Africa",
    "BDI","Eastern Africa","COM","Eastern Africa","DJI","Eastern Africa",
    "ERI","Eastern Africa","ETH","Eastern Africa","KEN","Eastern Africa",
    "MDG","Eastern Africa","RWA","Eastern Africa","SOM","Eastern Africa",
    "SSD","Eastern Africa","TZA","Eastern Africa","UGA","Eastern Africa",
    "SDN","Eastern Africa",
    "BWA","Southern Africa","LSO","Southern Africa","NAM","Southern Africa",
    "ZAF","Southern Africa","SWZ","Southern Africa","MWI","Southern Africa",
    "MOZ","Southern Africa","ZWE","Southern Africa","ZMB","Southern Africa"
  )

  data <- data %>%
    rename(iso3 = country) %>%
    left_join(africa_region_map, by = "iso3") %>%
    rename(country = iso3) %>%
    mutate(region = gsub(" ", "\n", region))

  top_n <- data %>%
    filter(passes_cutoff) %>%
    group_by(country) %>%
    arrange(desc(util_score)) %>%
    mutate(rank_within_country = row_number()) %>%
    slice_head(n = n) %>%
    ungroup() %>%
    mutate(is_top = TRUE) %>%
    select(country, alert_lab, is_top, rank_within_country)

  plot_dat <- data %>%
  left_join(top_n, by = c("country","alert_lab")) %>%
  mutate(
    size_rank = ifelse(is_top, rank_within_country, NA_real_)  # NEW
  )

  region_colors <- c(
    "Western\nAfrica"  = "#F4A676",
    "Central\nAfrica"  = "#A7B545",
    "Eastern\nAfrica"  = "#7C6A58",
    "Southern\nAfrica" = "#003F50"
  )

  alert_type_colors <- c(
    "trend"    = "#e74c3c",
    "case"     = "#f1c40f",
    "cum case" = "#1abc9c",
    "rate"     = "#9b59b6"
  )

  color_scale <- if (color_var == "alert_type") {
    scale_color_manual(values = alert_type_colors)
  } else {
    scale_color_manual(values = region_colors)
  }

  ggplot(plot_dat, aes(x = country, y = alert_lab)) +
    geom_point(
      aes(size = size_rank, color = !!sym(color_var)),
      na.rm = TRUE, alpha = 0.85
    ) +
    facet_grid(alert_type ~ region, scales = "free", space = "free") +
    scale_y_discrete(limits = function(x) unique(x)) +
    scale_x_discrete(limits = function(x) unique(x)) +
    color_scale +
    scale_size(
      range = c(6, 1.5),       
      limits = c(n, 1),        ## reverse bubble size, higher rank is larger bubble
      breaks = 1:n,
      labels = paste("Rank", 1:n)
    ) +
    theme_minimal() +
    labs(
      x = "Country",
      y = "Alert Definition",
      size = "Rank (1 = highest utility)",  
      color = ifelse(color_var == "region", "Region", "Alert Type")
  ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(face = "bold", size = 11),
      legend.position = "none"
    )
}

p_alerts <- plot_top_alerts(plot_base_dat, n = 3, color_var = "alert_type")

## region mini map 
colors_afro_regions <- c("#FFA378", "#A8B545", "#8C796D", "#024554")
names(colors_afro_regions) <- c(
  "Western Africa","Central Africa","Eastern Africa","Southern Africa"
)

africa_region_map <- tribble(
  ~iso3, ~region,
  "BEN","Western Africa","BFA","Western Africa","CPV","Western Africa",
  "CIV","Western Africa","GMB","Western Africa","GHA","Western Africa",
  "GIN","Western Africa","GNB","Western Africa","LBR","Western Africa",
  "MLI","Western Africa","MRT","Western Africa","NER","Western Africa",
  "NGA","Western Africa","SEN","Western Africa","SLE","Western Africa",
  "TGO","Western Africa",
  "AGO","Central Africa","CAF","Central Africa","CMR","Central Africa",
  "COD","Central Africa","COG","Central Africa","GNQ","Central Africa",
  "GAB","Central Africa","STP","Central Africa","TCD","Central Africa",
  "BDI","Eastern Africa","COM","Eastern Africa","DJI","Eastern Africa",
  "ERI","Eastern Africa","ETH","Eastern Africa","KEN","Eastern Africa",
  "MDG","Eastern Africa","RWA","Eastern Africa","SOM","Eastern Africa",
  "SSD","Eastern Africa","TZA","Eastern Africa","UGA","Eastern Africa",
  "SDN","Eastern Africa",
  "BWA","Southern Africa","LSO","Southern Africa","NAM","Southern Africa",
  "ZAF","Southern Africa","SWZ","Southern Africa","MWI","Southern Africa",
  "MOZ","Southern Africa","ZWE","Southern Africa","ZMB","Southern Africa"
)

africa_sf <- gisco_get_countries(region = "Africa", resolution = "3") %>%
  left_join(africa_region_map, by = c("ISO3_CODE"="iso3")) %>%
  filter(!is.na(region))

p_map <- ggplot(africa_sf) +
  geom_sf(aes(fill = region), color = "white", linewidth = 0.2) +
  scale_fill_manual(values = colors_afro_regions) +
  coord_sf(xlim = c(-20, 55), ylim = c(-35, 38), expand = FALSE) +
  theme_minimal() +
  theme(panel.grid = element_blank(), axis.text = element_blank(),
        axis.title = element_blank(), legend.position = "right")

africa_sf_t1 <- gisco_get_countries(
  region = "Africa",
  resolution = "3"
) %>%
  filter(ISO3_CODE %in% def_top1_by_country$country)

def_top1_valid <- def_top1_by_country %>%
  filter(!is.na(util_score) & !is.nan(util_score))

africa_alerts_sf <- africa_sf_t1 %>%
  left_join(def_top1_valid, by = c("ISO3_CODE" = "country")) %>%
  mutate(alert_stringency = factor(alert_stringency, ordered = TRUE))

alert_type_colors <- c(
  "trend"    = "#e74c3c",
  "case"     = "#f1c40f",
  "cum case" = "#1abc9c",
  "rate"     = "#9b59b6"
)

p_alert_map <- ggplot(africa_alerts_sf) +
  geom_sf(aes(fill = alert_type),
          color = "grey40", linewidth = 0.2) +   # removed alpha
  geom_sf_text(aes(label = ISO3_CODE), size = 2.5, color = "gray20") +
  scale_fill_manual(values = alert_type_colors, na.value = "lightgray", name = "Alert Type") +
  annotation_north_arrow(
    location = "tr",                 
    which_north = "true",
    height = unit(1, "cm"),
    width = unit(1, "cm"),
    style = north_arrow_fancy_orienteering()
  ) +
  annotation_scale(
    location = "bl",                
    width_hint = 0.25,
    text_cex = 0.8,                  
    bar_cols = c("grey60","white")   
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.key.height = unit(0.3, "cm"),
    legend.key.width = unit(0.3, "cm"),
    legend.spacing.y = unit(0.05, "cm"),
    legend.text = element_text(size = 7),
    legend.title = element_text(size = 8),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank()
  )


plot_alert_frequency_by_region <- function(data, n = 3) {

  africa_region_map <- africa_region_map

  data <- data %>%
    rename(iso3 = country) %>%
    left_join(africa_region_map, by = "iso3") %>%
    rename(country = iso3)

  top_n <- data %>%
    group_by(country) %>%
    arrange(desc(util_score)) %>%
    slice_head(n = n) %>%
    ungroup()

  alert_freq <- top_n %>%
    count(alert_type, alert_lab, region, name = "count") %>%
    group_by(alert_type, alert_lab) %>%
    mutate(total = sum(count)) %>%
    ungroup() %>%
    mutate(alert_lab_ord = reorder_within(alert_lab, total, alert_type))

  region_colors <- c(
    "Western Africa"  = "#F4A676",
    "Central Africa"  = "#A7B545",
    "Eastern Africa"  = "#7C6A58",
    "Southern Africa" = "#003F50"
  )

  ggplot(alert_freq, aes(x = alert_lab_ord, y = count, fill = region)) +
    geom_col() +
    coord_flip() +
    facet_wrap(~ alert_type, scales = "free_y", ncol = 1) +
    scale_x_reordered() +
    scale_fill_manual(values = region_colors) +
    theme_minimal() +
    labs(
      x = "",
      y = "Count",
      fill = "Region"
    ) +
    theme(
      strip.text = element_text(face = "bold"),
      legend.position = "none"  
    )
}

p_freq <- plot_alert_frequency_by_region(filtered_dat, n = 3)

## top panel
p_top <- plot_grid(
  p_alerts, p_map,
  ncol = 2,
  rel_widths = c(0.7, 0.3),
  labels = c("A", "B"),
  label_size = 14
)

## bottom panel
p_bottom <- plot_grid(
  p_freq,
  p_alert_map,
  ncol = 2,
  rel_widths = c(1, 1),
  labels = c("C", "D"),
  label_size = 14
)

p_map_nl <- p_map + theme(legend.position = "none")

p_C_with_B <- ggdraw() +
  draw_plot(p_freq, 0, 0, 1, 1) +   
draw_plot(p_map_nl,
          x = 0.75, y = 0.75,
          width = 0.28, height = 0.28)

p_top_row <- plot_grid(
  p_alert_map,
  p_C_with_B,
  ncol = 2,
  rel_widths = c(0.45, 0.55),
  labels = c("A","B"),
  label_size = 14
)

p_bottom_row <- plot_grid(
  p_alerts,         # A
  ncol = 1,
  labels = "C",
  label_size = 14
)

full_figure <- plot_grid(
  p_top_row,
  p_bottom_row,
  ncol = 1,
  rel_heights = c(1, 1.1)
)

ggsave(
  paste0(fig_path,"/combined_reordered_figure.png"),
  full_figure,
  width = 13,
  height = 11,
  dpi = 300
)

print(full_figure)

```


```{r}
#| label: fig4-singlepanel-version2

score_cols <- c("std_impact","std_eff","std_ppv","std_missed","std_delay")
cutoff_val <- -1

plot_base_dat <- def_complete_country_dat %>%
  filter(!is.na(util_score) & !is.nan(util_score)) %>%
  rowwise() %>%
  mutate(
    min_score = min(c_across(all_of(score_cols)), na.rm = TRUE),
    passes_cutoff = min_score >= cutoff_val
  ) %>%
  ungroup()

africa_region_map <- tribble(
  ~iso3, ~region,
  "BEN","Western Africa","BFA","Western Africa","CPV","Western Africa",
  "CIV","Western Africa","GMB","Western Africa","GHA","Western Africa",
  "GIN","Western Africa","GNB","Western Africa","LBR","Western Africa",
  "MLI","Western Africa","MRT","Western Africa","NER","Western Africa",
  "NGA","Western Africa","SEN","Western Africa","SLE","Western Africa",
  "TGO","Western Africa",
  "AGO","Central Africa","CAF","Central Africa","CMR","Central Africa",
  "COD","Central Africa","COG","Central Africa","GNQ","Central Africa",
  "GAB","Central Africa","STP","Central Africa","TCD","Central Africa",
  "BDI","Eastern Africa","COM","Eastern Africa","DJI","Eastern Africa",
  "ERI","Eastern Africa","ETH","Eastern Africa","KEN","Eastern Africa",
  "MDG","Eastern Africa","RWA","Eastern Africa","SOM","Eastern Africa",
  "SSD","Eastern Africa","TZA","Eastern Africa","UGA","Eastern Africa",
  "SDN","Eastern Africa",
  "BWA","Southern Africa","LSO","Southern Africa","NAM","Southern Africa",
  "ZAF","Southern Africa","SWZ","Southern Africa","MWI","Southern Africa",
  "MOZ","Southern Africa","ZWE","Southern Africa","ZMB","Southern Africa"
)

plot_top_alerts <- function(data, n = 3, color_var = c("region","alert_type")) {

  color_var <- match.arg(color_var)

  data <- data %>%
    rename(iso3 = country) %>%
    left_join(africa_region_map, by = "iso3") %>%
    rename(country = iso3) %>%
    mutate(region = gsub(" ", "\n", region))

  top_n <- data %>%
    filter(passes_cutoff) %>%
    group_by(country) %>%
    arrange(desc(util_score)) %>%
    mutate(rank_within_country = row_number()) %>%
    slice_head(n = n) %>%
    ungroup() %>%
    mutate(is_top = TRUE) %>%
    select(country, alert_lab, is_top, rank_within_country)

  plot_dat <- data %>%
    left_join(top_n, by = c("country","alert_lab")) %>%
    mutate(size_rank = ifelse(is_top, rank_within_country, NA_real_))

  region_colors <- c(
    "Western\nAfrica"  = "#F4A676",
    "Central\nAfrica"  = "#A7B545",
    "Eastern\nAfrica"  = "#7C6A58",
    "Southern\nAfrica" = "#003F50"
  )

  alert_type_colors <- c(
    "trend"    = "#e74c3c",
    "case"     = "#f1c40f",
    "cum case" = "#1abc9c",
    "rate"     = "#9b59b6"
  )

  color_scale <- if (color_var == "region") {
    scale_color_manual(values = region_colors)
  } else {
    scale_color_manual(values = alert_type_colors)
  }

  ggplot(plot_dat, aes(x = country, y = alert_lab)) +
    geom_point(
      aes(size = size_rank, color = .data[[color_var]]),
      na.rm = TRUE,
      alpha = 0.85
    ) +
    facet_grid(alert_type ~ region, scales = "free", space = "free") +
    color_scale +
    scale_size(
      range = c(6, 1.5),
      limits = c(n, 1),
      breaks = 1:n,
      labels = paste("Rank", 1:n)
    ) +
    theme_minimal() +
    labs(
      x = "Country",
      y = "Alert Definition",
      size = "Rank (1 = highest utility)",
      color = ifelse(color_var == "region","Region","Alert Type")
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(face = "bold", size = 11),
      legend.position = "none"
    )
}

p_alerts <- plot_top_alerts(plot_base_dat, n = 3, color_var = "region")

colors_afro_regions <- c(
  "Western Africa"  = "#F4A676",
  "Central Africa"  = "#A7B545",
  "Eastern Africa"  = "#7C6A58",
  "Southern Africa" = "#003F50"
)

africa_sf <- gisco_get_countries(region = "Africa", resolution = "3") %>%
  left_join(africa_region_map, by = c("ISO3_CODE"="iso3")) %>%
  filter(!is.na(region))

p_map <- ggplot(africa_sf) +
  geom_sf(aes(fill = region), color = "white", linewidth = 0.2) +
  scale_fill_manual(values = colors_afro_regions) +
  coord_sf(xlim = c(-20, 55), ylim = c(-35, 38), expand = FALSE) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.title = element_blank(),
    legend.position = "none",
    plot.margin = margin(10, 10, 10, 0),
    panel.border = element_rect(color = "grey80", fill = NA)
  )

p_final <- plot_grid(
  p_alerts,
  p_map,
  ncol = 2,
  rel_widths = c(0.86, 0.14)
)

ggsave(
  paste0(fig_path, "/bubble_alert_plot_",ocv_status,".png"),
  p_final,
  width = 12,
  height = 8,
  dpi = 300
)

print(p_final)

```

```{r}
#| label: fig4-singlepanel-coauthor

score_cols <- c("std_impact","std_eff","std_ppv","std_missed","std_delay")
cutoff_val <- -1

plot_base_dat <- def_complete_country_dat %>%
  filter(!is.na(util_score) & !is.nan(util_score)) %>%
  rowwise() %>%
  mutate(
    min_score = min(c_across(all_of(score_cols)), na.rm = TRUE),
    passes_cutoff = min_score >= cutoff_val
  ) %>%
  ungroup()

africa_region_map <- tribble(
  ~iso3, ~region,
  "BEN","Western Africa","BFA","Western Africa","CPV","Western Africa",
  "CIV","Western Africa","GMB","Western Africa","GHA","Western Africa",
  "GIN","Western Africa","GNB","Western Africa","LBR","Western Africa",
  "MLI","Western Africa","MRT","Western Africa","NER","Western Africa",
  "NGA","Western Africa","SEN","Western Africa","SLE","Western Africa",
  "TGO","Western Africa",
  "AGO","Central Africa","CAF","Central Africa","CMR","Central Africa",
  "COD","Central Africa","COG","Central Africa","GNQ","Central Africa",
  "GAB","Central Africa","STP","Central Africa","TCD","Central Africa",
  "BDI","Eastern Africa","COM","Eastern Africa","DJI","Eastern Africa",
  "ERI","Eastern Africa","ETH","Eastern Africa","KEN","Eastern Africa",
  "MDG","Eastern Africa","RWA","Eastern Africa","SOM","Eastern Africa",
  "SSD","Eastern Africa","TZA","Eastern Africa","UGA","Eastern Africa",
  "SDN","Eastern Africa",
  "BWA","Southern Africa","LSO","Southern Africa","NAM","Southern Africa",
  "ZAF","Southern Africa","SWZ","Southern Africa","MWI","Southern Africa",
  "MOZ","Southern Africa","ZWE","Southern Africa","ZMB","Southern Africa"
)

plot_top_alerts <- function(data, n = 3, color_var = c("region", "alert_type")) {

  color_var <- match.arg(color_var)

  data <- data %>%
    rename(iso3 = country) %>%
    left_join(africa_region_map, by = "iso3") %>%
    rename(country = iso3) %>%
    mutate(region = gsub(" ", "\n", region))

  top_n <- data %>%
    filter(passes_cutoff) %>%
    group_by(country) %>%
    arrange(desc(util_score)) %>%
    mutate(rank_within_country = row_number()) %>%
    slice_head(n = n) %>%
    ungroup() %>%
    mutate(is_top = TRUE) %>%
    select(country, alert_lab, is_top, rank_within_country)

  plot_dat <- data %>%
    left_join(top_n, by = c("country", "alert_lab")) %>%
    mutate(size_rank = ifelse(is_top, rank_within_country, NA_real_))

  region_colors <- c(
    "Western\nAfrica"  = "#F4A676",
    "Central\nAfrica"  = "#A7B545",
    "Eastern\nAfrica"  = "#7C6A58",
    "Southern\nAfrica" = "#003F50"
  )

  alert_type_colors <- c(
    "trend"    = "#e74c3c",
    "case"     = "#f1c40f",
    "cum case" = "#1abc9c",
    "rate"     = "#9b59b6"
  )

  color_scale <- if (color_var == "region") {
    scale_color_manual(values = region_colors)
  } else {
    scale_color_manual(values = alert_type_colors)
  }

  ggplot(plot_dat, aes(x = country, y = alert_lab)) +
    geom_point(
      aes(size = size_rank, color = .data[[color_var]]),
      na.rm = TRUE,
      alpha = 0.85
    ) +
    facet_grid(alert_type ~ region, scales = "free", space = "free") +
    color_scale +
    scale_size(
      range  = c(6, 1.5),
      limits = c(n, 1),
      breaks = 1:n,
      labels = paste("Rank", 1:n)
    ) +
    theme_minimal() +
    labs(
      x     = "Country",
      y     = "Alert Definition",
      size  = "Rank (1 = highest utility)",
      color = ifelse(color_var == "region", "Region", "Alert Type")
    ) +
    theme(
      axis.text.x     = element_text(angle = 45, hjust = 1),
      strip.text      = element_text(face = "bold", size = 11),
      legend.position = "none"
    )
}

p_alerts <- plot_top_alerts(plot_base_dat, n = 3, color_var = "region")

## Histogram
hist_dat <- plot_base_dat %>%
  rename(iso3 = country) %>%
  left_join(africa_region_map, by = "iso3") %>%
  rename(country = iso3) %>%
  filter(passes_cutoff) %>%
  group_by(country) %>%
  arrange(desc(util_score)) %>%
  mutate(rank_within_country = row_number()) %>%
  slice_head(n = 3) %>%
  ungroup() %>%
  count(alert_lab, alert_type, name = "n_top") %>%
  right_join(
    plot_base_dat %>% distinct(alert_lab, alert_type),
    by = c("alert_lab", "alert_type")
  ) %>%
  mutate(n_top = replace_na(n_top, 0))

p_hist <- ggplot(hist_dat, aes(x = n_top, y = alert_lab)) +
  geom_col(fill = "grey60", width = 0.6) +
  facet_grid(rows = vars(alert_type), scales = "free_y", space = "free_y") +
  scale_x_continuous(
    breaks = scales::breaks_width(2),
    expand = expansion(mult = c(0, 0.1))
  ) +
  theme_minimal() +
  labs(x = "No. countries\n(top 3)", y = NULL) +
  theme(
    axis.text.y        = element_blank(),
    axis.ticks.y       = element_blank(),
    strip.text.y       = element_blank(),
    strip.text.x       = element_blank(),
    panel.grid.major.y = element_blank(),
    legend.position    = "none"
  )

## Map
colors_afro_regions <- c(
  "Western Africa"  = "#F4A676",
  "Central Africa"  = "#A7B545",
  "Eastern Africa"  = "#7C6A58",
  "Southern Africa" = "#003F50"
)

africa_sf <- gisco_get_countries(region = "Africa", resolution = "3") %>%
  left_join(africa_region_map, by = c("ISO3_CODE" = "iso3")) %>%
  filter(!is.na(region))

p_map <- ggplot(africa_sf) +
  geom_sf(aes(fill = region), color = "white", linewidth = 0.2) +
  scale_fill_manual(values = colors_afro_regions) +
  coord_sf(xlim = c(-20, 55), ylim = c(-35, 38), expand = FALSE) +
  theme_minimal() +
  theme(
    panel.grid    = element_blank(),
    axis.text     = element_blank(),
    axis.title    = element_blank(),
    legend.position = "none",
    plot.margin   = margin(10, 10, 10, 0),
    panel.border  = element_rect(color = "grey80", fill = NA)
  )

## Size legend — extracted from a temporary version of p_alerts
p_legend <- cowplot::get_legend(
  plot_top_alerts(plot_base_dat, n = 3, color_var = "region") +
    ## show only the size legend, suppress the colour legend
    guides(color = "none") +
    theme(
      legend.position     = "right",
      legend.title        = element_text(size = 9),
      legend.text         = element_text(size = 8),
      legend.key.size     = unit(0.5, "cm"),
      legend.background   = element_blank()
    )
)

## Right column: map stacked above size legend
p_right <- plot_grid(
  p_map,
  p_legend,
  ncol        = 1,
  rel_heights = c(0.55, 0.45)
)

p_final <- plot_grid(
  p_alerts,
  p_hist,
  p_right,
  ncol       = 3,
  rel_widths = c(0.76, 0.10, 0.14),
  align      = "h",
  axis       = "tb"
)

ggsave(
  paste0(fig_path, "/bubble_alert_plot_", ocv_status, ".png"),
  p_final,
  width  = 12,
  height = 8,
  dpi    = 300
)

print(p_final)

```

```{r}
#| label: fig4-singlepanel

score_cols <- c("std_impact","std_eff","std_ppv","std_missed","std_delay")
cutoff_val <- -1

plot_base_dat <- def_complete_country_dat %>%
  filter(!is.na(util_score) & !is.nan(util_score)) %>%
  rowwise() %>%
  mutate(
    min_score = min(c_across(all_of(score_cols)), na.rm = TRUE),
    passes_cutoff = min_score >= cutoff_val
  ) %>%
  ungroup()

africa_region_map <- tribble(
  ~iso3, ~region,
  "BEN","Western Africa","BFA","Western Africa","CPV","Western Africa",
  "CIV","Western Africa","GMB","Western Africa","GHA","Western Africa",
  "GIN","Western Africa","GNB","Western Africa","LBR","Western Africa",
  "MLI","Western Africa","MRT","Western Africa","NER","Western Africa",
  "NGA","Western Africa","SEN","Western Africa","SLE","Western Africa",
  "TGO","Western Africa",
  "AGO","Central Africa","CAF","Central Africa","CMR","Central Africa",
  "COD","Central Africa","COG","Central Africa","GNQ","Central Africa",
  "GAB","Central Africa","STP","Central Africa","TCD","Central Africa",
  "BDI","Eastern Africa","COM","Eastern Africa","DJI","Eastern Africa",
  "ERI","Eastern Africa","ETH","Eastern Africa","KEN","Eastern Africa",
  "MDG","Eastern Africa","RWA","Eastern Africa","SOM","Eastern Africa",
  "SSD","Eastern Africa","TZA","Eastern Africa","UGA","Eastern Africa",
  "SDN","Eastern Africa",
  "BWA","Southern Africa","LSO","Southern Africa","NAM","Southern Africa",
  "ZAF","Southern Africa","SWZ","Southern Africa","MWI","Southern Africa",
  "MOZ","Southern Africa","ZWE","Southern Africa","ZMB","Southern Africa"
)

plot_top_alerts <- function(data, n = 3) {

  data <- data %>%
    rename(iso3 = country) %>%
    left_join(africa_region_map, by = "iso3") %>%
    rename(country = iso3) %>%
    mutate(region = gsub(" ", "\n", region))

  top_n <- data %>%
    filter(passes_cutoff) %>%
    group_by(country) %>%
    arrange(desc(util_score)) %>%
    mutate(rank_within_country = row_number()) %>%
    slice_head(n = n) %>%
    ungroup() %>%
    mutate(is_top = TRUE) %>%
    select(country, alert_lab, is_top, rank_within_country)

  plot_dat <- data %>%
    left_join(top_n, by = c("country","alert_lab")) %>%
    mutate(size_rank = ifelse(is_top, rank_within_country, NA_real_))

  alert_type_colors <- c(
    "trend"    = "#e74c3c",
    "case"     = "#f1c40f",
    "cum case" = "#1abc9c",
    "rate"     = "#9b59b6"
  )

  ggplot(plot_dat, aes(x = country, y = alert_lab)) +
    geom_point(
      aes(size = size_rank, color = alert_type),
      na.rm = TRUE,
      alpha = 0.85
    ) +
    facet_grid(alert_type ~ region, scales = "free", space = "free") +
    scale_color_manual(values = alert_type_colors) +
    scale_size(
      range = c(6, 1.5),
      limits = c(n, 1),
      breaks = 1:n,
      labels = paste("Rank", 1:n)
    ) +
    theme_minimal() +
    labs(
      x = "Country",
      y = "Alert Definition",
      size = "Rank (1 = highest utility)",
      color = "Alert Type"
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1),
      strip.text = element_text(face = "bold", size = 11),
      legend.position = "none"
    )
}

p_alerts <- plot_top_alerts(plot_base_dat, n = 3)

ggsave(
  paste0(fig_path, "/bubble_alert_plot.png"),
  p_alerts,
  width = 12,
  height = 8,
  dpi = 300
)

print(p_alerts)

```



```{r}
#| label: country-strat-utility-other-popgroups
prep_country_utility_data <- function(pop_group){

means_impact_eff_country_dat <- prep_impact_eff_dat %>%
  dplyr::filter(pop_brk == pop_group) %>%
  dplyr::group_by(alert_number, alert_type, country) %>%
  dplyr::summarise(
    impact_mean = mean(impact),
    eff_mean = mean(eff),
    .groups = "drop"
  )

means_delay_country_dat <- prep_missed_delay_dat %>%
  dplyr::filter(pop_brk == pop_group) %>%
  dplyr::group_by(alert_number, alert_type, country) %>%
  dplyr::summarise(
    delay_mean = mean(delay, na.rm = TRUE),
    .groups = "drop"
  )

n_country_outbreaks <- linked_cl_large %>%
  dplyr::filter(pop_brk == pop_group) %>%
  dplyr::distinct(outbreak_UID, country) %>%
  dplyr::group_by(country) %>%
  dplyr::summarise(n_outbreaks = n(), .groups = "drop")

fig_impact_country_dat <- dplyr::group_by(
  prep_impact_eff_dat %>% dplyr::filter(pop_brk == pop_group),
  alert_number, alert_type, country
) %>%
  dplyr::reframe(quibble2(impact)) %>%
  tidyr::pivot_wider(
    names_from = impact_q,
    names_glue = "{.value}_{impact_q}",
    values_from = impact
  ) %>%
  dplyr::full_join(
    dplyr::select(means_impact_eff_country_dat,
                  alert_number, alert_type, country, impact_mean),
    by = c("alert_number","alert_type","country")
  ) %>%
  prepare_fig_dat()

fig_eff_country_dat <- dplyr::group_by(
  prep_impact_eff_dat %>% dplyr::filter(pop_brk == pop_group),
  alert_number, alert_type, country
) %>%
  dplyr::reframe(quibble2(eff)) %>%
  tidyr::pivot_wider(
    names_from = eff_q,
    names_glue = "{.value}_{eff_q}",
    values_from = eff
  ) %>%
  dplyr::full_join(
    dplyr::select(means_impact_eff_country_dat,
                  alert_number, alert_type, country, eff_mean),
    by = c("alert_number","alert_type","country")
  ) %>%
  prepare_fig_dat()

fig_ppv_country_dat <- dplyr::group_by(
  prep_ppv_dat %>% dplyr::filter(pop_brk == pop_group),
  alert_number, alert_type, country
) %>%
  dplyr::summarise(
    alerts = n(),
    true_positives = sum(true_positives),
    ppv = round(true_positives/alerts, 2),
    .groups = "drop"
  ) %>%
  prepare_fig_dat()

fig_missed_country_dat <- dplyr::group_by(
  prep_missed_delay_dat %>% dplyr::filter(pop_brk == pop_group),
  alert_number, alert_type, country
) %>%
  dplyr::summarise(
    n_detected = sum(is_triggered),
    .groups = "drop"
  ) %>%
  dplyr::left_join(n_country_outbreaks, by = "country") %>%
  dplyr::mutate(
    n_missed = n_outbreaks - n_detected,
    missed_prop = n_missed / n_outbreaks
  ) %>%
  prepare_fig_dat()

fig_delay_country_dat <- dplyr::group_by(
  prep_missed_delay_dat %>% dplyr::filter(pop_brk == pop_group),
  alert_number, alert_type, country
) %>%
  dplyr::reframe(quibble2(delay)) %>%
  dplyr::ungroup() %>%
  tidyr::pivot_wider(
    names_from = delay_q,
    names_glue = "{.value}_{delay_q}",
    values_from = delay
  ) %>%
  dplyr::full_join(
    means_delay_country_dat,
    by = c("alert_number","alert_type","country")
  ) %>%
  prepare_fig_dat()

fig_complete_country_dat <- dplyr::full_join(
  fig_impact_country_dat, fig_eff_country_dat,
  by = c("alert_number","alert_type","country","alert_lab")
) %>%
  dplyr::full_join(fig_ppv_country_dat,
                   by = c("alert_number","alert_type","country","alert_lab")) %>%
  dplyr::full_join(fig_missed_country_dat,
                   by = c("alert_number","alert_type","country","alert_lab")) %>%
  dplyr::full_join(fig_delay_country_dat,
                   by = c("alert_number","alert_type","country","alert_lab"))

def_complete_country_dat <- fig_complete_country_dat %>%
  dplyr::group_by(country) %>%
  dplyr::mutate(
    std_impact = as.vector(scale(impact_mean)),
    std_eff = as.vector(scale(eff_mean)),
    std_ppv = as.vector(scale(ppv)),
    std_missed = as.vector(scale(missed_prop)) * -1,
    std_delay = as.vector(scale(delay_mean)) * -1
  ) %>%
  dplyr::ungroup() %>%
  OutbreakExtractR::add_alert_stringency() %>%
  dplyr::select(
    country, alert_lab, alert_type, alert_stringency,
    impact_mean, eff_mean, ppv, missed_prop, delay_mean,
    std_impact, std_eff, std_ppv, std_missed, std_delay
  ) %>%
  dplyr::mutate(
    dplyr::across(impact_mean:delay_mean, ~round(.x,1)),
    dplyr::across(dplyr::contains("std_"), ~round(.x,1))
  ) %>%
  dplyr::mutate(
    util_score = rowSums(dplyr::across(dplyr::contains("std_")))
  ) %>%
  dplyr::arrange(desc(util_score))

  ## save to file to generate manuscript figures in dedicated script
  clean_pop_group <- gsub("[^A-Za-z0-9]", "_", pop_group)
  readr::write_csv(def_complete_country_dat, paste0("country_utility_", clean_pop_group, ".csv"))

return(def_complete_country_dat)

}

## use function on all pop groups
pop_groups <- unique(prep_impact_eff_dat$pop_brk)

country_pop_results <- purrr::map(pop_groups, prep_country_utility_data)

names(country_pop_results) <- pop_groups

## get top x functions
get_top_alerts_by_country <- function(dat, n_top = 1, cutoff_val = -1){

score_cols <- c("std_impact","std_eff","std_ppv","std_missed","std_delay")

dat %>%
  dplyr::filter(!is.na(util_score) & !is.nan(util_score)) %>%
  dplyr::rowwise() %>%
  dplyr::mutate(
    min_score = min(dplyr::c_across(dplyr::all_of(score_cols)), na.rm = TRUE)
  ) %>%
  dplyr::ungroup() %>%
  dplyr::filter(min_score >= cutoff_val) %>%
  dplyr::group_by(country) %>%
  dplyr::arrange(desc(util_score)) %>%
  dplyr::slice_head(n = n_top) %>%
  dplyr::ungroup()

}

top3_results <- purrr::map(
  country_pop_results,
  ~ get_top_alerts_by_country(.x, n_top = 3)
)

top3_freq_by_pop <- purrr::imap_dfr(
  top3_results,           # list of Top-1 results per pop group
  ~ .x %>%
      count(alert_lab, name = "n_countries") %>%  ## count how many countries
      mutate(pop_brk = .y)                        ## add population group
)

top3_freq_by_pop <- top3_freq_by_pop %>%
  group_by(pop_brk) %>%
  arrange(desc(n_countries), .by_group = TRUE) %>%
  ungroup

readr::write_csv(top3_freq_by_pop, paste0(fig_path,"/country_utility_top3_ncountries","_", ocv_status,".csv"))

ggplot(top3_freq_by_pop,
       aes(x = reorder(alert_lab, n_countries), y = n_countries, fill = pop_brk)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~pop_brk, scales = "free_y") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Number of countries each alert Was Top-1 by population group",
    x = "Alert Definition",
    y = "Number of Countries"
  )

```

# Supplement Methods - General

## Unique population group distribution by admin scale

filter_ms_data is applied. Locations will only be double-counted if they switch population groups across years

```{r}
#| label: fig-pop-distr-supp
#| fig.width: 4.75
#| fig.height: 3.5
#| echo: true

## total subnational locations in pre_extract
total_subnat_locations <- pre_extract %>%
  dplyr::filter(stringr::str_count(location, "::") > 1) %>%
  dplyr::distinct(location) %>%
  nrow()

## collapse time dimension but allow population changes
outbreak_locs <- pre_extract %>%
  filter_ms_data(which_setting = which_setting) %>% ## filter out country and endemic locations
  dplyr::distinct(location, pop_brk, spatial_scale)

## subnational locations that appear in your final analysis dataset (fig denominator)
included_subnat_locations <- outbreak_locs %>%
  dplyr::distinct(location) %>%
  nrow()

## proportion
pct_included_subnat <- round(
  100 * included_subnat_locations / total_subnat_locations,
  1
)

coverage_text <- glue::glue(
  "In total, {total_subnat_locations} subnational locations were available. ",
  "Of these, {included_subnat_locations} ({pct_included_subnat}%) ",
  "are included in this analysis."
)

coverage_text

## population groups

## collapse time dimension but allow population changes
 pre_pops <- pre_extract %>%
   dplyr::distinct(location, pop_brk, spatial_scale)

fig_pops <- pre_pops %>%
  filter_ms_data(which_setting = which_setting) %>% ## filter out country and endemic locations
  dplyr::count(pop_brk, spatial_scale)

## plot
ggplot(fig_pops, aes(x = pop_brk, y = n)) +
  geom_col(aes(fill = spatial_scale)) +
  scale_x_discrete("Population") +
  scale_y_continuous("Locations") +
  scale_fill_brewer(palette = "Dark2") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    axis.text = element_text(size = 12)
  )

knitr::kable(fig_pops)

## totals by population bracket
summary_df <- fig_pops %>%
  dplyr::group_by(pop_brk) %>%
  dplyr::summarise(n = sum(n), .groups = "drop")

total_n <- sum(summary_df$n)

n_lt50   <- summary_df %>% dplyr::filter(pop_brk == "<50k") %>% dplyr::pull(n)
n_50_500 <- summary_df %>% dplyr::filter(pop_brk == "[50k, 500k)") %>% dplyr::pull(n)
n_gt500  <- summary_df %>% dplyr::filter(pop_brk == "≥ 500k") %>% dplyr::pull(n)

p_lt50   <- round(100 * n_lt50 / total_n, 1)
p_50_500 <- round(100 * n_50_500 / total_n, 1)
p_gt500  <- round(100 * n_gt500 / total_n, 1)

text_out <- glue::glue(
  "{total_n} location-population combinations were included. ",
  "{n_lt50} ({p_lt50}%) had a population less than 50,000, ",
  "{n_50_500} ({p_50_500}%) had a population between 50,000 and 500,000, ",
  "and {n_gt500} ({p_gt500}%) had a population over 500,000 ",
  "(Supp Fig - Population distribution plot)."
)

text_out

```

# Presentations Only

Prepare population distribution figure, cleaned just for presentations

```{r}
#| label: fig-pop-distr-prez
#| fig.width: 4.75
#| fig.height: 3.5

pre_pops <- dplyr::distinct(pre_extract, location, spatial_scale, pop_brk)

# checks
pre_pops %>% dplyr::group_by(location, pop_brk) %>% count() %>% filter(n>1)

fig_pops <- pre_pops %>%
  dplyr::group_by(pop_brk, spatial_scale) %>%
  dplyr::count() %>%
  dplyr::filter(!grepl("composite", spatial_scale)) ## remove composite locations just for a clean figure

ggplot(fig_pops, aes(x = pop_brk, y = n)) +
  geom_col(aes(fill = spatial_scale)) +
  scale_x_discrete("Population") +
  scale_y_continuous("Locations") +
  theme_bw() +
  theme(legend.position = "bottom", legend.title = element_blank(), axis.text = element_text(size=12))


```

# Proportion of alerts in the countries where the 50 weekly alert performs the best

```{r}
top1_countries <- c("COD", "ETH", "KEN", "TZA", "MWI")
alerts_subnational <- alerts %>% filter(spatial_scale != "country")
alerts_top1_countries <- alerts_subnational %>% filter(country %in% top1_countries)
proportion <- nrow(alerts_top1_countries)/nrow(alerts_subnational)

```

The countries where 50 weekly is the top-performing alert are: `r top1_countries`, and account for `r nrow(alerts_top1_countries)` out of 
`r nrow(alerts_subnational)` subnational alerts, representing 
`r round(proportion * 100, 1)`% of all subnational alerts.

# Country-level denominators for the top-3 heatmap (Fig 4)

In the country-stratified panel of the alert heatmap, each cell counts the number of
countries in which an alert definition ranked in the top three by utility score. The
appropriate denominator is defined **per alert definition**: a country can only rank a
definition it can score, and a complete utility score requires all five dimensions
(impact, efficiency, PPV, missed proportion, delay). The missed and delay dimensions in
turn require at least one recorded outbreak in the country's administrative units of that
population size, so a country may trigger a definition (raise at least one alert) without
being able to rank it. The number of rankable countries is therefore typically smaller
than the number of countries where a definition triggered, and it varies by definition
(stringent thresholds are rankable in fewer countries).

The chunk below computes, per population band and per alert definition:

- `n_countries_top3`: countries where the definition ranked in the top three (numerator)
- `n_countries_rankable`: countries with a complete utility score for the definition (the denominator)
- `n_countries_triggered`: countries where the definition triggered at least one alert

It also reports band-level maxima used in the figure caption.

```{r}
#| label: tbl-country-denominators

## "triggered" = definition raised >=1 alert in the country (ppv computable)
## "rankable"  = definition has a complete utility score (all 5 dimensions present)

## ---- per band, per definition ----
denom_by_def <- purrr::imap_dfr(
  country_pop_results,
  ~ .x %>%
      dplyr::group_by(alert_lab) %>%
      dplyr::summarise(
        n_countries_triggered = dplyr::n_distinct(country[!is.na(ppv)]),
        n_countries_rankable  = dplyr::n_distinct(
          country[!is.na(util_score) & !is.nan(util_score)]
        ),
        .groups = "drop"
      ) %>%
      dplyr::mutate(pop_brk = .y)
) %>%
  ## bring in the top-3 counts already computed for the figure
  dplyr::left_join(
    top3_freq_by_pop %>%
      dplyr::select(pop_brk, alert_lab, n_countries_top3 = n_countries),
    by = c("pop_brk", "alert_lab")
  ) %>%
  dplyr::mutate(n_countries_top3 = tidyr::replace_na(n_countries_top3, 0L)) %>%
  dplyr::select(
    pop_brk, alert_lab,
    n_countries_top3, n_countries_rankable, n_countries_triggered
  ) %>%
  dplyr::arrange(pop_brk, dplyr::desc(n_countries_triggered))

readr::write_csv(
  denom_by_def,
  paste0(fig_path, "/country_denominators_by_definition_", ocv_status, ".csv")
)

## ---- band-level summary (distinct countries + per-definition maxima) ----
denom_by_band <- purrr::imap_dfr(
  country_pop_results,
  ~ tibble::tibble(
      pop_brk               = .y,
      n_countries_with_data = dplyr::n_distinct(.x$country),
      n_countries_triggered = dplyr::n_distinct(.x$country[!is.na(.x$ppv)]),
      n_countries_rankable  = dplyr::n_distinct(
        .x$country[!is.na(.x$util_score) & !is.nan(.x$util_score)]
      )
    )
) %>%
  dplyr::left_join(
    denom_by_def %>%
      dplyr::group_by(pop_brk) %>%
      dplyr::summarise(
        max_triggered_any_def = max(n_countries_triggered),
        max_rankable_any_def  = max(n_countries_rankable),
        .groups = "drop"
      ),
    by = "pop_brk"
  )

readr::write_csv(
  denom_by_band,
  paste0(fig_path, "/country_denominators_by_band_", ocv_status, ".csv")
)

knitr::kable(
  denom_by_band,
  caption = "Country denominators per population band (Fig 4 country-stratified panel)"
)
```

Per-definition detail (numerator and both candidate denominators):

```{r}
#| label: tbl-country-denominators-detail

reactable_template(
  denom_by_def,
  col_names = list(
    pop_brk               = reactable::colDef(name = "Population Group"),
    alert_lab             = reactable::colDef(name = "Alert Def"),
    n_countries_top3      = reactable::colDef(name = "N Top-3", defaultSortOrder = "desc"),
    n_countries_rankable  = reactable::colDef(name = "N Rankable"),
    n_countries_triggered = reactable::colDef(name = "N Triggered", defaultSortOrder = "desc")
  )
)
```

Caption figures, pulled directly from the band-level summary so they update with the data:

```{r}
#| label: caption-country-denominators

rk <- setNames(denom_by_band$n_countries_rankable,  denom_by_band$pop_brk)
tg <- setNames(denom_by_band$max_triggered_any_def, denom_by_band$pop_brk)

caption_text <- glue::glue(
  "Cell values give the number of countries in which each alert definition ranked among ",
  "the top three by utility score, out of the countries with a complete utility score in ",
  "each band (up to {rk[['< 50k']]} for <50k, {rk[['[50k, 500k)']]} for 50k-500k, and ",
  "{rk[['\U2265 500k']]} for >=500k). The denominator is defined per definition, as a ",
  "country can only rank a definition it can score; the <50k band rests on a small ",
  "rankable set and should be interpreted with caution. Because alert definitions were not ",
  "triggered in every country, individual definitions need not reach these maxima \U2014 the ",
  "most general definitions triggered in up to {tg[['< 50k']]}, {tg[['[50k, 500k)']]}, and ",
  "{tg[['\U2265 500k']]} countries respectively."
)

caption_text
```
