---
title: "Hierarchical Log Ratio Models"
format:
  html:
    self-contained: true
params:
  model_choice: "modelB2" 
  rerun_model: false   
  alert_no: 8               
  pop_group: "50Kto500K"           
  location_filter: "outbreak-prone"   
  run_model_selection: false 
  constant: 1
  do_prior_check: true  
  compare_epidemic_endemic_posteriors: false 
  transform_posterior: true 
  rate_ratio_model: false
  remove_censored: false
  fig_path: "manuscript_figures"
editor: source
---

```{r setup, warning=FALSE, message=FALSE}
#| code-fold: true

library(tidyverse)
library(rstan)
library(loo)
library(bayesplot)
library(posterior)
library(glue)
library(parallel)
library(OutbreakExtractR)
library(kableExtra)
library(patchwork)
library(knitr)
library(DT)
library(stringr)

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = FALSE)
theme_set(theme_minimal())

## user selections
model_choice <- params$model_choice  ## options: "modelA1", "modelA2", "modelB", "modelB2", "modelC", "modelD"
rerun_model <- params$rerun_model    ## set to FALSE to load previous fit
alert_no <- params$alert_no               ## options: 1 to 24
pop_group <- params$pop_group          ## options: "under_50K", "50Kto500K", "over_500K"
location_filter <- params$location_filter  ## options: "all", "endemic", "outbreak-prone"
run_model_selection <- params$run_model_selection ## whether to run the model selection with LOO and WAIC
constant <- params$constant ## constant added to the calculation of the ratio to ensure we have no undefined ratios when cases are 0
do_prior_check <- params$do_prior_check  ## whether to run a prior predictive check
compare_epidemic_endemic_posteriors <- params$compare_epidemic_endemic_posteriors ## whether to compare posteriors of alpha0 between settings, use if model fits exist
transform_posterior <- params$transform_posterior ## whether to transform posterior to exp(posterior)-1 to get effect of delay per week in the natural scale
rate_ratio_model <- params$rate_ratio_model
remove_censored <- params$remove_censored
fig_path <- params$fig_path ## where we save slope summaries to get manuscript figure 5
pre_path <- "../data/"

## load stan model
stan_file <- switch(model_choice,
  "modelA1" = "../stan/log_ratio_model_A1.stan",
  "modelA2" = "../stan/log_ratio_model_A2.stan",
  "modelB" = "../stan/log_ratio_model_B.stan",
  "modelB2" = "../stan/log_ratio_model_B2.stan",
  "modelB3" = "../stan/log_ratio_model_B3.stan",  
  "modelC" = "../stan/log_ratio_model_C.stan",
  "modelD" = "../stan/log_ratio_model_D.stan",  
)

fit_file <- paste0("fit_", model_choice, "_a", alert_no, "_", pop_group, "_", location_filter,".rds")
```

```{r headline, echo=FALSE, results='asis'}

## map model choice to description
model_description <- switch(model_choice,
  "modelA1" = "A1: Simple null model with global rate",
  "modelA2" = "A2: Global slope and alert-specific deviation",
  "modelB" = "B: Global slope and country-specific deviation",
  "modelB2" = "B2: Global slope, country and alert-level deviation",
  "modelB3" = "B3: Global slope, country, admin1, and alert-level deviation",
  "modelC" = "C: Global slope and location-specific deviation",
  "modelD" = "D: Global slope, admin1, and alert-level deviation"
)

## create headline
cat(glue("# Alert {alert_no} – Population Group: {pop_group} – {model_description} - {location_filter}\n"))

```

## Model Specifications

::::{.center}
\begin{longtable}{p{2cm} p{12cm}}
\hline
\textbf{Model} & \textbf{Equations} \\
\hline

\textbf{A2} &
\[
r_i = \log\left(\frac{y_{d_i}}{y_{\text{delay at week 0}}}\right)
\]
\[
r_i \sim N(\mu_i,\sigma), \qquad \sigma \sim \text{HalfNormal}(1)
\]
\[
\mu_i = \beta_{alert[i]} \, d_i
\]
\[
\beta_{alert[i]} = \alpha_{global} + \alpha_{alert[i]}
\]
\[
\alpha_{global} \sim N(0,\sigma_{global}), \quad \sigma_{global} \sim \text{HalfNormal}(1)
\]
\[
\alpha_{alert} \sim N(0,\sigma_{alert}), \quad \sigma_{alert} \sim \text{HalfNormal}(0.25)
\]
\\[1em]

\textbf{B} &
\[
r_i = \log\left(\frac{y_{d_i}}{y_{\text{delay at week 0}}}\right)
\]
\[
r_i \sim N(\mu_i,\sigma), \qquad \sigma \sim \text{HalfNormal}(1)
\]
\[
\mu_i = \beta_{country[i]} \, d_i
\]
\[
\beta_{country[i]} = \alpha_{global} + \alpha_{country[i]}
\]
\[
\alpha_{global} \sim N(0,\sigma_{global}), \quad \sigma_{global} \sim \text{HalfNormal}(1)
\]
\[
\alpha_{country} \sim N(0,\sigma_{country}), \quad \sigma_{country} \sim \text{HalfNormal}(0.5)
\]
\\[1em]

\textbf{B2} &
\[
r_i = \log\left(\frac{y_{d_i}}{y_{\text{delay at week 0}}}\right)
\]
\[
r_i \sim N(\mu_i,\sigma), \qquad \sigma \sim \text{HalfNormal}(1)
\]
\[
\mu_i = \beta_{alert[i]} \, d_i
\]
\[
\beta_{alert[i]} = \alpha_{global} + \alpha_{country[i]} + \alpha_{alert[i]}
\]
\[
\alpha_{global} \sim N(0,\sigma_{global}), \quad \sigma_{global} \sim \text{HalfNormal}(1)
\]
\[
\alpha_{country} \sim N(0,\sigma_{country}), \quad \sigma_{country} \sim \text{HalfNormal}(0.5)
\]
\[
\alpha_{alert} \sim N(0,\sigma_{alert}), \quad \sigma_{alert} \sim \text{HalfNormal}(0.25)
\]
\\[1em]

\textbf{B3} &
\[
r_i = \log\left(\frac{y_{d_i}}{y_{\text{delay at week 0}}}\right)
\]
\[
r_i \sim N(\mu_i,\sigma), \qquad \sigma \sim \text{HalfNormal}(1)
\]
\[
\mu_i = \beta_{alert[i]} \, d_i
\]
\[
\beta_{alert[i]} = \alpha_{global} + \alpha_{country[i]} + \alpha_{admin1[i]} + \alpha_{alert[i]}
\]
\[
\alpha_{global} \sim N(0,\sigma_{global}), \quad \sigma_{global} \sim \text{HalfNormal}(1)
\]
\[
\alpha_{country} \sim N(0,\sigma_{country}), \quad \sigma_{country} \sim \text{HalfNormal}(0.5)
\]
\[
\alpha_{admin1} \sim N(0,\sigma_{admin1}), \quad \sigma_{admin1} \sim \text{HalfNormal}(0.5)
\]
\[
\alpha_{alert} \sim N(0,\sigma_{alert}), \quad \sigma_{alert} \sim \text{HalfNormal}(0.25)
\]
\\[1em]

\textbf{C} &
\[
r_i = \log\left(\frac{y_{d_i}}{y_{\text{delay at week 0}}}\right)
\]
\[
r_i \sim N(\mu_i,\sigma), \qquad \sigma \sim \text{HalfNormal}(1)
\]
\[
\mu_i = \beta_{location[i]} \, d_i
\]
\[
\beta_{location[i]} = \alpha_{global} + \alpha_{location[i]}
\]
\[
\alpha_{global} \sim N(0,\sigma_{global}), \quad \sigma_{global} \sim \text{HalfNormal}(1)
\]
\[
\alpha_{location} \sim N(0,\sigma_{location}), \quad \sigma_{location} \sim \text{HalfNormal}(0.5)
\]
\\[1em]

\textbf{D} &
\[
r_i = \log\left(\frac{y_{d_i}}{y_{\text{delay at week 0}}}\right)
\]
\[
r_i \sim N(\mu_i,\sigma), \qquad \sigma \sim \text{HalfNormal}(1)
\]
\[
\mu_i = \beta_{alert[i]} \, d_i
\]
\[
\beta_{alert[i]} = \alpha_{global} + \alpha_{admin1[i]} + \alpha_{alert[i]}
\]
\[
\alpha_{global} \sim N(0,\sigma_{global}), \quad \sigma_{global} \sim \text{HalfNormal}(1)
\]
\[
\alpha_{admin1} \sim N(0,\sigma_{admin1}), \quad \sigma_{admin1} \sim \text{HalfNormal}(0.5)
\]
\[
\alpha_{alert} \sim N(0,\sigma_{alert}), \quad \sigma_{alert} \sim \text{HalfNormal}(0.25)
\]
\\[1em]

\hline
\end{longtable}
::::


```{r}
#| label: load-data
#| code-fold: true

## load evaluation data for all delay values
delays <- seq(from = 0, to = 24, by = 1)

data_list <- map(delays, ~ readRDS(paste0("../../main_ecl_dec2024/outputs/evaluation_cases_dw", ., "_ew52.rds")) %>%
                   mutate(delay = .x))

data <- bind_rows(data_list)

data <- data %>%
  mutate(
    ## extract admin1
    admin1 = sapply(str_split(alert_id, "::"), function(x) if (length(x) >= 3) x[3] else NA),
    ## extract spatial scale based on number of "::"
    spatial_scale = case_when(
      str_count(alert_id, "::") == 1 ~ "country",
      str_count(alert_id, "::") == 2 ~ "admin1",
      str_count(alert_id, "::") >= 3 ~ "admin2",
      TRUE ~ "unknown"
    )
  )

## applying OutbreakExtractR filtering function
which_setting <- case_when(
  location_filter == "endemic" ~ "endemic",
  location_filter == "outbreak-prone" ~ "epidemic",
  TRUE ~ "all"
)

data <- OutbreakExtractR::filter_ms_data(
  df = data,
  which_setting = which_setting,
  incl_trend_alerts = TRUE
)

data <- data %>%
  mutate(
    alert_number = str_extract(alert_id, "(?<=^[A-Za-z])\\d+(?=_)")
  ) %>%
  filter(alert_number == alert_no) %>%
  select(alert_id, total_sCh, delay, location, admin1, spatial_scale) %>%
  rename(Sch = total_sCh) %>%
  mutate(Sch = as.integer(Sch)) %>%
  tidyr::extract(alert_id, into = "year", regex = "_(\\d{4})", remove = FALSE)

## load population data to select only required population group
ts_population <- readr::read_rds(file.path(pre_path, "time_series_preoutbreak_extraction_public_apr_2025.rds"))%>%
  transmute(location, year = as.character(year(TL)), pop) %>%
  distinct() %>%
  group_by(year, location) %>%
  slice(1) %>%
  ungroup()

data <- data %>%
  left_join(ts_population, by = c("location", "year")) %>%
  distinct()

## filter by population group
data <- data %>%
  filter(case_when(
    pop_group == "under_50K" ~ pop < 50000,
    pop_group == "50Kto500K" ~ pop >= 50000 & pop < 500000,
    pop_group == "over_500K" ~ pop >= 500000,
    TRUE ~ TRUE
  ))

data <- data %>%
  mutate(
    alert_id_index = as.integer(factor(alert_id)),
    location_id = as.integer(factor(location))
  )

## calculating ratios and log-ratios

if (rate_ratio_model) {

  ## load observed weeks
  observed_weeks <- readRDS("../../main_ecl_dec2024/outputs/evaluation_time_censoring.rds")

  ## join only the columns we need, avoid duplicate names
  data <- data %>%
    left_join(
      observed_weeks %>% select(alert_id, delay, O_weeks),  # only join needed columns
      by = c("alert_id", "delay")
    )

  ## calculate ratios and log ratios
  data <- data %>%
    group_by(alert_id) %>%
    mutate(
      Sch_delay0 = Sch[delay == 0],
      O_weeks_delay0 = O_weeks[delay == 0],  ## observed weeks at delay 0
      Sch_ratio = ((Sch + constant) / (O_weeks + constant)) / ((Sch_delay0 + constant) / (O_weeks_delay0 + constant)),
      Log_Sch_ratio = log(Sch_ratio)
    ) %>%
    ungroup() %>%
    filter(delay != 0)
  } else {

  ## optionally remove censored alerts
  if (remove_censored) {

    observed_weeks <- readRDS("../../main_ecl_dec2024/outputs/evaluation_time_censoring.rds")

    data <- data %>%
      left_join(
        observed_weeks %>% select(alert_id, delay, O_weeks),
        by = c("alert_id", "delay")
      ) %>%
      filter(O_weeks >= 52)
  }

  data <- data %>%
  group_by(alert_id) %>%
  mutate(Sch_delay0 = Sch[delay == 0]) %>%
  ungroup() %>%
  mutate(
    Sch_ratio = (Sch + constant) / (Sch_delay0 + constant),
    Log_Sch_ratio = log(Sch_ratio)
  ) %>%
  filter(delay != 0)
}

```

```{r}
#| label: prepare-stan-data
#| code-fold: true

prepare_stan_data <- function(data, model_choice) {
  factor_index <- function(x) as.integer(factor(x))
  if (model_choice == "modelA1") {
    stan_data <- list(
      N = nrow(data),
      r = data$Log_Sch_ratio,
      d = data$delay
    )
    
  } else if (model_choice == "modelA2") {
  data <- data %>% mutate(alert_index = factor_index(alert_id))
  
  stan_data <- list(
    N = nrow(data),
    J = length(unique(data$alert_index)),
    alert = data$alert_index,
    r = data$Log_Sch_ratio,
    d = data$delay
  )
  
  lookup <- list(
    alert_lookup = data %>%
      distinct(alert_index, alert_id) %>%
      arrange(alert_index)
  )
  
  return(list(data = data, stan_data = stan_data, lookup = lookup))
} else if (model_choice == "modelB") {
  data <- data %>%
    mutate(
      country = str_extract(location, "(?<=::)[A-Z]{3}(?=::|$)"),  
      country_id = factor_index(country)
    )
  
  stan_data <- list(
    N = nrow(data),
    C = length(unique(data$country_id)),
    country = data$country_id,   
    r = data$Log_Sch_ratio,
    d = data$delay
  )
  
  lookup <- list(
    country_lookup = data %>% distinct(country_id, country)
  )
  } else if (model_choice == "modelB2") {
  data <- data %>%
    mutate(
      country = str_extract(location, "(?<=::)[A-Z]{3}(?=::|$)"),  
      country_id = factor_index(country),
      alert_index = factor_index(alert_id)
    )

  alert_country_lookup <- data %>%
    distinct(alert_index, country_id, alert_id, country) %>%
    arrange(alert_index)

  stan_data <- list(
    N = nrow(data),
    C = length(unique(data$country_id)),
    J = length(unique(data$alert_index)),
    country = data$country_id,
    alert = data$alert_index,
    country_alert = alert_country_lookup$country_id,  
    r = data$Log_Sch_ratio,
    d = data$delay
  )

  lookup <- list(
    alert_country_lookup = alert_country_lookup,
    country_lookup       = data %>% distinct(country_id, country)
  )
  } else if (model_choice == "modelB3") {
    if (!"admin1" %in% names(data)) stop("admin1 column missing before prepare_stan_data() for modelB3")
    
    data <- data %>%
      mutate(
        country = str_extract(location, "(?<=::)[A-Z]{3}(?=::|$)"),
        country_id = factor_index(country),
        admin1_combined = paste(country, admin1, sep = "::"),
        admin1_index = factor_index(admin1_combined),
        alert_index = factor_index(alert_id)
      )
    
    admin1_lookup <- data %>%
      distinct(admin1_index, country_id, country, admin1) %>%
      arrange(admin1_index)
    
    alert_lookup <- data %>%
      distinct(alert_index, admin1_index, alert_id) %>%
      arrange(alert_index)
    
    alert_country_admin1_lookup <- alert_lookup %>%
      left_join(admin1_lookup, by = "admin1_index")  
    
    U <- nrow(admin1_lookup)
    J <- nrow(alert_lookup)
    
    stan_data <- list(
      N = nrow(data),
      C = length(unique(data$country_id)),
      U = U,
      J = J,
      country = data$country_id,
      admin1 = data$admin1_index,
      alert = data$alert_index,
      country_admin1 = admin1_lookup$country_id,  # length U
      admin1_alert = alert_lookup$admin1_index,   # length J
      r = data$Log_Sch_ratio,
      d = data$delay
    )
    
    lookup <- list(
      country_lookup = data %>% distinct(country_id, country),
      admin1_lookup  = admin1_lookup,
      alert_lookup   = alert_lookup,
      alert_country_admin1_lookup = alert_country_admin1_lookup
    )
    return(list(data = data, stan_data = stan_data, lookup = lookup))
    
  } else if (model_choice == "modelC") {

  location_lookup <- data %>%
    distinct(location_id, location) %>%
    arrange(location_id) %>%
    mutate(loc_index = 1:n())
  
  data <- data %>%
    left_join(location_lookup, by = c("location_id", "location"))

  stan_data <- list(
    N = nrow(data),
    E = nrow(location_lookup),   
    loc = data$loc_index,
    r = data$Log_Sch_ratio,
    d = data$delay
  )
  
  lookup <- list(
    location_lookup = location_lookup
  )
  
  return(list(data = data, stan_data = stan_data, lookup = lookup))
} else if (model_choice == "modelD") {
  if (!"admin1" %in% names(data)) stop("admin1 column missing before prepare_stan_data() for modelD")

  data <- data %>%
    mutate(
      country = str_extract(location, "(?<=::)[A-Z]{3}(?=::|$)"),
      country_id = factor_index(country),
      admin1_combined = paste(country, admin1, sep = "::"),
      admin1_index = factor_index(admin1_combined),
      alert_index = factor_index(alert_id)
    )

  admin1_lookup <- data %>%
    distinct(admin1_index, country_id, country, admin1) %>%
    arrange(admin1_index)

  alert_lookup <- data %>%
    distinct(alert_index, alert_id, admin1_index) %>%
    arrange(alert_index)

  alert_admin1_lookup <- alert_lookup %>%
    left_join(admin1_lookup, by = "admin1_index")

  U <- nrow(admin1_lookup)
  J <- nrow(alert_lookup)

  stan_data <- list(
    N = nrow(data),
    U = U,
    J = J,
    admin1 = data$admin1_index,
    alert = data$alert_index,
    admin1_alert = alert_lookup$admin1_index,  
    r = data$Log_Sch_ratio,
    d = data$delay
  )

  lookup <- list(
    admin1_lookup = admin1_lookup,
    alert_lookup  = alert_lookup,
    alert_admin1_lookup = alert_admin1_lookup
  )

  return(list(data = data, stan_data = stan_data, lookup = lookup))
}
 else {
    stop("Unknown model_choice")
  }
  
  return(list(data = data, stan_data = stan_data, lookup = lookup))
}



```

```{r}
#| label: prepared-data-checks

models_need_admin1 <- c("modelB3", "modelD")

if (model_choice %in% models_need_admin1) {

  n_before <- nrow(data)
  n_na_admin1 <- sum(is.na(data$admin1) | data$admin1 == "")

  if (n_na_admin1 > 0) {
    message(glue::glue("Model {model_choice} requires admin1. Removing {n_na_admin1} observations with missing admin1 (out of {n_before})."))
  }

  data <- data %>%
    filter(!(is.na(admin1) | admin1 == ""))

  n_after <- nrow(data)
  if (n_after == 0) stop("No data left after removing rows with missing admin1. Check your filters (alert_no, population, location_filter).")

  ## quick sanity checks
  n_admin1 <- data %>% distinct(admin1) %>% nrow()
  if (n_admin1 < 5) {
    warning(glue::glue("Only {n_admin1} unique admin1 units remain — hierarchical admin1 estimates may be weak."))
  }
}

## prepare stan_data
prepared <- prepare_stan_data(data, model_choice)
data <- prepared$data
stan_data <- prepared$stan_data

message(glue::glue("Prepared stan_data for {model_choice}: N={stan_data$N}"))
if (!is.null(stan_data$C)) message(glue::glue("Countries: {stan_data$C}"))
if (!is.null(stan_data$U)) message(glue::glue("Admin1 units: {stan_data$U}"))
if (!is.null(stan_data$J)) message(glue::glue("Alerts: {stan_data$J}"))


```

```{r}
#| label: run-or-reload-model
#| code-fold: true

if (rerun_model || !file.exists(fit_file)) {
  message("Running Stan model...")

  fit <- rstan::stan(
    file = stan_file,
    data = stan_data,
    chains = 4,
    iter = 6000,
    warmup = 3000,
    seed = 123,
    control = list(adapt_delta = 0.98, max_treedepth = 15)
  )
 
  message("Finished running Stan model...")
  saveRDS(fit, fit_file)
  message("Model fit saved to: ", fit_file)

} else {
  if (!file.exists(fit_file)) {
    stop("Saved model file not found: ", fit_file)
  }

  message("Loading saved model fit from: ", fit_file)
  fit <- readRDS(fit_file)
}

```

## Priors Used

```{r}
#| label: print-priors
#| code-fold: true

## function to extract priors from stan model
extract_stan_priors <- function(stan_file) {
  stan_lines <- readLines(stan_file)
  
  ## keep only lines with ~
  prior_lines <- stan_lines[grepl("~", stan_lines)]
  
  ## remove comments after // or #
  prior_lines <- sub("//.*", "", prior_lines)
  prior_lines <- sub("#.*", "", prior_lines)
  
  prior_lines <- prior_lines[grepl("[a-zA-Z0-9_]+\\s*~", prior_lines)]
  
  priors_list <- data.frame(
    Parameter = sub("\\s*~.*", "", prior_lines) |> trimws(),
    Prior     = sub(".*~\\s*", "", prior_lines) |> trimws(),
    stringsAsFactors = FALSE
  )
  
  priors_list$Prior <- gsub(";", "", priors_list$Prior)
  
  half_normal_idx <- grepl("^(tau|sigma)", priors_list$Parameter)
  
  scales <- sub(".*normal\\([^,]+,\\s*([0-9\\.eE+-]+)\\).*", "\\1", priors_list$Prior[half_normal_idx])
  
  priors_list$Prior[half_normal_idx] <- paste0("half-normal(", scales, ")")
  
  return(priors_list)
}

priors_list <- extract_stan_priors(stan_file)

priors_list %>%
  kable(format = "html", caption = paste0("Priors extracted from ", model_choice)
  ) %>%
  kable_styling(full_width = FALSE, bootstrap_options = c("hover", "striped", "condensed", "responsive")
  )

```

```{r}
#| label: print-priors-2
#| code-fold: true
#| results: asis

## priors from all models for manuscript table with priors

## model lists
all_models <- c("modelA1","modelA2","modelB","modelB2","modelB3","modelC","modelD")

stan_paths <- list(
  modelA1 = "../stan/log_ratio_model_A1.stan",
  modelA2 = "../stan/log_ratio_model_A2.stan",
  modelB  = "../stan/log_ratio_model_B.stan",
  modelB2 = "../stan/log_ratio_model_B2.stan",
  modelB3 = "../stan/log_ratio_model_B3.stan",
  modelC  = "../stan/log_ratio_model_C.stan",
  modelD  = "../stan/log_ratio_model_D.stan"
)

model_descriptions <- list(
  modelA1 = "A1",
  modelA2 = "A2",
  modelB  = "B",
  modelB2 = "B2",
  modelB3 = "B3",
  modelC  = "C",
  modelD  = "D"
)

param_labels <- tibble::tribble(
  ~Parameter,       ~ParamLabel,
  "tau_alert",      "<em>σ<sub>alert</sub></em>",
  "alpha_alert",    "<em>α<sub>alert</sub></em>",
  "tau_country",    "<em>σ<sub>country</sub></em>",
  "alpha_country",  "<em>α<sub>country</sub></em>",
  "tau_admin1",     "<em>σ<sub>admin1</sub></em>",
  "alpha_admin1",   "<em>α<sub>admin1</sub></em>",
  "tau_loc",        "<em>σ<sub>loc</sub></em>",
  "alpha_loc",      "<em>α<sub>loc</sub></em>",
  "sigma",          "<em>σ</em>",
  "beta",           "<em>β</em>",
  "alpha0",         "<em>α<sub>global</sub></em>",
  "alpha_admin1_std","<em>α<sub>admin1_std</sub></em>"
)

## extract priors
priors_all <- purrr::map_df(all_models, function(m){
  p <- extract_stan_priors(stan_paths[[m]])
  p$model <- model_descriptions[[m]]
  p
})

priors_all <- priors_all %>%
  mutate(
    Prior = as.character(Prior),
    Prior = gsub("(?i)Normal", "𝒩", Prior, perl = TRUE),       
    Prior = gsub("(?i)HalfNormal", "<em>HalfNormal</em>", Prior, perl = TRUE)
  )

param_lookup <- setNames(param_labels$ParamLabel, param_labels$Parameter)

replace_params_in_prior <- function(txt) {
  purrr::reduce(names(param_lookup), function(acc, key) {
    gsub(key, param_lookup[[key]], acc, fixed = TRUE)
  }, .init = txt)
}

priors_all <- priors_all %>%
  left_join(param_labels, by = "Parameter") %>%
  mutate(
    ParamLabel = coalesce(ParamLabel, Parameter),
    Prior = replace_params_in_prior(Prior)
  )

priors_table <- priors_all %>%
  select(model, ParamLabel, Prior) %>%
  pivot_wider(
    names_from = ParamLabel,
    values_from = Prior,
    values_fill = "",
    values_fn = list(Prior = ~as.character(.))  # force character
  ) %>%
  arrange(model) %>%
  select(-any_of("r"))

## print
priors_table %>%
  kable("html", escape = FALSE, caption = "Model Priors") %>%
  kable_styling(full_width = TRUE, position = "center") %>%
  as.character() %>%
  htmltools::HTML() %>%
  htmltools::div(style = "width:80%; margin:auto;") %>%
  print()


```

## Model Fit Summary


```{r}
#| label: model-fit-summary-stan
#| code-fold: true
#| eval: true
#| results: asis

extract_summary_table <- function(fit, pars = NULL, probs = c(0.025, 0.5, 0.975)) {
  all_pars <- names(rstan::extract(fit))
  if (!is.null(pars)) pars <- intersect(pars, all_pars)
  
  s <- summary(fit, pars = pars, probs = probs)$summary
  df <- as.data.frame(s)
  
  df$Parameter <- rownames(df)
  rownames(df) <- NULL
  
  df <- df[, c("Parameter","mean","sd","2.5%","50%","97.5%","n_eff","Rhat")]
  colnames(df) <- c("Parameter","Mean","SD","2.5%","Median","97.5%","n_eff","Rhat")
  
  df[, -1] <- round(df[, -1], 3)
  df
}

pretty_table <- function(df, caption) {
  df %>%
    kbl(
      caption = caption,
      align = paste0("l", strrep("c", ncol(df) - 1)),
      booktabs = TRUE,
      linesep = ""
    ) %>%
    kable_styling(
      latex_options = "basic",
      full_width = TRUE,
      position = "center",
      font_size = 18
    )
}


report_model <- function(fit, lookup, model_name = "", location_filter = "", return_df = FALSE) {

  # Parameters to extract
  pars_to_include <- c(
    "alpha_global_exp_m1",
    "alpha_country_eff_exp_m1",
    "alpha_country",            # raw alpha_country
    "alpha_admin1_eff_exp_m1",
    "alpha_alert_eff_exp_m1",
    "alpha_loc_eff_exp_m1",
    "sigma",
    "tau_country",
    "tau_admin1",
    "tau_alert",
    "tau_loc"
  )
  
  df <- extract_summary_table(fit, pars = pars_to_include)

  df$Parameter[df$Parameter == "alpha_global_exp_m1"] <- "Global effect (% change)"
  
  country_rows <- grepl("^alpha_country_eff_exp_m1", df$Parameter)
  alpha_country_rows <- grepl("^alpha_country\\[", df$Parameter)
  
  if (any(country_rows | alpha_country_rows) && !is.null(lookup$country_lookup)) {
    # Extract indices
    idx_country_eff <- as.integer(str_extract(df$Parameter[country_rows], "(?<=\\[)\\d+(?=\\])"))
    idx_alpha_country <- as.integer(str_extract(df$Parameter[alpha_country_rows], "(?<=\\[)\\d+(?=\\])"))
    
    country_codes_eff <- lookup$country_lookup$country[match(idx_country_eff, lookup$country_lookup$country_id)]
    country_codes_raw <- lookup$country_lookup$country[match(idx_alpha_country, lookup$country_lookup$country_id)]
    
    df$Parameter[country_rows] <- paste0("Country: ", country_codes_eff, " (% change)")
    df$Parameter[alpha_country_rows] <- paste0("Country (deviation): ", country_codes_raw)
  }

  admin1_rows <- grepl("^alpha_admin1_eff_exp_m1", df$Parameter)
  if(any(admin1_rows) && !is.null(lookup$admin1_lookup)) {
    idx <- as.integer(str_extract(df$Parameter[admin1_rows], "(?<=\\[)\\d+(?=\\])"))
    admin1_names <- lookup$admin1_lookup$admin1[match(idx, lookup$admin1_lookup$admin1_index)]
    countries <- lookup$admin1_lookup$country[match(idx, lookup$admin1_lookup$admin1_index)]
    df$Parameter[admin1_rows] <- paste0("Admin1: ", countries, "::", admin1_names, " (% change)")
  }

  loc_rows <- grepl("^alpha_loc_eff_exp_m1", df$Parameter)
  if(any(loc_rows) && !is.null(lookup$location_lookup)) {
    idx <- as.integer(str_extract(df$Parameter[loc_rows], "(?<=\\[)\\d+(?=\\])"))
    loc_names <- lookup$location_lookup$location[match(idx, lookup$location_lookup$loc_index)]
    df$Parameter[loc_rows] <- paste0("Location: ", loc_names, " (% change)")
  }

  alert_rows <- grepl("^alpha_alert_eff_exp_m1", df$Parameter)
  if(any(alert_rows)) {
    idx <- as.integer(str_extract(df$Parameter[alert_rows], "(?<=\\[)\\d+(?=\\])"))
    
    if(!is.null(lookup$alert_country_admin1_lookup)) {
      tmp <- lookup$alert_country_admin1_lookup
      df$Parameter[alert_rows] <- paste0(
        "Alert: ", tmp$country[match(idx, tmp$alert_index)], "::",
        tmp$admin1[match(idx, tmp$alert_index)], "::",
        tmp$alert_id[match(idx, tmp$alert_index)], " (% change)"
      )
    } else if(!is.null(lookup$alert_country_lookup)) {
      tmp <- lookup$alert_country_lookup
      df$Parameter[alert_rows] <- paste0(
        "Alert: ", tmp$country[match(idx, tmp$alert_index)], "::",
        tmp$alert_id[match(idx, tmp$alert_index)], " (% change)"
      )
    } else if(!is.null(lookup$alert_admin1_lookup)) {
      tmp <- lookup$alert_admin1_lookup
      df$Parameter[alert_rows] <- paste0(
        "Alert: ", tmp$admin1[match(idx, tmp$alert_index)], "::",
        tmp$alert_id[match(idx, tmp$alert_index)], " (% change)"
      )
    } else if(!is.null(lookup$alert_lookup)) {
      tmp <- lookup$alert_lookup
      df$Parameter[alert_rows] <- paste0(
        "Alert: ", tmp$alert_id[match(idx, tmp$alert_index)], " (% change)"
      )
    }
  }
  
  df$Parameter[df$Parameter == "sigma"] <- "Measurement SD"
  df$Parameter[df$Parameter == "tau_country"] <- "Country-level SD"
  df$Parameter[df$Parameter == "tau_admin1"] <- "Admin1-level SD"
  df$Parameter[df$Parameter == "tau_alert"] <- "Alert-level SD"
  df$Parameter[df$Parameter == "tau_loc"] <- "Location-level SD"
  
  if(return_df) return(df)
  pretty_table(df, paste("Posterior summary for", model_name, location_filter))
}

posterior_summaries <- report_model(fit = fit, lookup = prepared$lookup)
posterior_summaries



```


## Traceplots and Posterior Densities

```{r}
#| label: traceplots
#| code-fold: true
#| eval: true

# ## helper function to plot traceplots for n random indices of a vector parameter
plot_random_indices <- function(fit, param, n = 3) {

  ## extract parameter names from Stan fit
  param_names <- dimnames(rstan::extract(fit, permuted = FALSE))[[3]]

  candidates <- grep(paste0("^", param, "\\["), param_names, value = TRUE)

  if (length(candidates) == 0) {
    if (param %in% param_names) {
      message("Parameter '", param, "' is scalar — plotting directly.")
      print(mcmc_trace(fit, pars = param))
    } else {
      message("Parameter '", param, "' not found in fit (check spelling or model_choice).")
    }
    return(invisible(NULL))
  }

  ## select random indices for plotting
  sel_pars <- sample(candidates, min(n, length(candidates)))
  message("Plotting random indices of ", param, ": ", paste(sel_pars, collapse = ", "))

  ## traceplots
  print(mcmc_trace(fit, pars = sel_pars))
}


## helper function to plot alpha_alert in modelB2 and B3
plot_random_alpha_alert <- function(fit, n = 3) {
  param_names <- dimnames(rstan::extract(fit, permuted = FALSE))[[3]]
  candidates <- grep("^alpha_alert\\[", param_names, value = TRUE)

  if(length(candidates) == 0) {
    message("No alpha_alert parameters found in fit.")
    return(invisible(NULL))
  }

  sel_pars <- sample(candidates, min(n, length(candidates)))
  message("Plotting random alpha_alert indices: ", paste(sel_pars, collapse = ", "))

  print(mcmc_trace(fit, pars = sel_pars))
}


## plotting depending on the model choice

if(model_choice == "modelA1"){
  print(mcmc_trace(fit, pars = c("beta", "sigma")))
  print(mcmc_dens_overlay(fit, pars = c("beta", "sigma")))
  plot_random_indices(fit, "beta")
}

if(model_choice == "modelA2"){
  print(mcmc_trace(fit, pars = c("alpha0", "tau_alert", "sigma")))
  print(mcmc_dens_overlay(fit, pars = c("alpha0", "tau_alert", "sigma")))
  plot_random_indices(fit, "alpha_alert")
}

if(model_choice == "modelB"){
  print(mcmc_trace(fit, pars = c("alpha0", "tau_country", "sigma")))
  print(mcmc_dens_overlay(fit, pars = c("alpha0", "tau_country", "sigma")))
  plot_random_indices(fit, "alpha_country")
}

if(model_choice == "modelB2"){
  print(mcmc_trace(fit, pars = c("alpha0", "tau_country", "tau_alert", "sigma")))
  print(mcmc_dens_overlay(fit, pars = c("alpha0", "tau_country", "tau_alert", "sigma")))

  plot_random_indices(fit, "alpha_country")   ## country deviations
  plot_random_alpha_alert(fit, n = 5)        ## flattened alert deviations, length J
}

if(model_choice == "modelB3"){
  print(mcmc_trace(fit, pars = c("alpha0", "tau_country", "tau_alert","tau_admin1", "sigma")))
  print(mcmc_dens_overlay(fit, pars = c("alpha0", "tau_country", "tau_alert","tau_admin1", "sigma")))

  plot_random_indices(fit, "alpha_country")   ## country deviations
  plot_random_indices(fit, "alpha_admin1")    ## admin1 deviations
  plot_random_alpha_alert(fit, n = 5)         ## alert deviations
}

if(model_choice == "modelC"){
  print(mcmc_trace(fit, pars = c("alpha0", "tau_loc", "sigma")))
  print(mcmc_dens_overlay(fit, pars = c("alpha0", "tau_loc", "sigma")))
  plot_random_indices(fit, "alpha_loc")
}

if(model_choice == "modelD"){
  print(mcmc_trace(fit, pars = c("alpha0", "tau_alert","tau_admin1", "sigma")))
  print(mcmc_dens_overlay(fit, pars = c("alpha0", "tau_alert","tau_admin1", "sigma")))

  plot_random_indices(fit, "alpha_admin1")    ## admin1 deviations
  plot_random_alpha_alert(fit, n = 5)         ## alert deviations
}
```

```{r}
#| label: rhats
#| code-fold: true
#| eval: true

## function to compute R-hat per observation for r_rep ===
compute_rhat_rrep <- function(fit) {
  
  draws <- as_draws_df(fit)
  
  if (transform_posterior) {
    param_cols <- !grepl("^\\.", colnames(draws))
    draws[param_cols] <- lapply(draws[param_cols], function(x) exp(x) - 1)
  }
  
  r_rep_cols <- grep("^r_rep\\[", colnames(draws), value = TRUE)
  if (length(r_rep_cols) == 0) stop("No r_rep columns found.")
  
  tibble(
    Observation = seq_along(r_rep_cols),
    Rhat = sapply(r_rep_cols, function(col) rhat(draws[[col]]))
  )
}

## use the function
rhat_df <- compute_rhat_rrep(fit)

## the plot
ggplot(rhat_df, aes(x = Observation, y = Rhat)) +
  geom_point(color = "darkblue") +
  labs(
    title = "R-hat per observation",
    x = expression(hat(R)),
    y = "Observation"
  ) +
  theme_minimal()



```

## Prior Predictive Checks

```{r}
#| code-fold: true

## running prior predictive check
if (do_prior_check) {
  message("Running prior predictive simulation...")

  fit_prior <- rstan::stan(
    file = stan_file,
    data = stan_data,
    chains = 1,
    iter = 6000,
    seed = 123,
    algorithm = "Fixed_param"
  )

  ## explore prior predictive simulation
  prior_samples <- rstan::extract(fit_prior)
  
  r_prior_pred <- prior_samples$r_prior_pred
  tau_country_prior <- prior_samples$tau_country
  tau_alert_prior <- prior_samples$tau_alert
  sigma_prior <- prior_samples$sigma
  
  ## plot of prior predictive vs observed

  prior_pred_df <- tibble(
  observed = stan_data$r,
  prior_predicted = apply(r_prior_pred, 2, mean)  ## mean across prior draws
)
  
ggplot(prior_pred_df, aes(x = observed, y = prior_predicted)) +
  geom_point(alpha = 0.6, color = "lightblue4") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = "Prior Predictive Check: Observed vs Predicted",
       x = "Observed log-ratio",
       y = "Mean of Prior Predicted")

  
}

```

```{r}
#| label: prior-predictive-check
#| code-fold: true

if (do_prior_check) {

  r_mat <- as.matrix(r_prior_pred)
  ndraws <- nrow(r_mat)
  N <- ncol(r_mat)

  sel_draws <- sample(1:ndraws, min(200, ndraws))

  sim_df <- tibble(
    draw = rep(sel_draws, each = N),
    r = as.vector(t(r_mat[sel_draws, , drop = FALSE])),
    type = "Prior Simulated"
  )

  obs_df <- tibble(
    r = stan_data$r,
    type = "Observed"
  )

  plot_df <- bind_rows(sim_df, obs_df) %>%
    mutate(type = factor(type, levels = c("Prior Simulated", "Observed")))

  ggplot() +
    geom_density(
      data = plot_df %>% filter(type == "Prior Simulated"),
      aes(x = r, group = draw),
      fill = "steelblue", color = NA, alpha = 0.05
    ) +
    geom_density(
      data = plot_df %>% filter(type == "Observed"),
      aes(x = r),
      color = "black", size = 1.2
    ) +
    labs(
      title = "Density of Observed and Prior Predicted Log Ratios",
      x = "Log ratio",
      y = "Density"
    ) +
    theme_minimal(base_size = 14)
}


```

## Posterior Predictive Check Scatterplot

Here we generate data from the posterior and compare them to our observed log-ratios.

```{r}
#| code-fold: true
#| eval: true

posterior_samples <- rstan::extract(fit)

## generated data from the posterior
r_rep <- posterior_samples$r_rep  
r_hat <- apply(r_rep, 2, mean)   ## posterior predictive mean for each obs

ppc_df <- tibble(
  observed = stan_data$r,
  predicted = r_hat,
  delay = stan_data$d,
  loc = data$location
)

ggplot(ppc_df, aes(x = observed, y = predicted)) +
  geom_point(alpha = 0.6, color = "skyblue") +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(title = "Posterior Predictive Check: Observed vs Predicted",
       x = "Observed log-ratio",
       y = "Predicted mean log-ratio")



```

## Compare Prior Predictive, Posterior Predictive, and Observed

```{r}
#| eval: true

## prior predictive
r_mat_prior <- as.matrix(r_prior_pred)
sel_prior <- sample(1:nrow(r_mat_prior), min(200, nrow(r_mat_prior)))
prior_df <- tibble(
  draw = rep(sel_prior, each = ncol(r_mat_prior)),
  r    = as.vector(t(r_mat_prior[sel_prior, , drop = FALSE])),
  type = "Prior Predictive"
)

## posterior predictive
r_mat_post <- as.matrix(r_rep)
sel_post <- sample(1:nrow(r_mat_post), min(200, nrow(r_mat_post)))
post_df <- tibble(
  draw = rep(sel_post, each = ncol(r_mat_post)),
  r    = as.vector(t(r_mat_post[sel_post, , drop = FALSE])),
  type = "Posterior Predictive"
)

## observed
obs_df <- tibble(
  r    = stan_data$r,
  type = "Observed"
)

## combine
all_df <- bind_rows(prior_df, post_df, obs_df)

## plot
ggplot(all_df, aes(x = r, fill = type, color = type)) +
  geom_density(alpha = 0.2, size = 0.7) +
  scale_color_viridis_d(option = "D") +
  scale_fill_viridis_d(option = "D") +
  labs(
    title = "Observed vs Prior & Posterior Predictive Densities",
    x = "Log ratio",
    y = "Density",
    color = "",
    fill = ""
  ) +
  theme_minimal()


```

## Effect of Delay on Measurement Variability

```{r}
#| eval: true

## explore the effect of delay in measurement variability in observed vs posterior predictive log-ratios

## sd of observed log ratios
obs_summary <- data %>%
  group_by(delay) %>%
  summarise(
    sd_r = sd(Log_Sch_ratio, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(source = "Observed")

## sd of posterior predictive log ratios
ppc_summary <- ppc_df %>%
  group_by(delay) %>%
  summarise(
    sd_r = sd(predicted, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(source = "Posterior Predictive")

combined_summary <- bind_rows(obs_summary, ppc_summary)

## plot
ggplot(combined_summary, aes(x = delay, y = sd_r, color = source)) +
  geom_point(size = 2) +
  geom_line(linewidth = 0.8) +
  labs(
    title = "Standard deviation of log-ratios by delay",
    x = "Delay",
    y = "SD of log-ratios",
    color = "Source"
  ) +
  theme_minimal()

## plot with raw values

## raw values
obs_raw <- data %>%
  select(delay, value = Log_Sch_ratio) %>%
  mutate(source = "Observed")

ppc_raw <- ppc_df %>%
  select(delay, value = predicted) %>%
  mutate(source = "Posterior Predictive")

combined_raw <- bind_rows(obs_raw, ppc_raw)

## plot raw values per delay
ggplot(combined_raw, aes(x = factor(delay), y = value, color = source)) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 1.5) +
  labs(
    title = "Raw log-ratio values by delay",
    x = "Delay",
    y = "Log-ratio",
    color = "Source"
  ) +
  theme_minimal()

```

## Effective Slopes per level (summing global and deviation level posteriors per draw)

```{r,  fig.height=35, fig.width=15, warning=FALSE}
#| code-fold: true
#| eval: true

## function that gets the summed posteriors from the stan fit to plot them
summarize_effective_slopes <- function(fit, model_choice, prep, transform_posterior = TRUE) {

  draws <- as_draws_df(fit)
  
  if (transform_posterior) {
    param_cols <- !grepl("^\\.|^lp__", colnames(draws))
    draws[param_cols] <- lapply(draws[param_cols], function(x) exp(x) - 1)
  }

  lookup <- prep$lookup

  summarize_draws_df <- function(x) {
    tibble(
      median = apply(x, 2, median),
      q5   = apply(x, 2, quantile, 0.05),
      q95  = apply(x, 2, quantile, 0.95)
    )
  }

  ## global
  global <- tibble(
    median = median(draws$alpha0),
    q5     = quantile(draws$alpha0, 0.05),
    q95    = quantile(draws$alpha0, 0.95),
    level  = "global",
    id     = "global"
  )

  ## country (% change)
  country <- NULL
  country_cols <- names(draws)[grepl("^alpha_country_eff\\[", names(draws))]
  if (length(country_cols) > 0) {
    country <- summarize_draws_df(draws %>% select(all_of(country_cols))) %>%
      mutate(level = "country", country_index = row_number())

    if (!is.null(lookup$country_lookup)) {
      country <- country %>%
        left_join(lookup$country_lookup %>% select(country_id, country),
                  by = c("country_index" = "country_id")) %>%
        mutate(id = country)
    } else {
      country <- country %>% mutate(id = paste0("country_", country_index))
    }
  }
  
  ## raw country deviation
  alpha_country <- NULL
  alpha_country_cols <- names(draws)[grepl("^alpha_country\\[", names(draws))]
  if (length(alpha_country_cols) > 0) {
    alpha_country <- summarize_draws_df(draws %>% select(all_of(alpha_country_cols))) %>%
      mutate(level = "country_raw", country_index = row_number())

    if (!is.null(lookup$country_lookup)) {
      alpha_country <- alpha_country %>%
        left_join(lookup$country_lookup %>% select(country_id, country),
                  by = c("country_index" = "country_id")) %>%
        mutate(id = paste0(country, " (deviation)"))
    } else {
      alpha_country <- alpha_country %>% mutate(id = paste0("country_raw_", country_index))
    }
  }

  ## admin1
  admin1 <- NULL
  admin1_cols <- names(draws)[grepl("^alpha_admin1_eff\\[", names(draws))]
  if (length(admin1_cols) > 0) {
    admin1 <- summarize_draws_df(draws %>% select(all_of(admin1_cols))) %>%
      mutate(level = "admin1", admin1_index = row_number())

    if (!is.null(lookup$admin1_lookup)) {
      admin1 <- admin1 %>%
        left_join(lookup$admin1_lookup %>% select(admin1_index, admin1, country),
                  by = "admin1_index") %>%
        mutate(id = paste0(country, "::", admin1))
    } else {
      admin1 <- admin1 %>% mutate(id = paste0("admin1_", admin1_index))
    }
  }

  ## alert
  alert <- NULL
  alert_cols <- names(draws)[grepl("^alpha_alert_eff\\[", names(draws))]
  if (length(alert_cols) > 0) {
    alert <- summarize_draws_df(draws %>% select(all_of(alert_cols))) %>%
      mutate(level = "alert", alert_index = row_number())

    if (!is.null(lookup$alert_country_admin1_lookup)) {
      alert <- alert %>%
        left_join(lookup$alert_country_admin1_lookup %>%
                    select(alert_index, alert_id, country, admin1),
                  by = "alert_index") %>%
        mutate(id = paste0(country, "::", admin1, "::", alert_id))
    } else if (!is.null(lookup$alert_country_lookup)) {
      alert <- alert %>%
        left_join(lookup$alert_country_lookup %>% select(alert_index, alert_id, country),
                  by = "alert_index") %>%
        mutate(id = paste0(country, "::", alert_id))
    } else if (!is.null(lookup$alert_admin1_lookup)) {
      alert <- alert %>%
        left_join(lookup$alert_admin1_lookup %>% select(alert_index, alert_id, admin1),
                  by = "alert_index") %>%
        mutate(id = paste0(admin1, "::", alert_id))
    } else if (!is.null(lookup$alert_lookup)) {
      alert <- alert %>%
        left_join(lookup$alert_lookup %>% select(alert_index, alert_id),
                  by = "alert_index") %>%
        mutate(id = alert_id)
    } else {
      alert <- alert %>% mutate(id = paste0("alert_", alert_index))
    }
  }

  ## location
  location <- NULL
  if (model_choice == "modelC") {
    loc_cols <- names(draws)[grepl("^alpha_loc_eff\\[", names(draws))]
    if (length(loc_cols) > 0) {
      location <- summarize_draws_df(draws %>% select(all_of(loc_cols))) %>%
        mutate(level = "location", loc_index = row_number())

      if (!is.null(lookup$location_lookup)) {
        location <- location %>%
          left_join(lookup$location_lookup %>% select(loc_index, location),
                    by = "loc_index") %>%
          mutate(id = location)
      } else {
        location <- location %>% mutate(id = paste0("location_", loc_index))
      }
    }
  }

  bind_rows(global, country, alpha_country, admin1, alert, location)
}

slope_summaries <- summarize_effective_slopes(fit, model_choice, transform_posterior = transform_posterior ,prepared)
readr::write_rds(slope_summaries,
  paste0(fig_path, "/slope_summaries_", model_choice, "_a", alert_no, "_", pop_group, "_", location_filter,".rds")
)

```

```{r, fig.height=35, fig.width=15, warning=FALSE}

df_for_plot <- report_model(
  fit, 
  lookup = prepared$lookup, 
  location_filter, 
  return_df = TRUE
)

slope_summaries2 <- df_for_plot %>%
  select(
    id = Parameter,
    median = Median,
    q5 = `2.5%`,
    q95 = `97.5%`
  ) %>%
  mutate(
    level = case_when(
      grepl("^Global", id)   ~ "global",
      grepl("^Country:", id) ~ "country",
      grepl("^Admin1:", id)  ~ "admin1",
      grepl("^Location:", id)~ "location",
      grepl("^Alert:", id)   ~ "alert",
      TRUE                   ~ NA_character_
    )
  ) %>%
  filter(!is.na(level))

plot_effective_slopes <- function(slope_summaries, title = "Posterior effects") {

  ggplot(slope_summaries, aes(x = median, y = id, color = level)) +
    geom_point(size = 2.5) +
    geom_errorbarh(aes(xmin = q5, xmax = q95), height = 0.2, linewidth = 0.6) +
    
    # global reference line (if exists)
    geom_vline(
      data = subset(slope_summaries, level == "global"),
      aes(xintercept = median),
      linetype = "dashed",
      color = "black"
    ) +
    labs(
      title = title,
      x = "% change",
      y = "",
      color = "Level"
    ) +
    theme_minimal(base_size = 16) +
    theme(
      legend.position = "top",
      axis.text.y = element_text(size = 12),
      panel.grid.minor = element_blank()
    )
}

plot_effective_slopes(slope_summaries2)

```

```{r}
#| eval: true

## Map with country-level slopes 
if (model_choice %in% c("modelB", "modelB2", "modelB3")) {

  country_exp_summary <- df_for_plot %>%
    dplyr::filter(grepl("^Country:", .data[["Parameter"]])) %>%
    dplyr::mutate(
      country_code = gsub("^Country: (.+) \\(% change\\)$", "\\1", .data[["Parameter"]]),
      perc_change = .data[["Median"]],   
      perc_q5    = .data[["2.5%"]],
      perc_q95   = .data[["97.5%"]]
    )

  africa <- giscoR::gisco_get_countries(
    region = "Africa",
    resolution = "3"
  )

  africa_slopes <- africa %>%
    left_join(
      country_exp_summary,
      by = c("ISO3_CODE" = "country_code")
    )

    ggplot(africa_slopes) +
    geom_sf(aes(fill = perc_change), color = "gray70") +
    scale_fill_viridis_c(
      option = "D",
      direction = 1,
      na.value = "white",
      name = "Effect per week (%)"
    ) +
    ggspatial::annotation_scale(
      location = "bl",
      width_hint = 0.25,
      text_cex = 0.6
    ) +
    ggspatial::annotation_north_arrow(
      location = "tl",
      which_north = "true",
      style = ggspatial::north_arrow_fancy_orienteering,
      height = unit(0.8, "cm"),
      width  = unit(0.8, "cm")
    ) +
    theme_minimal() +
    theme(
      legend.position = "right",
      plot.title = element_text(hjust = 0.5),
      axis.ticks = element_blank(),
      axis.text = element_blank(),
      axis.title = element_blank()
    )
}

```

```{r, fig.height=20, fig.width=15, warning=FALSE}
#| code-fold: true
#| eval: true

if (model_choice == "modelB2") {
  
  plot_alerts_by_country_B2 <- function(summaries) {
  summaries <- summaries %>%
    filter(level %in% c("country", "alert")) %>%
    mutate(
      ## extract date from alert id
      date = as.Date(sub(".*_(\\d{4}-\\d{2}-\\d{2})$", "\\1", alert_id)),
      ## get administrative level
      admin_level = case_when(
        str_count(alert_id, "::") == 1 ~ "0",
        str_count(alert_id, "::") == 2 ~ "1",
        str_count(alert_id, "::") == 3 ~ "2",
        str_count(alert_id, "::") == 4 ~ "3",
        TRUE ~ "other"
      )
    )
  
  ggplot() +
    geom_errorbarh(
      data = summaries %>% filter(level == "alert"),
      aes(xmin = q5, xmax = q95, x = median, y = date, color = admin_level),
      height = 0.3
    ) +
    geom_point(
      data = summaries %>% filter(level == "alert"),
      aes(x = median, y = date, color = admin_level),
      size = 3
    ) +
    
    ## country slope 
    geom_errorbarh(
      data = summaries %>% filter(level == "country") %>%
        left_join(summaries %>% group_by(country) %>%
                    summarise(min_date = min(date, na.rm = TRUE)), by = "country") %>%
        mutate(plot_date = min_date - 10),
      aes(xmin = q5, xmax = q95, x = median, y = plot_date),
      height = 0.5, color = "black", size = 1.2,
      inherit.aes = FALSE
    ) +
    geom_point(
      data = summaries %>% filter(level == "country") %>%
        left_join(summaries %>% group_by(country) %>%
                    summarise(min_date = min(date, na.rm = TRUE)), by = "country") %>%
        mutate(plot_date = min_date - 10),
      #aes(x = mean, y = plot_date),
      aes(x = median, y = plot_date),
      color = "black", size = 5,
      inherit.aes = FALSE
    ) +
    
    coord_flip() +
    facet_wrap(~ country, scales = "free") +
    scale_y_date(
      date_labels = "%Y-%m-%d"  
    ) +
    labs(
      x = "Slope", 
      y = "Date", 
      color = "Administrative Level",
      title = "Country slopes and individual alerts"
    ) +
    theme_minimal(base_size = 20) +
    theme(
      axis.text.x = element_text(angle = 30)
    )
}
  
  plot_alerts_by_country_B2(slope_summaries)
} 

## MODEL B3
plot_alerts_by_country_B3 <- function(summaries) {

  summaries <- summaries %>%
    filter(level %in% c("alert", "country"))
  message("Rows after filtering levels: ", nrow(summaries))

  summaries <- summaries %>%
    mutate(
      country_iso3 = case_when(
        level == "country" ~ id,
        level == "alert" ~ str_split(id, "::", simplify = TRUE)[,1],
        TRUE ~ NA_character_
      )
    )
  message("Unique country_iso3 values:")
  print(unique(summaries$country_iso3))

  summaries <- summaries %>%
    mutate(
      admin_level = case_when(
        level == "alert" & str_count(alert_id, "::") == 2 ~ "1",
        level == "alert" & str_count(alert_id, "::") == 3 ~ "2",
        level == "alert" & str_count(alert_id, "::") == 4 ~ "3",
        level == "alert" ~ "0",
        TRUE ~ NA_character_
      )
    )

  summaries <- summaries %>%
    mutate(
      date = as.Date(sub(".*_(\\d{4}-\\d{2}-\\d{2})$", "\\1", id), format = "%Y-%m-%d")
    )

  country_positions <- summaries %>%
    filter(level == "country") %>%
    #select(country_iso3, mean, q5, q95) %>%
    select(country_iso3, median, q5, q95) %>%
    mutate(plot_date = min(summaries$date[summaries$country_iso3 == country_iso3], na.rm = TRUE) - 10)

  summaries <- summaries %>%
    left_join(country_positions %>% select(country_iso3, plot_date), by = "country_iso3")

  ggplot() +
    geom_errorbarh(
      data = summaries %>% filter(level == "alert"),
      aes(xmin = q5, xmax = q95, x = median, y = date, color = admin_level),
      height = 0.3
    ) +
    geom_point(
      data = summaries %>% filter(level == "alert"),
      aes(x = median, y = date, color = admin_level),
      size = 3
    ) +
    geom_errorbarh(
      data = summaries %>% filter(level == "country" & !is.na(plot_date)),
      aes(xmin = q5, xmax = q95, x = median, y = plot_date),
      height = 0.5, color = "black", size = 1.2,
      inherit.aes = FALSE
    ) +
    geom_point(
      data = summaries %>% filter(level == "country" & !is.na(plot_date)),
      aes(x = median, y = plot_date),
      color = "black", size = 5,
      inherit.aes = FALSE
    ) +
    coord_flip() +
    facet_wrap(~ country_iso3, scales = "free_y") +
    scale_y_date(date_labels = "%Y-%m-%d") +
    labs(
      x = "Slope",
      y = "Date",
      color = "Administrative Level",
      title = "Country slopes and individual alerts"
    ) +
    theme_minimal(base_size = 20) +
    theme(axis.text.x = element_text(angle = 30))
}

if (model_choice == "modelB3") {
  plot_alerts_by_country_B3(slope_summaries)
}

```

```{r}
#| code-fold: true
#| eval: true

## check that a soft sum to zero constraint is indeed in place 
draws_list <- rstan::extract(fit, permuted = TRUE)

if ("alpha_country" %in% names(draws_list)) {
  alpha_country_draws <- draws_list$alpha_country  
  ## posterior means per country
  country_means <- colMeans(alpha_country_draws)
  sum_country_means <- sum(country_means)
  message("Sum of posterior means (country slopes): ", sum_country_means)

  ## distribution of the sum across draws
  sum_country_draws <- rowSums(alpha_country_draws)  
  mean_sum_country <- mean(sum_country_draws)
  sd_sum_country   <- sd(sum_country_draws)
  ci_sum_country   <- quantile(sum_country_draws, c(0.025, 0.5, 0.975))
  print(list(mean = mean_sum_country, sd = sd_sum_country, ci = ci_sum_country))
}

if ("alpha_alert" %in% names(draws_list)) {
  alpha_alert_draws <- draws_list$alpha_alert  ## draws x J
  alert_means <- colMeans(alpha_alert_draws)
  sum_alert_means <- sum(alert_means)
  message("Sum of posterior means (alert slopes): ", sum_alert_means)

  sum_alert_draws <- rowSums(alpha_alert_draws)
  print(list(mean = mean(sum_alert_draws),
             sd = sd(sum_alert_draws),
             ci = quantile(sum_alert_draws, c(0.025,0.5,0.975))))
}


```

```{r,fig.height=32, fig.width=15}
#| code-fold: true
#| eval: true

### MODELB2
## VERSION 1: Creating a straight line going through all the delay values and just plotting the predicted slope per alert and per country

if(model_choice == "modelB2"){
  
  obs_logratio_country <- data %>%
  group_by(country, delay) %>%
  summarise(
    obs_log_ratio = mean(Log_Sch_ratio, na.rm = TRUE),
    .groups = "drop"
  )

  ## function version with observed
  plot_slopes_B2 <- function(slope_summaries, obs_country,
                           delays = seq(from=0,to=24, by=1)) {

  plot_data <- slope_summaries %>%
    filter(level %in% c("country", "alert")) %>%
    tidyr::expand_grid(delay = delays) %>%
    mutate(
      #predicted = mean * delay,
      predicted = median * delay,
      predicted_low = q5 * delay,
      predicted_high = q95 * delay,
      id_short = case_when(
        level == "alert" & !is.na(alert_index) ~ paste0("alert_", alert_index),
        level == "alert" & is.na(alert_index) ~ "alert",
        level == "country" ~ country
      )
    )

  ggplot(plot_data, aes(x = delay)) +

    ## alert
    geom_ribbon(data = plot_data %>% filter(level == "alert"),
                aes(ymin = predicted_low, ymax = predicted_high,
                    group = id_short, fill = "Alert"),
                alpha = 0.2) +
    geom_line(data = plot_data %>% filter(level == "alert"),
              aes(y = predicted, group = id_short, color = "Alert"),
              alpha = 0.6, size = 0.8) +

    ## country
    geom_ribbon(data = plot_data %>% filter(level == "country"),
                aes(ymin = predicted_low, ymax = predicted_high,
                    fill = "Country"),
                alpha = 0.2) +
    geom_line(data = plot_data %>% filter(level == "country"),
              aes(y = predicted, color = "Country"),
              size = 1.5, linetype = "dashed") +

    ## observed average
    geom_point(data = obs_country,
               aes(x = delay, y = obs_log_ratio),
               shape = 21, fill = "black", color = "black", size = 2.5) +

    facet_wrap(~ country, scales = "free_y") +

    scale_color_manual(name = "Level",
                       values = c("Alert" = "steelblue",
                                  "Country" = "firebrick")) +
    scale_fill_manual(name = "Level",
                      values = c("Alert" = "steelblue",
                                 "Country" = "firebrick")) +

    labs(
      x = "Delay",
      y = "Log-ratio",
      title = "Predicted slopes with observed country-average log-ratios"
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "top")
}

 ## apply plotting function for model B2
  plot_slopes_B2(slope_summaries, obs_logratio_country)
}

## VERSION 2: creating the alert line using the predicted ratio for each observation at different delay values 
## and linking observation points with lines if they belong to the same alert
## here the country line is just the average of the predicted ratios per observation at the country level

if (model_choice == "modelB2") {

  draws <- as_draws_df(fit)
  ## predicted log-ratios
  r_rep_cols <- grep("^r_rep\\[", colnames(draws), value = TRUE)
  r_rep_long <- draws %>%
    as_tibble() %>%   # drop draws_df class
    select(all_of(r_rep_cols)) %>%
    pivot_longer(cols = everything(),
                 names_to = "obs",
                 values_to = "r_rep") %>%
    mutate(obs = as.integer(str_extract(obs, "\\d+")))

  ## table for each observation
  obs_lookup <- tibble(
    obs = seq_along(stan_data$d),
    delay = as.numeric(stan_data$d),
    alert_index = stan_data$alert,
    country_index = stan_data$country
  ) %>%
    left_join(
      data %>% distinct(alert_id_index, alert_id, country_id, country),
      by = c("alert_index" = "alert_id_index",
             "country_index" = "country_id")
    )

  r_rep_long <- r_rep_long %>%
    left_join(obs_lookup, by = "obs")

  ## summarising posterior predictive distribution
  pred_summary <- r_rep_long %>%
    group_by(country, alert_id, delay) %>%
    summarise(
      mean = mean(r_rep),
      q5   = quantile(r_rep, 0.05),
      q95  = quantile(r_rep, 0.95),
      .groups = "drop"
    )

  ggplot(pred_summary, aes(x = delay, y = mean, group = alert_id)) +
    geom_ribbon(aes(ymin = q5, ymax = q95, group = alert_id),
                fill = "grey70", alpha = 0.2) +
    geom_line(color = "grey40", size = 0.6, alpha = 0.7) +
    geom_point(color = "grey20", size = 1) +

    ## country averages (mean of alerts per country)
    stat_summary(
      fun = mean,
      geom = "line",
      aes(group = country),
      color = "firebrick", size = 1.2
    ) +

    facet_wrap(~ country, scales = "free_y") +

    labs(
      x = "Delay",
      y = "Predicted log-ratio",
      title = "Posterior predictive log-ratio by delay"
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")

}


```

```{r}
#| code-fold: true
#| eval: true

## line plot
### MODEL A2

## VERSION 1: Creating a straight line going through all the delay values and just plotting the predicted slope per alert and per country

if (model_choice == "modelA2") {
  
  plot_slopes_A2 <- function(slope_summaries, delay_values) {
    
    plot_data <- slope_summaries %>%
      filter(level == "alert") %>%
      tidyr::expand_grid(delay = delay_values) %>%
      mutate(
        predicted = median * delay,
        predicted_low = q5 * delay,
        predicted_high = q95 * delay,
        id_short = ifelse(!is.na(alert_id), alert_id, paste0("alert_", id))
      )
    
    ggplot(plot_data, aes(x = delay)) +
      
      ## alerts + CI
      geom_ribbon(aes(ymin = predicted_low, ymax = predicted_high, group = id_short, fill = "Alert"),
                  alpha = 0.1) +
      geom_line(aes(y = predicted, group = id_short, color = "Alert"),
                alpha = 0.7, size = 0.9) +
      
      scale_color_manual(name = "Level", values = c("Alert" = "steelblue")) +
      scale_fill_manual(name = "Level", values = c("Alert" = "steelblue")) +
      
      labs(
        x = "Delay",
        y = "Predicted log-ratio",
        title = "Predicted log-ratio by delay for alerts (Model A2)"
      ) +
      theme_minimal(base_size = 14) +
      theme(legend.position = "top")
  }
  
  ## apply plotting function for model A2
  plot_slopes_A2(slope_summaries, delay_values = delays)
}

## VERSION 2: creating the alert line using the predicted ratio for each observation at different delay values 
## and linking observation points with lines if they belong to the same alert
## here the country line is just the average of the predicted ratios per observation at the country level

if (model_choice == "modelA2") {

  draws <- as_draws_df(fit)

  ## getting fitted values per observation
  r_rep_cols <- grep("^r_rep\\[", colnames(draws), value = TRUE)
  r_rep_long <- draws %>%
    select(all_of(r_rep_cols)) %>%
    pivot_longer(cols = everything(),
                 names_to = "obs",
                 values_to = "r_rep") %>%
    mutate(obs = as.integer(str_extract(obs, "\\d+")))
  
  ## one row per observation
  obs_lookup <- tibble(
    obs = seq_along(stan_data$d),
    delay = as.numeric(stan_data$d),
    alert_index = stan_data$alert
  ) %>%
    left_join(data %>% distinct(alert_id_index, alert_id),
              by = c("alert_index" = "alert_id_index"))
  
  r_rep_long <- r_rep_long %>%
    left_join(obs_lookup, by = "obs")
  
  ## summarise posterior predictive distribution
  pred_summary <- r_rep_long %>%
    group_by(alert_id, delay) %>%
    summarise(
      mean = mean(r_rep),
      q5 = quantile(r_rep, 0.05),
      q95 = quantile(r_rep, 0.95),
      .groups = "drop"
    )
  
  ## plot
  ggplot(pred_summary, aes(x = delay, y = mean, group = alert_id)) +
    geom_ribbon(aes(ymin = q5, ymax = q95, group = alert_id), 
                fill = "grey70", alpha = 0.2) +
    geom_line(color = "grey40", size = 0.7, alpha = 0.7) +
    geom_point(color = "grey20", size = 1) +
    labs(
      x = "Delay",
      y = "Predicted log-ratio",
      title = "Posterior predictive log-ratio by delay (each line is an alert)"
    ) +
    theme_minimal(base_size = 14) +
    theme(legend.position = "none")
}

```

## Model Selection with LOO and WAIC

We compare the candidate models with Leave One Out Cross-Validation (LOO) and the Widely Applicable Information Criterion (WAIC) because they work best in the context of Bayesian Hierarchical Models. They use samples from the posterior distribution rather than using a single point estimate like the Akaike Information Criterion (AIC) which uses Maximum Likelihood Estimation.

WAIC calculates a bias-corrected estimate of predictive accuracy based on the log-likelihood evaluated across all posterior samples, while LOO approximates cross-validation by evaluating how well the model predicts each data point when that point is left out during fitting. Both criteria provide uncertainty (as standard errors) instead of just point estimates.

Another benefit of using information criteria is that they allow selecting the most parsimonious model. While criteria like AIC do this by penalizing models with a larger number of parameters, in Bayesian Hierarchical Models the notion of "effective number of parameters" is less clear: The number of effective parameters in this context can be ambiguous because the model "shares" information across groups (partial pooling means some of the variation can be explained by a higher level) and incorporates priors which limit how much a parameter can vary, reducing its "effective" contribution to model complexity.

**The model with the highest LOO and the WAIC values is preferred**

```{r model_comparison, eval=run_model_selection}
#| code-fold: true
#| results: asis

model_ids <- c("modelA1", "modelA2", "modelB", "modelB2","modelB3", "modelC", "modelD")
model_labels <- c(
  "A1: Simple null model with global rate",
  "A2: Global slope and alert-specific deviation",
  "B: Global slope and country-specific deviation",
  "B2: Global slope, country-specific deviation, and alert-level deviation",
  "B3: Global slope, country, admin1, and alert-level deviation",
  "C: Global slope and location-specific deviation", 
  "D: Global slope, admin1, and alert-specific deviation"
)

fit_files <- paste0("fit_", model_ids, "_a", alert_no, "_", pop_group, "_", location_filter,".rds")

## only use the models whose fit files actually exist
existing_models <- file.exists(fit_files)
fit_files <- fit_files[existing_models]
model_ids <- model_ids[existing_models]
model_labels <- model_labels[existing_models]

if (length(fit_files) < 1) {
  stop("No model fit files found for this alert and population group.")
}

## memory-safe LOO/WAIC computation
compute_loo_waic <- function(file) {
  fit <- readRDS(file)

  loglik <- extract_log_lik(fit, merge_chains = FALSE)

  out <- list(
    loo = loo(loglik),
    waic = waic(loglik)
  )

  rm(fit, loglik)
  gc()

  out
}

results <- purrr::map(fit_files, compute_loo_waic)

loos  <- purrr::map(results, "loo")
waics <- purrr::map(results, "waic")


```

**LOO**

```{r,  eval=run_model_selection}
#| code-fold: true
#| results: asis

names(loos) <- model_labels
if (length(loos) >= 2) {
  print(loo::loo_compare(loos))
} else {
  message("Not enough models for loo_compare (requires at least 2).")
}
```

**WAIC**

```{r,  eval=run_model_selection}
#| code-fold: true
#| results: asis

## create a WAIC summary table
waic_df <- tibble(
  Model = model_labels,
  elpd_waic = map_dbl(waics, ~ .x$estimates["elpd_waic", "Estimate"]),
  se_waic = map_dbl(waics, ~ .x$estimates["elpd_waic", "SE"])
)

waic_df

```

```{r, eval=run_model_selection}
#| code-fold: true
#| results: asis

## manuscript LOO table

names(loos) <- model_labels

if (length(loos) >= 2) {
  loo_tbl <- as.data.frame(loo::loo_compare(loos))
  loo_tbl$Model <- rownames(loo_tbl)
  rownames(loo_tbl) <- NULL
  loo_tbl <- loo_tbl[, c("Model", setdiff(names(loo_tbl), "Model"))]

  loo_table <- kable(
    loo_tbl,
    caption = paste0("Model Comparison using LOO for ",location_filter, " locations"),
    digits = 2,
    align = c("l", rep("r", ncol(loo_tbl) - 1))
  ) %>%
    kable_styling(full_width = FALSE, position = "center") 

    print(loo_table)

    save_kable(loo_table, file = "loo_comparison.html")
  
} else {
  message("Not enough models for loo_compare (requires at least 2).")
}

## manuscript WAIC table

names(waics) <- model_labels

if (length(waics) >= 2) {
  waic_tbl <- as.data.frame(loo::loo_compare(waics))
  waic_tbl$Model <- rownames(waic_tbl)
  rownames(waic_tbl) <- NULL
  waic_tbl <- waic_tbl[, c("Model", setdiff(names(waic_tbl), "Model"))]

  waic_pretty <- waic_tbl %>%
    kable(
      caption = paste0("Model Comparison using WAIC for ",location_filter, " locations"),
      digits = 2,
      align = c("l", rep("r", ncol(waic_tbl) - 1))
    ) %>%
    kable_styling(
      full_width = FALSE,
      position = "center"
    )

  print(waic_pretty)

  save_kable(waic_pretty, file = "waic_comparison.html")

} else {
  message("Not enough models for loo_compare (requires at least 2).")
}


```

```{r compare_B2_D_alpha0, eval=compare_epidemic_endemic_posteriors}
#| code-fold: true
#| warning: false
#| message: false
#| fig-height: 6

## load models fits
fit_file_B2 <- glue::glue("fit_modelB2_a{alert_no}_{pop_group}_outbreak-prone.rds")
fit_file_D  <- glue::glue("fit_modelD_a{alert_no}_{pop_group}_endemic.rds")

fit_B2 <- readRDS(fit_file_B2)
fit_D  <- readRDS(fit_file_D)

## convert draws to data frames
draws_B2 <- posterior::as_draws_df(fit_B2)
draws_D  <- posterior::as_draws_df(fit_D)

## posterior density for alpha0
alpha_df <- bind_rows(
  draws_B2 %>% select(alpha0) %>% mutate(model = "Model B2 (Outbreak-prone)"),
  draws_D  %>% select(alpha0) %>% mutate(model = "Model D (Endemic)")
)

p_alpha0 <- ggplot(alpha_df, aes(x = alpha0, fill = model, color = model)) +
  geom_density(alpha = 0.3, size = 1) +
  labs(
    title = expression("Posterior distributions of " ~ alpha[0]),
    x = expression(alpha[0]),
    y = "Density"
  ) +
  theme_minimal(base_size = 18) +
  theme(legend.position = "top", plot.title = element_text(face = "bold"))

## posterior predictive density for r_rep
rrep_B2 <- draws_B2 %>%
  dplyr::select(starts_with("r_rep")) %>%
  tidyr::pivot_longer(everything(), values_to = "r_rep") %>%
  dplyr::mutate(model = "Model B2 (Outbreak-prone)")

rrep_D <- draws_D %>%
  dplyr::select(starts_with("r_rep")) %>%
  tidyr::pivot_longer(everything(), values_to = "r_rep") %>%
  dplyr::mutate(model = "Model D (Endemic)")

## combine
set.seed(123)
rrep_combined <- bind_rows(rrep_B2, rrep_D) %>%
  sample_n(min(5000, nrow(.)))  ## limit to 5000 draws

p_rrep <- ggplot(rrep_combined, aes(x = r_rep, fill = model, color = model)) +
  geom_density(alpha = 0.3, size = 1) +
  labs(
    title = "Posterior predictive densities for log-ratios",
    x = "Predicted log-ratio (r_rep)",
    y = "Density"
  ) +
  theme_minimal(base_size = 18) +
  theme(legend.position = "top")

## plots
combined_plot <- p_alpha0 / p_rrep + 
  plot_annotation(title = "")

combined_plot

```
